### 2015-02-18 ParInfpaper/stochInf1D.tex
First upload of Tex files (paper)

\documentclass[11pt, a4paper]{article}

\usepackage{a4wide}
\usepackage{amsmath,amssymb,amsthm,amscd}
\usepackage{natbib}
\usepackage[pdftex]{graphicx}
\usepackage[affil-it]{authblk}
\usepackage{sidecap}
\usepackage[font=small,labelfont=bf]{caption}

%\usepackage{showkeys}

\usepackage[all]{xy}

\newcommand{\bb}[1]{\begin{equation}\label{#1}}
\newcommand{\ee}{\end{equation}}
\newcommand{\vc}[1]{{\bf #1 }}
\renewcommand{\vec}[1]{{\bf #1}}
\newcommand{\Sum}[1]{\sum\limits_{#1}}
\newcommand{\Int}[1]{\int\limits_{#1}}
\newcommand{\Prod}[1]{\prod\limits_{#1}}
\newcommand{\const}{\operatorname{const}}
\newcommand{\ul}{\underline}
\newcommand{\ol}{\overline}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\bbA}{\mathbb A}
\newcommand{\bbR}{\mathbb R}
\newcommand{\bbN}{\mathbb N}
\newcommand{\bbZ}{\mathbb Z}
\newcommand{\bbC}{\mathbb C}
\newcommand{\bbT}{\mathbb T}
\newcommand{\QB}{\mathcal H}
\newcommand{\PQB}{\tilde{\mathcal H}}
\newcommand{\CQB}{\mathcal H'}
\newcommand{\Ad}{\operatorname{Ad}}
\newcommand{\ad}{\operatorname{ad}}
\newcommand{\gh}{\operatorname{gh}}
\newcommand{\Der}{\operatorname{Der}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\aut}{\operatorname{aut}}
\newcommand{\Diff}{\operatorname{Diff}}
\renewcommand{\dim}{\operatorname{dim}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\Pf}{\operatorname{Pf}}
\newcommand{\Cov}{\operatorname{Cov}}

\newcommand{\g}{{\bf g}}
\newcommand{\h}{{\bf h}}
\renewcommand{\d}{{\bf d}}
\newcommand{\e}{{\bf e}}
\renewcommand{\t}{{\bf t}}
\newcommand{\lpartial}{\overset{\rightarrow}{\partial}}
\newcommand{\rpartial}{\overset{\leftarrow}{\partial}}
\newcommand{\lder}[1]{\overset{\rightarrow}{\vc #1}}
\newcommand{\rder}[1]{\overset{\leftarrow}{\vc #1}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\im}{\operatorname{Im}}
\renewcommand{\ker}{\operatorname{Ker}}
\newcommand{\Var}{\operatorname{Var}}

%define the \identy operator
\def\litID{{\sf id}}
\def\identy{{\mathsurround0pt\mathchoice{\textidenty}{\textidenty}{\scptidenty}{\scptidenty}}}
\def\scptidenty{\setbox0\hbox{$\scriptstyle1$}\bothidenty}
\def\textidenty{\setbox0\hbox{$1$}\bothidenty}
\def\bothidenty{\rlap{\hbox to.97\wd0{\hss\vrule height.06\ht0 width.82\wd0}}
 \copy0\rlap{\kern-.36\wd0\vrule height1.05\ht0 width.05\ht0}\kern.14\wd0}
%end of definition \identy

\newcommand{\bbM}{\mathbb M}
\newcommand{\bt}{\pmb\theta}
\newcommand{\Bt}{\pmb\Theta}
\newcommand{\dt}{\mathit{dt}}
\newcommand{\eq}{\text{eq}}

\begin{document}
\title{Parameter Inference for a Simple Stochastic Hydrological Model}

\author{C. Albert\thanks{\noindent Email: carlo.albert@eawag.ch} \,,
S. Ulzega \,, P. Reichert}
\affil{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
\maketitle

The dynamics of a linear reservoir with state-dependent noise can be described by a stochastic differential equation, interpreted according to the Stratonovich convention \textbf{[ref?]},

\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}

where $r(t)$ describes the rain input, $S(t)$ is the volume of water in the system, $K$ is the {\em retention time} and the dimensionless parameter $\gamma$ quantifies the noise ${\eta}(t)$. Henceforth we shall assume standard white noise, i.e.,

\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}

The term $\left(1+\gamma/2\right)$ in (\ref{sde}) arises from the Stratonovich interpretation. According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S(T)$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,

\begin{equation}\label{pathint}
P(S) = \int_{S_0}^{S(T)} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{DS} \,,
\end{equation}

where the integral extends over all paths starting at $S_0$ and ending at $S$. Following \cite{lau_2007_StateDepDiff}, one can derive from (\ref{sde}) and (\ref{whitenoise}) the formal path-measure,

\begin{equation}\label{pathmeasure}
{\mathcal DS}
=
\prod_{t}
\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
\end{equation}

defined as an infinite product in the formal limit $\Delta t \rightarrow 0$, and the {\em action} $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$,

\begin{equation}\label{action}
  \tilde{\mathcal S}[\dot{S(t)},S(t)]
  =
  \int_0^T
  dt
  \left\{
    \frac{K}{2\gamma S^2(t)}
    \left(
        \dot S(t)
        -
        r(t)
        +
        \frac{1+\gamma}{K}S(t)
    \right)^2
    -
    \frac{2+\gamma}{4K}
  \right\}
 \,.
\end{equation}

Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S) \mathit{dS}}=1$.

It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S(T))$ in the simple case of a constant rain input $r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,

\begin{equation}\label{detailed_balance}
P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,,
\end{equation}

where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S} = S(-t)$. Assuming a solution in the form

\begin{equation}\label{Peq}
  P_{\text{eq}}(S)
  \propto
  e^{-\mathcal{H}(S)}\,,
\end{equation}

equation (\ref{detailed_balance}) can be rewritten as

\begin{equation}\label{detailed_balance_2}
P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\partial \mathcal{H}}{\partial S}}\,.
\end{equation}

After plugging in (\ref{pathint}) for the conditional probabilities, and using the action (\ref{action}), equation (\ref{detailed_balance_2}) yields

\begin{equation}\label{constant_rain_H}
-r_{0}+\frac{1+\gamma}{K}S
=
\frac{\gamma S^2}{2K}
\frac{d \mathcal{H}(S)}{dS}\,.
\end{equation}

In order to go from (\ref{detailed_balance_2}) to (\ref{constant_rain_H}) we have exploited the property of the Stratonovich convention that the backward path is evaluated at the same points as the forward path. The solution to (\ref{constant_rain_H}) is, up to an additive constant,

\begin{equation}\label{solution_const_rain}
\mathcal{H}(S)=\frac{2 K r_0}{\gamma S}+\frac{2(1+\gamma)}{\gamma}\ln(S)\,.
\end{equation}

%\begin{figure}
%    \centering
%    \includegraphics[width=0.6\textwidth]{Figs/InvGamma.png}
%    \caption{Inverse gamma distributions with scale parameter $\beta = 5$ and shape parameters $\alpha = 3,4,5,6$.}
%    \label{fig:inv_gamma}
%\end{figure}

Thus, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution, $\Gamma^{-1}(\alpha,\beta)$ with scale parameter $\beta=2Kr_{0}/\gamma$ and shape parameter $\alpha=(2+\gamma)/\gamma$, i.e.,

\begin{equation}\label{inverse_gamma}
  P_{\text{eq}}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}

whose mean equals the equilibrium solution of the unperturbed system,

\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}

and whose variance, for $\gamma< 2$, is given by

\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}

Note that the variance diverges for $\gamma\geq 2$.

\vspace{\baselineskip}

Let us consider now the general case of a time dependent rain input $r(t)$ with the following time-series of {\em observed} outputs,

\begin{equation}\label{data}
  y_i=\frac{1}{K}S(t_i)+\sigma\epsilon_i\,,\quad i=1,\dots,n\,,
\end{equation}

where the times $t_i$ are chosen so that $0=t_1<t_2<\dots < t_n=T$ and the $\epsilon_i$ are uncorrelated standard normal random variables.
The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral

\begin{equation}\label{likelihood}
  f(\vc y | \bt)
  \propto
  \int
  \exp\bigg[
    -{\tilde{\mathcal S}}[\dot{S},S]
    -\frac{1}{2}
    \sum_{i=1}^n
    \frac
    {(y_i-S(t_i)/K)^2}
    {\sigma^2}
  \bigg]
  {\mathcal D} S(\bt)\,,
\end{equation}

with the parameter vector $\bt=(K,\gamma)^T$. Choosing for the parameters $K$ and $\gamma$ the improper prior given by

\begin{equation}\label{prior}
  f(\bt)
  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
\end{equation}

{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
\begin{equation}\label{posterior}
  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
\end{equation}

After the transformation $S(t) = K r(t) e^{q(t)}$, the posterior reads

\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -{\tilde{\mathcal S}}[\dot{q(t)},q(t)]
    -\frac{1}{2}
    \sum_{i=1}^n
    \frac
    {(y_i-r_ie^{q_i})^2}
    {\sigma^2}
    -\ln(K\gamma)
  \bigg]
  {\mathcal Dq}({\pmb\theta})
  \,,
\end{equation}

%\prod_{t}
%\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,

where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,

\begin{equation}\label{pathmeasure_q}
{\mathcal Dq} (\bt)
=
\prod_{t}
\sqrt{\frac{K}{\gamma}} dq(t)\,,
\end{equation}

while the action can be rewritten in the compact form,

\begin{equation}\label{action_new}
{\mathcal S}[\dot{q},q]
=
\frac{1}{2\gamma K}
\int_0^T \dt \left\{\left(
    K^2\dot q^2(t) +
    \rho(t)-e^{-q(t)}\right)^2 -
    \frac{\gamma(2+\gamma)}{2}
\right\} \,,
\end{equation}

with
\begin{equation}\label{rho}
\rho(t)=K\frac{\dot r(t)}{r(t)}+\gamma+1 = K\frac{d \left[\ln r(t)\right]}{dt}+\gamma+1\,.
\end{equation}

In order to ease the numerical computations, it is useful to introduce a generalized time-dependent Hamiltonian,

\begin{equation}\label{H}
  \mathcal{H}_q(q,t)
  =
  e^{-q}+q\rho(t)\,,
\end{equation}

and rewrite the action (\ref{action_new}) as,

\begin{multline}\label{action_final}
{\mathcal S}[\dot{q},q]
=
\frac{1}{2\gamma K}
\int_0^T dt\left\{
    K^2\dot q^2(t) +
    (\rho(t)-e^{-q(t)})^2 -
    2K\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
    \frac{\gamma(2+\gamma)}{2}
\right\}
\\
+
\frac{1}{\gamma}\left(
    \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
\right)
\\
= \frac{1}{2\gamma K}
\int_0^T dt\left\{
    K^2\dot q^2(t) +
    (\rho(t)-e^{-q(t)})^2 -
    2Kq(t)\dot\rho(t) -
    \frac{\gamma(2+\gamma)}{2}
\right\}
\\
+
\frac{1}{\gamma}\left(
    e^{-q(T)}+q(T)\rho(T)
    -e^{-q(0)}-q(0)\rho(0)
\right)
\,.
\end{multline}

Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample simultaneously parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since this distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian

\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}

where

\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int \frac{ p^2(t)}{2m_q}dt
   + \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}

where in our case $s=2$, and $V( q,\bt)$ is the logarithm of the kernel of (\ref{posterior_pathint}).
The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps

\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically (leap-frog algorithm).
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}

The last step is the standard Metropolis algorithm, while the first two steps allow to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

In particular, step 2 requires the calculation of the gradient of (\ref{Hamiltonian}), which is a somewhat tedious manual labour.
Before we set out to do so, let us first write down the proper discretization of the path-integral (\ref{posterior_pathint}), with the action (\ref{action_final}).
Therefore, let us assume that the measurement time points $\left\{ y_i \right\}_{i=1,\dots, n}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$.
We further partition this interval into $N$ bins, with $N >> 1$, according to the time series $0=t_1<t_2,\dots<t_N=T$, such that the time points $t_{a_{\kappa}}$ with indexes $a_{\kappa} = \kappa (N-1)/n + 1$, where $\kappa = 1,\dots, n$, coincide with the data points $y_i$.
Under these assumptions the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ read

\begin{align}
   K( p,{\pmb\pi})
   &=
   \sum_{i=1}^N \frac{ p_i^2}{2m_q}\Delta t
   + \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt)
   &\approx
   \sum_{i=2}^{N}
  \frac{K}{2\gamma\Delta t}
  (q_i-q_{i-1})^2+\sum_{i=2}^{N-1} \frac{\Delta t}{\gamma} \left\{\frac{(\rho_i-e^{-q_i})^2}{2 K}  - q_i \frac{(\rho_i-\rho_{i-1})}{\Delta t} \right\}
  \\
  &- \frac{T (\gamma + 2)}{4 K}+\frac{\left(e^{-q_N}+q_N \rho_{N-1}- e^{-q_1}-q_1 \rho_{1} \right)}{\gamma} \nonumber
  \\
  & + \sum_{\kappa=1}^n \frac{(y_\kappa-r_{a_{\kappa}} e^{q_{a_{\kappa}}})^2}{2\sigma^2} +\ln \left(K^{1-N/2} \gamma^{1+N/2}\right)\nonumber
   \,,\label{Vdisc}
\end{align}

where the last term comes from the path measure (\ref{pathmeasure_q}) and

\begin{equation}\label{rhodisc}
\rho_i = K \frac{\ln(r_{i+1}/r_i)}{\Delta t}+\gamma+1\,.
\end{equation}

%\begin{equation}
%  \dot{q}_i:=\frac{ q_i -  q_{i-1}}{\Delta t}\,,\quad
%  \bar{ q}_i:=\frac{q_i + q_{i-1}}{2}\,,
%\end{equation}
%and the analogous definition of $\dot r_i$ and $\bar r_i$.

For strongly varying rain input or high-frequency output data, $\Delta t$ needs to be chosen small. Thus, the harmonic part of the Hamiltonian becomes stiff and, consequently, the time-step of MD needs to be chosen small, which leads to slow convergence.
To solve this problem, we change coordinates, such that the harmonic part of the Hamiltonian becomes (partly) diagonal and the particle interaction moves to the potential.
One option would be to use the Fourier modes of ${\eta}(t)$ as configurational variables. Calculating the potential requires then an inversion of a Toeplitz matrix. This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes.

Another approach is the so-called \emph{staging}, described in Tuckerman et al. (1993). This approach is interesting if different time-scales are involved.
We first note that the Hamiltonian $\mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})$ (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$. Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics. Thus, let us place $n+1$ equidistant (data) beads on the time domain, the first sitting on $t=0$ and the last on $t=T$. In between, we place $n(j-1)$ equidistant staging beads. So that, in total, we have $nj+1=N$ equidistant beads.

The discretized harmonic part of the action can be rewritten as

\begin{multline}
  \sum_{i=2}^{N}
  \frac{K}{2\gamma\Delta t}
  (q_i-q_{i-1})^2
  \\ =
  \frac{K}{2\gamma}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{multline}

with

\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}

Following Tuckerman et al. (1993) we apply the coordinate transformation

\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}

with its inverse given by

\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}

or equivalently by the recursive relation

\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{s+k+1}+ \frac{1}{k}q_{sj+1} \,.
\end{equation}

Moreover, we introduce the following transformation for the parameters,

\begin{equation}\label{params_transform}
\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
\end{equation}

and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.

The potential $V(q,\tilde{\bt})$ can now be split into a fast component that scales linearly or quadratically with $N$ or $n$,

\begin{multline}
  {\mathcal V}_f(\vc u,\tilde{\bt})
  =
  \frac{N e^{\beta-\tau}}{2}
  \sum_{s=1}^n\left\{
    \frac{(u_{(s-1)j+1} - u_{sj+1})^2}{j}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)}
    u_{(s-1)j+k}^2
  \right\}
  \\
  +
  \frac{N}{2}(\beta-\tau)
      -
    \sum_{s=0}^{n}
    \frac
    {(y_s-r_{sj+1}e^{u_{sj+1}})^2}
    {2 \sigma^2}
  \,,
\end{multline}

where we have used $T = N \Delta t$, and a slow $N-$ and $n-$independent component,

\begin{multline}
{\mathcal V}_s(\vc q,\tilde{\bt})
=
    \sum_{i=2}^{N-1}
        \left\{
            \frac{e^{-(\beta+\tau)}}{2N}
            (\rho_i-e^{-q_i})^2 - e^{-\tau}
            q_i (\rho_i-\rho_{i-1})
        \right\}-\frac{(2+e^\tau)e^{-\beta}}{4}
        \\
    + e^{-\tau} \left( e^{-q_N} + q_N\rho_{N-1}
    - e^{-q_1} - q_1\rho_1\right)
\,.
\end{multline}

In the expressions above we have included the Jacobian associated with (\ref{params_transform}).

\vspace{\baselineskip}

We can now fully exploit the two different time-scales that are involved in the problem. First, the harmonic part of the fast component ${\mathcal V}_f[\dot{\vc u},\vc u]$ can be seen as describing the dynamics of heavy coupled \emph{boundary} data-point beads and lighter uncoupled \emph{staging} beads. The different weights of boundary and staging beads are defined by adjusting their \emph{effective masses}, $m_{\text{bdy}}$ and $m_{\text{st}}$, respectively. These values are used in (\ref{Kdisc}) for the masses $m_{q}$.
Second, the two different ranges of the forces associated with ${\mathcal V}_f(\vc u,\tilde{\bt})$ and ${\mathcal V}_s(\vc q,\tilde{\bt})$ can be efficiently taken into account by applying a reversible reference system propagator algorithm (RESPA), as outlined in Tuckerman et al. (1992). The RESPA algorithm, based on appropriate reversible molecular dynamics integrators, can significantly accelerate simulations of systems that are simultaneously subject to short- and long-range forces by separating the two different time-scales.

The basic idea of the RESPA algorithm is to apply a Trotter decomposition to the system Liouville propagator $\mathcal{L}$ to generate a discrete time propagator factorized into fast and slow contributions. Indeed, as shown in Tuckerman et al. (1992), the propagator $\exp[i \mathcal{L} \Delta \tau ]$ can be written as

\begin{equation}\label{G}
  \mathcal{G}(\Delta \tau)=\exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\} \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\} \exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\}\,,
\end{equation}

where for the sake of clarity we have assumed a one-dimensional system. The argument can be easily extended to our multi-dimensional case. In (\ref{G}) $F_s$ is the force associated with the slowly varying long-range potential ${\mathcal V}_s$ and $\mathcal{L}_f = \dot{x}(\partial/\partial x) + F_f(x)(\partial/\partial p)$ is the Liouville operator of the fast short-range potential ${\mathcal V}_f$. The propagator of the slow force $F_s$ can be well approximated with a large time step $\Delta \tau$. On the other hand, in order to ensure stable dynamics, the short-range middle propagator is further factorized into $n_f$ smaller time steps of duration $\delta \tau = \Delta \tau / n_f$, according to the Trotter expansion

\begin{equation}\label{G_fast}
  \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\}
  =\left[
  \exp\left\{ \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
  \exp\left\{  \delta\tau \dot{x} \frac{\partial}{\partial x}  \right\}
  \exp\left\{  \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
  \right]^{n_f} \,.
\end{equation}

Coordinates and conjugate momenta, i.e., $(q,\bt; p,{\pmb\pi})$ in our case, are propagated using (\ref{G}) and (\ref{G_fast}) iteratively by applying a velocity Verlet algorithm.

\textbf{Some preliminary results}. The results shown in the figures refer to synthetic data obtained assuming a simple sinusoidal model for the rain input. The experimental data were then generated using the "true" parameter values $K=200$ and $\gamma = 0.2$. We considered $n=51$ data points, $j=10$ staging points, and therefore $N=501$ discretization points. The total time $T$ is $\approx 833$ [units of time], so that the discretization points are separated by a time interval $\Delta t \approx 1.6 $ [units of time]. Following the RESPA method described above, we used a long time step $\Delta \tau = 0.02$ [units of time] for the evolution of the slow (long-range) potential, further subdivided in 16 smaller steps for the evolution of the fast (short-range) potential.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
    \caption{Rain input (dashed), system realization (solid line) and synthetic experimental data. The system output and the observed data were generated using $K=200$ and $\gamma = 0.2$.}
    \label{fig:rain_data_S}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{Figs/FigSpaghetti.png}
    \caption{Simulated system realizations and synthetic experimental data obtained assuming a simple sinusoidal model for the rain input. In the inset one may appreciate the different dynamics of the heavy data-point beads and the light staging beads.}
    \label{fig:spaghetti}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{Figs/FigFinalState.png}
    \caption{PDF for the system final state expressed via the coordinate set $\left\{ q_N \right\}$. The dashed curve represents a normal distribution with the same mean and variance. The final state is basically normally distributed.}
    \label{fig:final_state}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{Figs/FigChains.png}
    \caption{Markov chains for the two parameters $K$ (top) and $\gamma$ (bottom).}
    \label{fig:chains}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{Figs/FigPhaseSpaceEvol.png}
    \caption{System dynamics in the phase space $K-\gamma$. The green dot is the initial state, the red dot is the final state, the yellow dots are the last 50 states of the HMC chain, and the orange square corresponds to the true parameter values used to generate the data.}
    \label{fig:phase_space_evol}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{Figs/FigKg.png}
    \caption{PDF for the parameters $K$ (left) and $\gamma$ (right). The true values used to generate the data are $K=200$ and $\gamma = 0.2$. The initial values used in the simulations were $K=50$ and $\gamma = 0.4$.}
    \label{fig:K_distr}
\end{figure}

\bibliographystyle{apalike} % chicago \shortcite(N)
\bibliography{C:/Users/ulzegasi/SWITCHdrive/refs}

\end{document}

### 2015-02-18 TeXFiles/stochInf1D.tex
First upload of Tex files

--- 2015-02-18 ParInfpaper/stochInf1D.tex
@@
 
-The dynamics of a linear reservoir with state-dependent noise can be described by a stochastic differential equation, interpreted according to the Stratonovich convention \textbf{[ref?]},
-
+In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
+We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
 \begin{equation}\label{sde}
@@
 \end{equation}
-
-where $r(t)$ describes the rain input, $S(t)$ is the volume of water in the system, $K$ is the {\em retention time} and the dimensionless parameter $\gamma$ quantifies the noise ${\eta}(t)$.
-Henceforth we shall assume standard white noise, i.e.,
-
+where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
 \begin{equation}\label{whitenoise}
@@
 \end{equation}
+Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the {\em retention time}, and the noise.
+The noise parameter is dimensionless, which renders the noise term scale-invariant.
+Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
+For computational reasons, we work with the {\em Stratonovich} convention.
+Our parametrization is such that, for constant rain input $r(t)=r$, the mean of $S$ converges to $Kr$, in the long-time limit (see below).
 
-The term $\left(1+\gamma/2\right)$ in (\ref{sde}) arises from the Stratonovich interpretation.
-According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S(T)$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,
+According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,
 
 \begin{equation}\label{pathint}
-P(S) = \int_{S_0}^{S(T)} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{DS} \,,
+P(S) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
 \end{equation}
-
-where the integral extends over all paths starting at $S_0$ and ending at $S$. Following \cite{lau_2007_StateDepDiff}, one can derive from (\ref{sde}) and (\ref{whitenoise}) the formal path-measure,
-
+where the integral extends over all paths starting at $S_0$ and ending at $S$.
+The path-measure $\mathcal DS$ is the formal limit $\Delta t\rightarrow 0$ of the discretized measure
 \begin{equation}\label{pathmeasure}
-{\mathcal DS}
+{\mathcal D}S
 =
-\prod_{t}
-\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
+\prod_{i}
+\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,.
 \end{equation}
-
-defined as an infinite product in the formal limit $\Delta t \rightarrow 0$, and the {\em action} $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$,
-
+The {\em action} is a functional defined on paths $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$ and reads as
 \begin{equation}\label{action}
-  \tilde{\mathcal S}[\dot{S(t)},S(t)]
+  \tilde{\mathcal S}[\dot{S}(t),S(t)]
   =
@@
 \end{equation}
-
 Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S) \mathit{dS}}=1$.
@@
 Systems at thermal equilibrium must fulfill detailed balance, i.e.,
-
 \begin{equation}\label{detailed_balance}
@@
 \end{equation}
-
-where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S} = S(-t)$.
+where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S}(t) = S(-t)$.
 Assuming a solution in the form
@@
 \begin{equation}\label{detailed_balance_2}
-P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\partial \mathcal{H}}{\partial S}}\,.
+P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\mathcal{H}}{S}}\,.
 \end{equation}
@@
 
-After the transformation $S(t) = K r(t) e^{q(t)}$, the posterior reads
-
+We replace the variable $S(t)$ and the parameter $K$ by the dimensionless quantities, $q(t)$ and $\beta$, respectively, by means of the transformations
+\begin{equation}
+  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
+  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
+\end{equation}
+In terms of these dimensionless variables and parameters the posterior reads (without prior)
 \begin{equation}\label{posterior_pathint}
@@
   \exp\bigg[
-    -{\tilde{\mathcal S}}[\dot{q(t)},q(t)]
+    -{\tilde{\mathcal S}}[\dot{q}(t),q(t)]
     -\frac{1}{2}
@@
     \frac
-    {(y_i-r_ie^{q_i})^2}
+    {(y_i-r_ie^{\beta q_i})^2}
     {\sigma^2}
-    -\ln(K\gamma)
+    %-\ln(K\gamma)
   \bigg]
-  {\mathcal Dq}({\pmb\theta})
+  {\mathcal Dq}
   \,,
 \end{equation}
-
 %\prod_{t}
 %\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
-
 where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,
@@
 \begin{equation}\label{pathmeasure_q}
-{\mathcal Dq} (\bt)
+{\mathcal Dq}
 =
 \prod_{t}
-\sqrt{\frac{K}{\gamma}} dq(t)\,,
+dq(t)\,,
 \end{equation}
@@
 \begin{equation}\label{action_new}
-{\mathcal S}[\dot{q},q]
+\tilde {\mathcal S}[\dot{q},q]
 =
-\frac{1}{2\gamma K}
+\frac{1}{2T}
 \int_0^T \dt \left\{\left(
-    K^2\dot q^2(t) +
-    \rho(t)-e^{-q(t)}\right)^2 -
-    \frac{\gamma(2+\gamma)}{2}
+    T\dot q(t)
+    +
+    \rho(t)
+    -
+    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
+    -
+    \frac{(2+\gamma)\beta^2}{2\gamma}
 \right\} \,,
@@
 \begin{equation}\label{rho}
-\rho(t)=K\frac{\dot r(t)}{r(t)}+\gamma+1 = K\frac{d \left[\ln r(t)\right]}{dt}+\gamma+1\,.
+\rho(t)=\frac{T\dot r(t)}{\beta r(t)}+\frac{\beta(\gamma+1)}{\gamma}
+=
+\frac{T}{\beta}\frac{d \left[\ln r(t)\right]}{dt}+\frac{\beta(\gamma+1)}{\gamma}\,.
 \end{equation}
@@
   =
-  e^{-q}+q\rho(t)\,,
+  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
 \end{equation}
@@
 =
-\frac{1}{2\gamma K}
+\frac{1}{2T}
 \int_0^T dt\left\{
-    K^2\dot q^2(t) +
-    (\rho(t)-e^{-q(t)})^2 -
-    2K\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
-    \frac{\gamma(2+\gamma)}{2}
+    T^2\dot q^2(t) +
+    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
+    2T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
+     \frac{(2+\gamma)\beta^2}{2\gamma}
+\right\}
+\\
++ \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
+\\
+= \frac{1}{2T}
+\int_0^T dt\left\{
+    T^2\dot q^2(t) +
+    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
+    2Tq(t)\dot\rho(t) -
+    \frac{(2+\gamma)\beta^2}{2\gamma}
 \right\}
@@
 +
-\frac{1}{\gamma}\left(
-    \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
-\right)
-\\
-= \frac{1}{2\gamma K}
-\int_0^T dt\left\{
-    K^2\dot q^2(t) +
-    (\rho(t)-e^{-q(t)})^2 -
-    2Kq(t)\dot\rho(t) -
-    \frac{\gamma(2+\gamma)}{2}
-\right\}
-\\
-+
-\frac{1}{\gamma}\left(
-    e^{-q(T)}+q(T)\rho(T)
-    -e^{-q(0)}-q(0)\rho(0)
-\right)
+    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
+   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
 \,.

### 2015-03-09 TeXFiles/stochInf1D.tex
first NAPA version of tex File

@@
 \author{C. Albert\thanks{\noindent Email: carlo.albert@eawag.ch} \,,
-S. Ulzega \,, P. Reichert}
+S. Ulzega}
 \affil{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
@@
 \end{equation}
-Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the {\em retention time}, and the noise.
-The noise parameter is dimensionless, which renders the noise term scale-invariant.
+Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.
+The scale-invariance of the noise term renders the noise parameter dimensionless.
 Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
 For computational reasons, we work with the {\em Stratonovich} convention.
-Our parametrization is such that, for constant rain input $r(t)=r$, the mean of $S$ converges to $Kr$, in the long-time limit (see below).
+Our parametrization is such that, for constant rain input $r(t)=r$ and in the long-time limit, the mean of $S$ converges to the limit of the unperturbed ($\gamma=0$) system $Kr$ (see below).
 
-According to \cite{lau_2007_StateDepDiff}, the probability $P(S)$ of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as,
-
+According to \cite{lau_2007_StateDepDiff}, the probability, $P(S,T|S_0,t_0)$, of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as
 \begin{equation}\label{pathint}
-P(S) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
+P(S,T|S_0,t_0) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
 \end{equation}
 where the integral extends over all paths starting at $S_0$ and ending at $S$.
-The path-measure $\mathcal DS$ is the formal limit $\Delta t\rightarrow 0$ of the discretized measure
+The path-measure $\mathcal DS$ is the formal limit
 \begin{equation}\label{pathmeasure}
@@
 =
+\lim_{\Delta t\rightarrow 0}
 \prod_{i}
-\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,.
+\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,,
 \end{equation}
-The {\em action} is a functional defined on paths $\tilde{\mathcal S}:[0,T]\longrightarrow \mathbb R_+$ and reads as
+for a discretization $S_i = S(t_i)$ with $t_i-t_{i-1}=\Delta t$.
+The {\em action} is a functional defined on paths $S:[0,T]\longrightarrow \mathbb R_+$ and reads as
 \begin{equation}\label{action}
@@
 \end{equation}
-Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S) \mathit{dS}}=1$.
+Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S,T|S_0,t_0) \mathit{dS}}=1$.
 
-It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S(T))$ in the simple case of a constant rain input $r_{0}$.
+It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S,T|S_0,t_0)$ in the simple case of a constant rain input $r_{0}$.
 Systems at thermal equilibrium must fulfill detailed balance, i.e.,
 \begin{equation}\label{detailed_balance}
-P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,,
+P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,.
 \end{equation}
-where $P(S_1 t_1 | S_0 t_0 )$ is the conditional probability for the state $S_1$ at time $t_1$ given the initial state $S_0$ at time $t_0$ and $P(S_0 t_1 | S_1 t_0 )$ is the conditional probability for the reverse path $\overline{S}(t) = S(-t)$.
-Assuming a solution in the form
+Assuming a solution of the form
 
@@
 \end{equation}
-
-equation (\ref{detailed_balance}) can be rewritten as
+equation (\ref{detailed_balance}) is rewritten as
 
 \begin{equation}\label{detailed_balance_2}
-P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 ) \exp^{-\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{\mathcal{H}}{S}}\,.
+P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 )
+\exp\left\{
+        -\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{d\mathcal{H}}{dS}
+    \right\}
+\,,
 \end{equation}
-
-After plugging in (\ref{pathint}) for the conditional probabilities, and using the action (\ref{action}), equation (\ref{detailed_balance_2}) yields
+where $S(t)$ is an arbitrary path connecting $S_0$ with $S_1$.
+After plugging in (\ref{pathint}) and (\ref{action}), and using the Stratonovich property, which allows us to transform a path from $S_0$ to $S_1$ into a path from $S_1$ to $S_0$ through the transformation $S(t) \rightarrow S(-t)$, we get
 
@@
 \end{equation}
-
-In order to go from (\ref{detailed_balance_2}) to (\ref{constant_rain_H}) we have exploited the property of the Stratonovich convention that the backward path is evaluated at the same points as the forward path.
-The solution to (\ref{constant_rain_H}) is, up to an additive constant,
-
+The solution to (\ref{constant_rain_H}) is, up to an additive constant, given by eq.
 \begin{equation}\label{solution_const_rain}
@@
 \end{equation}
-
 %\begin{figure}
@@
 %\end{figure}
-
-Thus, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution, $\Gamma^{-1}(\alpha,\beta)$ with scale parameter $\beta=2Kr_{0}/\gamma$ and shape parameter $\alpha=(2+\gamma)/\gamma$, i.e.,
+Thus, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
 
@@
 \end{equation}
-
-whose mean equals the equilibrium solution of the unperturbed system,
-
+whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
 \begin{equation}
@@
 \end{equation}
-
 and whose variance, for $\gamma< 2$, is given by
-
 \begin{equation}
@@
 \end{equation}
-
-Note that the variance diverges for $\gamma\geq 2$.
-
-\vspace{\baselineskip}
+Note that the variance diverges, for $\gamma\geq 2$.
+The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.
 
 Let us consider now the general case of a time dependent rain input $r(t)$ with the following time-series of {\em observed} outputs,
-
 \begin{equation}\label{data}
-  y_i=\frac{1}{K}S(t_i)+\sigma\epsilon_i\,,\quad i=1,\dots,n\,,
+  y_i=\frac{1}{K}S(t_i)+\sigma\epsilon_i\,,\quad i=1,\dots,n+1\,,
 \end{equation}
-
-where the times $t_i$ are chosen so that $0=t_1<t_2<\dots < t_n=T$ and the $\epsilon_i$ are uncorrelated standard normal random variables.
+where the times $t_i$ are chosen so that $0=t_1<t_2<\dots < t_{n+1}=T$ and the $\epsilon_i$ are uncorrelated standard normal random variables.
 The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
-
 \begin{equation}\label{likelihood}
@@
     -\frac{1}{2}
-    \sum_{i=1}^n
+    \sum_{i=1}^{n+1}
     \frac
@@
 \end{equation}
-
-with the parameter vector $\bt=(K,\gamma)^T$. Choosing for the parameters $K$ and $\gamma$ the improper prior given by
+with the parameter vector $\bt=(K,\gamma)^T$.
+Choosing for the parameters $K$ and $\gamma$ the improper prior given by
 
@@
   \exp\bigg[
-    -{\tilde{\mathcal S}}[\dot{q}(t),q(t)]
+    -{{\mathcal S}}[\dot{q},q]
     -\frac{1}{2}
-    \sum_{i=1}^n
+    \sum_{i=1}^{n+1}
     \frac
@@
 
-while the action can be rewritten in the compact form,
+while the action is rewritten in the compact form,
 
 \begin{equation}\label{action_new}
-\tilde {\mathcal S}[\dot{q},q]
+{\mathcal S}[\dot{q},q]
 =
-\frac{1}{2T}
-\int_0^T \dt \left\{\left(
+\frac{1}{T}
+\int_0^T \dt \left\{
+\frac{1}{2}
+\left(
     T\dot q(t)
@@
     -
-    \frac{(2+\gamma)\beta^2}{2\gamma}
+    \frac{(2+\gamma)\beta^2}{4\gamma}
 \right\} \,,
@@
 
-In order to ease the numerical computations, it is useful to introduce a generalized time-dependent Hamiltonian,
+In order to ease the numerical computations, it is useful to introduce the time-dependent Hamiltonian,
 
@@
 =
-\frac{1}{2T}
+\frac{1}{T}
 \int_0^T dt\left\{
+    \frac{1}{2}
     T^2\dot q^2(t) +
+    \frac{1}{2}
     \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
-    2T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
-     \frac{(2+\gamma)\beta^2}{2\gamma}
+    T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
+     \frac{(2+\gamma)\beta^2}{4\gamma}
 \right\}
@@
 \\
-= \frac{1}{2T}
+= \frac{1}{T}
 \int_0^T dt\left\{
+    \frac{1}{2}
     T^2\dot q^2(t) +
+    \frac{1}{2}
     \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
-    2Tq(t)\dot\rho(t) -
-    \frac{(2+\gamma)\beta^2}{2\gamma}
+    Tq(t)\dot\rho(t) -
+    \frac{(2+\gamma)\beta^2}{4\gamma}
 \right\}
@@
    =
-   \int \frac{ p^2(t)}{2m_q}dt
-   + \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,
+   \int_0^T \frac{ p^2(t)}{2m_q}dt
+   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
 \end{equation}
 
-where in our case $s=2$, and $V( q,\bt)$ is the logarithm of the kernel of (\ref{posterior_pathint}).
+where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
 The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps
@@
 
-In particular, step 2 requires the calculation of the gradient of (\ref{Hamiltonian}), which is a somewhat tedious manual labour.
-Before we set out to do so, let us first write down the proper discretization of the path-integral (\ref{posterior_pathint}), with the action (\ref{action_final}).
-Therefore, let us assume that the measurement time points $\left\{ y_i \right\}_{i=1,\dots, n}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$.
-We further partition this interval into $N$ bins, with $N >> 1$, according to the time series $0=t_1<t_2,\dots<t_N=T$, such that the time points $t_{a_{\kappa}}$ with indexes $a_{\kappa} = \kappa (N-1)/n + 1$, where $\kappa = 1,\dots, n$, coincide with the data points $y_i$.
-Under these assumptions the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ read
+Let us first write down the proper discretization of the path-integral (\ref{posterior_pathint}), with the action (\ref{action_final}).
+Therefore, let us assume that the measurement time points $\left\{ y_i \right\}_{i=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
+Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
+The discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ read
 
@@
    K( p,{\pmb\pi})
-   &=
-   \sum_{i=1}^N \frac{ p_i^2}{2m_q}\Delta t
-   + \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
+   &\approx
+   \sum_{i=1}^N
+   \frac{ p_i^2}{2m_q}\Delta t
+   +
+   \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
    \\
@@
    &\approx
+   \frac{\Delta t}{T}
    \sum_{i=2}^{N}
-  \frac{K}{2\gamma\Delta t}
-  (q_i-q_{i-1})^2+\sum_{i=2}^{N-1} \frac{\Delta t}{\gamma} \left\{\frac{(\rho_i-e^{-q_i})^2}{2 K}  - q_i \frac{(\rho_i-\rho_{i-1})}{\Delta t} \right\}
+   \left\{
+    \frac{1}{2} T^2 \dot q_i^2
+    +
+   \left(
+        \frac{1}{2}
+        \left(
+            \rho_i-\frac{\beta}{\gamma}e^{-\beta \bar q_i}
+        \right)^2
+        -
+        T\bar q_i\dot\rho_i
+    \right)
+   \right\}
   \\
-  &- \frac{T (\gamma + 2)}{4 K}+\frac{\left(e^{-q_N}+q_N \rho_{N-1}- e^{-q_1}-q_1 \rho_{1} \right)}{\gamma} \nonumber
+  &-
+  \frac{(2+\gamma)\beta^2}{4\gamma}
+  +
+  \frac{1}{\gamma}
+  e^{-\beta q_N}
+  +
+  q_N \rho_{N-1}
+  -
+  \frac{1}{\gamma}
+  e^{-\beta q_1}
+  -
+  q_1 \rho_{1}
+  \nonumber
   \\
-  & + \sum_{\kappa=1}^n \frac{(y_\kappa-r_{a_{\kappa}} e^{q_{a_{\kappa}}})^2}{2\sigma^2} +\ln \left(K^{1-N/2} \gamma^{1+N/2}\right)\nonumber
+  & +
+  \sum_{s=1}^{n+1}
+  \frac{(y_s-r_{(s-1)j+1} e^{\beta q_{(s-1)j+1}})^2}{2\sigma^2}
+  % +\ln \left(K^{1-N/2} \gamma^{1+N/2}\right)\nonumber
    \,,\label{Vdisc}
 \end{align}
+with
+\begin{equation}
+  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,\quad
+  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
+\end{equation}
+and
+\begin{equation}\label{rhodisc}
+\rho_i = K \frac{\ln(r_{i+1}/r_i)}{\Delta t}+\gamma+1\,,\quad
+\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
+\end{equation}
 
-where the last term comes from the path measure (\ref{pathmeasure_q}) and
-
-\begin{equation}\label{rhodisc}
-\rho_i = K \frac{\ln(r_{i+1}/r_i)}{\Delta t}+\gamma+1\,.
-\end{equation}
+Note that we've neglected terms of order $\mathcal O(1/N)$ in the action (\ref{Vdisc}) and that we've assumed $r(t)$ to be at least twice differentiable.
 
@@
 Calculating the potential requires then an inversion of a Toeplitz matrix.
-This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes.
+This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes (see \cite{doll_1985_fourier}).
 
-Another approach is the so-called \emph{staging}, described in Tuckerman et al. (1993). This approach is interesting if different time-scales are involved.
+We choose another approach, the so-called \emph{staging} as described in Tuckerman et al. (1993), and diagonalize the harmonic part only in between the measurement points.
 We first note that the Hamiltonian $\mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})$ (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
+Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom. 
 Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.
-Thus, let us place $n+1$ equidistant (data) beads on the time domain, the first sitting on $t=0$ and the last on $t=T$.
-In between, we place $n(j-1)$ equidistant staging beads.
-So that, in total, we have $nj+1=N$ equidistant beads.
 
-The discretized harmonic part of the action can be rewritten as
+The discretized harmonic part of the action is rewritten as
 
@@
   \sum_{i=2}^{N}
-  \frac{K}{2\gamma\Delta t}
+  \frac{T}{2\Delta t}
   (q_i-q_{i-1})^2
   \\ =
-  \frac{K}{2\gamma}
+  \frac{T}{2}
   \sum_{s=1}^{n}\left\{
@@
 
-Moreover, we introduce the following transformation for the parameters,
+%Moreover, we introduce the following transformation for the parameters,
+%
+%\begin{equation}\label{params_transform}
+%\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
+%\end{equation}
+%
+%and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
+%
+This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal H_{HMC}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
+We write
+\begin{equation}
+  \mathcal H_{HMC}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
+\end{equation}
+where
+\begin{align}
+  \mathcal H_N
+  &=
+  \frac{1}{2}
+  \sum_{s=1}^{n}
+  \sum_{k=2}^j
+  \left\{
+    \frac{\Delta t}{m_q}p_{(s-1)j+k}^2
+    +
+    \frac{Tk}{\Delta t(k-1)}
+    u_{(s-1)j+k}^2
+  \right\}\,,\label{H_N}
+  \\
+  \mathcal H_n
+  &=
+  \frac{1}{2}
+  \sum_{s=1}^{n+1}
+  \left\{
+   \frac{\Delta t }{m_q}p_{(s-1)j+1}^2
+    +
+    \frac{T}{j\Delta t}
+    (u_{(s-1)j+1} - u_{sj+1})^2
+    -
+    \frac
+    {(y_s-r_{(s-1)j+1}e^{u_{(s-1)j+1}})^2}
+    {\sigma^2}
+   \right\}\,,\\
+  \mathcal H_1
+  &=
+   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}
+   +
+  \frac{\Delta t}{T}
+   \sum_{i=2}^{N}
+   \left\{
+       \frac{1}{2}
+        \left(
+            \rho_i-\frac{\beta}{\gamma}e^{-\beta \bar q_i}
+        \right)^2
+        -
+        T\bar q_i\dot\rho_i
+   \right\}
+  \\
+  &-
+  \frac{(2+\gamma)\beta^2}{4\gamma}
+  +
+  \frac{1}{\gamma}
+  e^{-\beta q_N}
+  +
+  q_N \rho_{N-1}
+  -
+  \frac{1}{\gamma}
+  e^{-\beta q_1}
+  -
+  q_1 \rho_{1}
+\,.
+\end{align}
+The harmonic part $\mathcal H_N$, for the staging beads, scales linearly with $N$.
+The harmonic term $\mathcal H_n$, for the boundary beads, as well as the measurement term scale linearly with $n$.
+Finally, the potential does not scale with neither $n$ nor $N$.
+Thus, depending on the values of $N$ and $n$, the dynamics happens on very different time scales.
+Therefore, we use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
+In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes.
+\begin{enumerate}
+  \item[i]
+  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$
+  \item[ii]
+  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$
+  \item[iii]
+  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$
+\end{enumerate}
 
-\begin{equation}\label{params_transform}
-\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
+In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
+\begin{equation}
+  \mathcal H_{HMC}=\mathcal H_N + \mathcal H'\,.
+\end{equation}
+In order to design a reversible integrator, for the associated Hamiltonian equations, we define the Liouville operators
+\begin{equation}
+  iL_N=\{\cdot\,,\,\mathcal H_N\}\,,\quad
+  iL'=\{\cdot\,,\,\mathcal H'\}\,,
+\end{equation}
+where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
+Trotter's formula allows us to write the Hamiltonian propagator as
+\begin{equation}
+  e^{i(L_N+L')\tau}
+  =
+  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
+  +
+  \mathcal O(P^{-2})\,,
+\end{equation}
+for $\tau =P\Delta \tau$.
+In regime (ii) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
+However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
+Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
+\begin{equation}
+  m=m_q/\Delta t\,,\quad
+  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
+\end{equation}
+The fast outer propagator is then given by eqs.
+\begin{align}
+  u_{(s-1)j+k}(\Delta\tau/2)
+  &=
+  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
+  +
+  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
+  p_{(s-1)j+k}(\Delta\tau/2)
+  &=
+  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
+  -
+  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
+\end{align}
+for $k=2,\dots,j$.
+
+For the inner, slow propagator, we employ the velocity Verlet algorithm.
+For the boundary beads, it reads
+\begin{align}
+  u_{(s-1)j+1}(\Delta\tau)
+  &=
+  u_{(s-1)j+1}(0)
+  +
+  \frac{\Delta\tau}{m} p_{(s-1)j+1}(0)
+  +
+  \frac{\Delta \tau^2}{2m}
+  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
+  p_{(s-1)j+1}(\Delta\tau)
+  &=
+  p_{(s-1)j+1}(0)
+  +
+  \frac{\Delta\tau}{2}
+  (
+  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
+  +
+  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
+  )\,,
+\end{align}
+where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
+For the model parameters, analogous equations apply.
+When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
+Thus,
+\begin{equation}
+  p_{(s-1)j+k}(\Delta\tau)
+  =
+  p_{(s-1)j+k}(0)
+  +
+  \frac{\Delta\tau}{2}
+  (
+  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
+  +
+  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
+  )
+  \,.
 \end{equation}
 
-and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-The potential $V(q,\tilde{\bt})$ can now be split into a fast component that scales linearly or quadratically with $N$ or $n$,
-
-\begin{multline}
-  {\mathcal V}_f(\vc u,\tilde{\bt})
-  =
-  \frac{N e^{\beta-\tau}}{2}
-  \sum_{s=1}^n\left\{
-    \frac{(u_{(s-1)j+1} - u_{sj+1})^2}{j}
-    +
-    \sum_{k=2}^j
-    \frac{k}{(k-1)}
-    u_{(s-1)j+k}^2
-  \right\}
-  \\
-  +
-  \frac{N}{2}(\beta-\tau)
-      -
-    \sum_{s=0}^{n}
-    \frac
-    {(y_s-r_{sj+1}e^{u_{sj+1}})^2}
-    {2 \sigma^2}
-  \,,
-\end{multline}
-
-where we have used $T = N \Delta t$, and a slow $N-$ and $n-$independent component,
-
-\begin{multline}
-{\mathcal V}_s(\vc q,\tilde{\bt})
-=
-    \sum_{i=2}^{N-1}
-        \left\{
-            \frac{e^{-(\beta+\tau)}}{2N}
-            (\rho_i-e^{-q_i})^2 - e^{-\tau}
-            q_i (\rho_i-\rho_{i-1})
-        \right\}-\frac{(2+e^\tau)e^{-\beta}}{4}
-        \\
-    + e^{-\tau} \left( e^{-q_N} + q_N\rho_{N-1}
-    - e^{-q_1} - q_1\rho_1\right)
-\,.
-\end{multline}
-
-In the expressions above we have included the Jacobian associated with (\ref{params_transform}).
-
-\vspace{\baselineskip}
-
-We can now fully exploit the two different time-scales that are involved in the problem.
+%We can now fully exploit the two different time-scales that are involved in the problem.
 First, the harmonic part of the fast component ${\mathcal V}_f[\dot{\vc u},\vc u]$ can be seen as describing the dynamics of heavy coupled \emph{boundary} data-point beads and lighter uncoupled \emph{staging} beads.
@@
 These values are used in (\ref{Kdisc}) for the masses $m_{q}$.
-Second, the two different ranges of the forces associated with ${\mathcal V}_f(\vc u,\tilde{\bt})$ and ${\mathcal V}_s(\vc q,\tilde{\bt})$ can be efficiently taken into account by applying a reversible reference system propagator algorithm (RESPA), as outlined in Tuckerman et al.
+%Second, the two different ranges of the forces associated with ${\mathcal V}_f(\vc u,\tilde{\bt})$ and ${\mathcal V}_s(\vc q,\tilde{\bt})$ can be efficiently taken into account by applying a reversible reference system propagator algorithm (RESPA), as outlined in Tuckerman et al.
 (1992).
@@
 
-The basic idea of the RESPA algorithm is to apply a Trotter decomposition to the system Liouville propagator $\mathcal{L}$ to generate a discrete time propagator factorized into fast and slow contributions.
+%The basic idea of the RESPA algorithm is to apply a Trotter decomposition to the system Liouville propagator $\mathcal{L}$ to generate a discrete time propagator factorized into fast and slow contributions.
 Indeed, as shown in Tuckerman et al.
@@
 
-\begin{equation}\label{G}
-  \mathcal{G}(\Delta \tau)=\exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\} \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\} \exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\}\,,
-\end{equation}
-
-where for the sake of clarity we have assumed a one-dimensional system.
+%\begin{equation}\label{G}
+%  \mathcal{G}(\Delta \tau)=\exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\} \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\} \exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\}\,,
+%\end{equation}
+%
+%where for the sake of clarity we have assumed a one-dimensional system.
 The argument can be easily extended to our multi-dimensional case.
@@
 On the other hand, in order to ensure stable dynamics, the short-range middle propagator is further factorized into $n_f$ smaller time steps of duration $\delta \tau = \Delta \tau / n_f$, according to the Trotter expansion
-
-\begin{equation}\label{G_fast}
-  \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\}
-  =\left[
-  \exp\left\{ \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
-  \exp\left\{  \delta\tau \dot{x} \frac{\partial}{\partial x}  \right\}
-  \exp\left\{  \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
-  \right]^{n_f} \,.
-\end{equation}
-
-Coordinates and conjugate momenta, i.e., $(q,\bt; p,{\pmb\pi})$ in our case, are propagated using (\ref{G}) and (\ref{G_fast}) iteratively by applying a velocity Verlet algorithm.
-
+%
+%\begin{equation}\label{G_fast}
+%  \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\}
+%  =\left[
+%  \exp\left\{ \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
+%  \exp\left\{  \delta\tau \dot{x} \frac{\partial}{\partial x}  \right\}
+%  \exp\left\{  \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
+%  \right]^{n_f} \,.
+%\end{equation}
+%
+%Coordinates and conjugate momenta, i.e., $(q,\bt; p,{\pmb\pi})$ in our case, are propagated using (\ref{G}) and (\ref{G_fast}) iteratively by applying a velocity Verlet algorithm.
+%
 \textbf{Some preliminary results}.
@@
 \bibliographystyle{apalike} % chicago \shortcite(N)
-\bibliography{C:/Users/ulzegasi/SWITCHdrive/refs}
+\bibliography{C:/Users/albertca/SWITCHdrive/refs}
 

### 2015-03-09 Main_HMC_3_2.jl
backup update

@@ -1,6 +1,5 @@
 ## ============================================================================================
 ## Project: Parameter Inference for a Simple Stochastic Hydrological Model
-## File: Main_HMC_1.jl
 ##
 ## Description: Single bucket, scale invariant noise, no evaporation yet
 ##              dS(t)/dt = r(t) - S(t)/k + sqrt(gamma/k) S(t) eta(t)
@@ -14,7 +13,9 @@
 ##
 ## NEW IN VERSION 2. THE PRIOR FOR PARAMETER K IS A LOG-NORMAL DISTRIBUTION.
 ##                   We use the tentative function f(K) = (3 Exp(-(Log(K)-4)^2/2))/K 
-## NEW IN VERSION 3. We use the reverse diff package 
+## NEW IN VERSION 3. We use the reverse diff package for automated differentiation w.r.t. 
+##                   both parameters and coordinates.
+##                   Both fast and slow potentials (and their derivatives) are fnctions of {u} only. 
 ##             
 ## GitHub repository: https://github.com/ulzegasi/Julia_ParInf.git
 ##
@@ -150,7 +151,7 @@ reload("$dir/ParInf_Fun_AD_3_2.jl")
 ## --------------------------------------------------------------------------------------------
 
 M_bdy   = 6.0*K*N/(gamma*T)
-M_theta = M_bdy*0.02
+M_theta = M_bdy/N
 M_stage = M_bdy*0.01
 
 mp[1:s] = M_theta
@@ -230,7 +231,7 @@ M_theta_burnin = M_theta
 M_stage_burnin = M_stage
 
 M_bdy   = 6.0*K*N/(gamma*T)
-M_theta = M_bdy*[0.03, 0.03]						    # Masses for K and gamma
+M_theta = M_bdy./[1.0*N, 1.0*N]						    # Masses for K and gamma
 M_stage = M_bdy*0.02
 
 mp = zeros(Float64, s+N)
@@ -249,6 +250,9 @@ reject_counter = 0
 
 time_Hold       = Array(Float64,nsample_eff)
 time_respa      = Array(Float64,nsample_eff)
+time_respa_s    = zeros(nsample_eff)
+time_respa_f_u  = zeros(nsample_eff)
+time_respa_f_d  = zeros(nsample_eff)
 time_Hnew       = Array(Float64,nsample_eff)
 time_metropolis = Array(Float64,nsample_eff) 
 
@@ -281,7 +285,7 @@ for counter = (nsample_burnin + 1):nsample
     t1=time()
[…]

### 2015-03-09 ParInf_Fun_AD_3_2.jl

@@ -220,24 +220,25 @@ V_slow_der_u     = rdiff(V_slow, outsym=:out_slow, u = ones(Float64, N) )
 ## RESPA
 ##########################################################################
 
-function RESPA(theta, u, p, mp, dtau, nn, deltatau)     # (Tuckerman et al., JCP 97(3), 1992 )
+function RESPA(theta, u, p, mp, dtau, nn, deltatau, counter)     # (Tuckerman et al., JCP 97(3), 1992 )
     
 
     # First long-range step (dtau):
     # ---------------------------
-    
+    time1 = time()
     force = -V_slow_der(theta,u)
     for i=1:(N+s) 
         p[i] = p[i]  + (dtau/2)* force[i]
     end
-
+    time_respa_s[counter-nsample_burnin] += (time()-time1) 
     # Short-range steps (nn*deltatau):
     # ---------------------------
     
     for counter_fast_respa = 1:nn                       # Verlet integrator 
-        
+        timefd = time()
         force_old = -V_fast_der(theta,u)                # (Tuckerman et al., JCP 97(3), 1992, eq. 2.17)
-
+        time_respa_f_d[counter-nsample_burnin] += (time()-timefd) 
+        timefu = time()
         for i=1:s
             theta[i] = theta[i] + deltatau * ( p[i]  + (deltatau/2) * force_old[i] )/mp[i]
         end
@@ -245,12 +246,15 @@ function RESPA(theta, u, p, mp, dtau, nn, deltatau)     # (Tuckerman et al., JCP
             u[i] = u[i]  + deltatau * ( p[i+s]  + (deltatau/2) * force_old[i+s] )/mp[i+s]
         end
         
+        time_respa_f_u[counter-nsample_burnin] += (time()-timefu)
+        timefd = time()
         force_new = -V_fast_der(theta,u)
-
+        time_respa_f_d[counter-nsample_burnin] += (time()-timefd)
         for i=1:(N+s) 
             p[i] = p[i]  + (deltatau/2)*( force_old[i] + force_new[i] )
         end
     end
+
     
     # Back transformations (u -> q) to update the {q} variables :
     # (Tuckerman et al., JCP 99 (1993), eq. 2.19)
@@ -270,9 +274,10 @@ function RESPA(theta, u, p, mp, dtau, nn, deltatau)     # (Tuckerman et al., JCP
 
     # Long-range step:
     # ---------------------------
+    time2 = time()
[…]

### 2015-03-09 Analysis_HMC_3.jl
backup update 2

@@ -1,0 +1,276 @@
+## ============================================================================================
+## Project: Parameter Inference for a Simple Stochastic Hydrological Model
+## File: Analysis_HMC_1.jl
+##
+## Description: Single bucket, scale invariant noise, no evaporation yet
+##              dS(t)/dt = r(t) - S(t)/k + sqrt(gamma/k) S(t) eta(t)
+## In Main:------------------------------------------------------------------------------------
+## The {q} variable set is used for slow (long-range) components. 
+## The {u} variable set is used for fast (short-range) components.
+## 
+## The derivative of the potential V(theta,u/q) w.r.t. theta can be calculated explicitely
+## or by automated differentiation (ForwardDiff package). 
+## 
+## Different masses are used for the burn-in phase.
+## --------------------------------------------------------------------------------------------            
+## GitHub repository: https://github.com/ulzegasi/Julia_ParInf.git
+##
+## Authors: Simone Ulzega, Carlo Albert
+##
+## simone.ulzega@eawag.ch
+## carlo.albert@eawag.ch
+## ============================================================================================
+##
+##
+
+dir      = "C:/Users/ulzegasi/SWITCHdrive/JuliaTemp/Data"
+fname    = string("_N401")
+
+##
+##
+## ============================================================================================
+## Get stored parameters and data
+## ============================================================================================
+##
+##
+
+params   = readdlm("$dir/params$fname.dat")
+
+N              = int64(params[findin(params[:,1],["N"]),2])[1]
+j              = int64(params[findin(params[:,1],["j"]),2])[1]
+n              = int64(params[findin(params[:,1],["n"]),2])[1] 
+s 			   = int64(params[findin(params[:,1],["s"]),2])[1] 
+nsample_burnin = int64(params[findin(params[:,1],["nsample_burnin"]),2])[1]
+nsample_eff    = int64(params[findin(params[:,1],["nsample_eff"]),2])[1]  
[…]

### 2015-03-09 Main_HMC_3_2_orders.jl

@@ -1,0 +1,420 @@
+## ============================================================================================
+## Project: Parameter Inference for a Simple Stochastic Hydrological Model
+## 
+## Description: Single bucket, scale invariant noise, no evaporation yet
+##              dS(t)/dt = r(t) - S(t)/k + sqrt(gamma/k) S(t) eta(t)
+## The {q} variable set is used for slow (long-range) components. (NOT ANY MORE TRUE!)
+## The {u} variable set is used for fast (short-range) components.
+## 
+## The derivative of the potential V(theta,u/q) w.r.t. theta can be calculated explicitely
+## or by automated differentiation (ForwardDiff package). 
+## 
+## Different masses are used for the burn-in phase.
+##
+## NEW IN VERSION 2. THE PRIOR FOR PARAMETER K IS A LOG-NORMAL DISTRIBUTION.
+##                   We use the tentative function f(K) = (3 Exp(-(Log(K)-4)^2/2))/K 
+## NEW IN VERSION 3. We use the reverse diff package for automated differentiation w.r.t. 
+##                   both parameters and coordinates.
+##                   Both fast and slow potentials (and their derivatives) are fnctions of {u} only. 
+##             
+## GitHub repository: https://github.com/ulzegasi/Julia_ParInf.git
+##
+## Authors: Simone Ulzega, Carlo Albert
+##
+## simone.ulzega@eawag.ch
+## carlo.albert@eawag.ch
+## ============================================================================================
+dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC"       # Main project directory, local repository   
+dir2  = "C:/Users/ulzegasi/SWITCHdrive/JuliaTemp/Data"   # Secondary directory (e.g., data storage)
+dir3  = "C:/Users/ulzegasi/SWITCHdrive/JuliaTemp/Orders"
+using ReverseDiffSource  #ForwardDiff                    # Automated differentiation package
+##
+## ============================================================================================
+## System parameters
+## ============================================================================================
+##
+range = 2:5002
+t  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
+
+const n = 10             # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
+const j = 40             # n(j-1) -> total number of staging beads, j-1 -> staging beads per segment
+const N = int64(n*j+1)   # Total number of discrete time points
[…]

### 2015-03-09 Main_HMC_4.jl

@@ -1,0 +1,436 @@
+## ============================================================================================
+## Project: Parameter Inference for a Simple Stochastic Hydrological Model
+##
+## Description: Single bucket, scale invariant noise, no evaporation yet
+##              dS(t)/dt = r(t) - S(t)/k + sqrt(gamma/k) S(t) eta(t)
+## The {q} variable set is used for slow (long-range) components. (NOT ANY MORE TRUE!)
+## The {u} variable set is used for fast (short-range) components.
+## 
+## The derivative of the potential V(theta,u/q) w.r.t. theta can be calculated explicitely
+## or by automated differentiation (ForwardDiff package). 
+## 
+## Different masses are used for the burn-in phase.
+##
+## NEW IN VERSION 2. THE PRIOR FOR PARAMETER K IS A LOG-NORMAL DISTRIBUTION.
+##                   We use the tentative function f(K) = (3 Exp(-(Log(K)-4)^2/2))/K 
+## NEW IN VERSION 3. We use the REVERSE DIFF package for AUTOMATED DIFFERENTIATION w.r.t. 
+##                   both parameters and coordinates.
+##                   Both fast and slow potentials (and their derivatives) are FUNCTIONS OF {u} ONLY. 
+## NEW IN VERSION 4. PARALLEL.
+##             
+## GitHub repository: https://github.com/ulzegasi/Julia_ParInf.git
+##
+## Authors: Simone Ulzega, Carlo Albert
+##
+## simone.ulzega@eawag.ch
+## carlo.albert@eawag.ch
+## ============================================================================================
+addprocs(2)
+num_workers  = nworkers()
+first_worker = workers()[1]
+last_worker  = workers()[end]
+
+dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC"       # Main project directory, local repository   
+dir2  = "C:/Users/ulzegasi/SWITCHdrive/JuliaTemp/Data"   # Secondary directory (e.g., data storage)
+fname = string("_testing")                               # Output files
+@everywhere using ReverseDiffSource #, ForwardDiff       # Automated differentiation package
+##
+## ============================================================================================
+## System parameters
+## ============================================================================================
+##
+range = 2:5002
+t  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
+
+const n = 10             # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
[…]

### 2015-03-09 Main_HMC_4_2.jl

@@ -1,0 +1,504 @@
+## ============================================================================================
+## Project: Parameter Inference for a Simple Stochastic Hydrological Model
+##
+## Description: Single bucket, scale invariant noise, no evaporation yet
+##              dS(t)/dt = r(t) - S(t)/k + sqrt(gamma/k) S(t) eta(t)
+## The {q} variable set is used for slow (long-range) components. (NOT ANY MORE TRUE!)
+## The {u} variable set is used for fast (short-range) components.
+## 
+## The derivative of the potential V(theta,u/q) w.r.t. theta can be calculated explicitely
+## or by automated differentiation (ForwardDiff package). 
+## 
+## Different masses are used for the burn-in phase.
+##
+## NEW IN VERSION 2. THE PRIOR FOR PARAMETER K IS A LOG-NORMAL DISTRIBUTION.
+##                   We use the tentative function f(K) = (3 Exp(-(Log(K)-4)^2/2))/K 
+## NEW IN VERSION 3. We use the REVERSE DIFF package for AUTOMATED DIFFERENTIATION w.r.t. 
+##                   both parameters and coordinates.
+##                   Both fast and slow potentials (and their derivatives) are FUNCTIONS OF {u} ONLY. 
+## NEW IN VERSION 4. PARALLEL.
+##             
+## GitHub repository: https://github.com/ulzegasi/Julia_ParInf.git
+##
+## Authors: Simone Ulzega, Carlo Albert
+##
+## simone.ulzega@eawag.ch
+## carlo.albert@eawag.ch
+## ============================================================================================
+addprocs(3)
+num_workers  = nworkers()
+first_worker = workers()[1]
+last_worker  = workers()[end]
+
+dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC"       # Main project directory, local repository   
+dir2  = "C:/Users/ulzegasi/SWITCHdrive/JuliaTemp/Data"   # Secondary directory (e.g., data storage)
+fname = string("_testing")                               # Output files
+@everywhere using ReverseDiffSource #, ForwardDiff       # Automated differentiation package
+reload("$dir/myDArray.jl")
+##
+## ============================================================================================
+## System parameters
+## ============================================================================================
+##
+trange = 2:5002
+t  = float64(readdlm("$dir/t.dat")[trange,2]) # Time points t.dat
+
+const n = 10             # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
[…]

### 2015-03-09 ParInf_Fun_AD_4.jl

@@ -1,0 +1,445 @@
+##########################################################################
+## Fast potential as a function
+##########################################################################
+
+function V_fast_fun(theta,u)    # Fast dynamics (harmonic and data terms)
+    
+    # Variables:
+    # ---------------------------
+    
+    out = 0.0
+
+    # beta    = theta[1]
+    # tau     = theta[2]
+    
+    K  = T*exp(theta[1])
+    g  = exp(theta[2])
+
+    out = ( y[1] - r[1]*exp(u[1]) )^2/(2*sigma^2)
+    for i=1:n
+        out += K * ( u[(i-1)*j+1] - u[i*j+1] )^2/( j*dt )   / ( 2*g ) +
+                    ( y[i+1] - r[i*j+1]*exp(u[i*j+1]) )^2/(2*sigma^2)
+        for k=2:j
+            out += k*u[(i-1)*j+k]^2/( (k-1)*dt ) * K/(2*g)
+        end
+    end
+
+    return out
+end
+
+##
+##
+# Calculate derivatives of V_fast w.r.t. the parameters theta:
+# --------------------------- 
+# V_fast_fun_theta      = function(theta) V_fast_fun(theta,u) end
+# V_fast_fun_der_theta  = forwarddiff_gradient(V_fast_fun_theta, Float64, fadtype=:dual, n=s)
+
+# V_fast_fun_u      = function(u) V_fast_fun(theta,u) end
+# V_fast_fun_der_u  = forwarddiff_gradient(V_fast_fun_u, Float64, fadtype=:dual, n=N)
+##
+##
+
+##########################################################################
+## Fast potential as an expression (-> array of expressions)
+##########################################################################
+
+V_fast            = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+V_fast_der_u_expr = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+V_fast_der_t_expr = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+
+@spawnat first_worker ( 
+
+    firstind = proc_indexes[1][1];
+    lastind  = proc_indexes[1][end];
+
+    localpart(V_fast)[1] = quote
+        K = T*exp(theta[1])
+        g =   exp(theta[2])
+
+        out_fast = ( y[1] - r[1]*exp(u[1]) )^2/(2*sigma^2)
[…]

### 2015-03-09 ParInf_Fun_AD_4_2.jl

@@ -1,0 +1,456 @@
+##########################################################################
+## Fast potential as a function
+##########################################################################
+
+function V_fast_fun(theta,u)    # Fast dynamics (harmonic and data terms)
+    
+    # Variables:
+    # ---------------------------
+    
+    out = 0.0
+
+    # beta    = theta[1]
+    # tau     = theta[2]
+    
+    K  = T*exp(theta[1])
+    g  = exp(theta[2])
+
+    out = ( y[1] - r[1]*exp(u[1]) )^2/(2*sigma^2)
+    for i=1:n
+        out += K * ( u[(i-1)*j+1] - u[i*j+1] )^2/( j*dt )   / ( 2*g ) +
+                    ( y[i+1] - r[i*j+1]*exp(u[i*j+1]) )^2/(2*sigma^2)
+        for k=2:j
+            out += k*u[(i-1)*j+k]^2/( (k-1)*dt ) * K/(2*g)
+        end
+    end
+
+    return out
+end
+
+##
+##
+# Calculate derivatives of V_fast w.r.t. the parameters theta:
+# --------------------------- 
+# V_fast_fun_theta      = function(theta) V_fast_fun(theta,u) end
+# V_fast_fun_der_theta  = forwarddiff_gradient(V_fast_fun_theta, Float64, fadtype=:dual, n=s)
+
+# V_fast_fun_u      = function(u) V_fast_fun(theta,u) end
+# V_fast_fun_der_u  = forwarddiff_gradient(V_fast_fun_u, Float64, fadtype=:dual, n=N)
+##
+##
+
+##########################################################################
+## Fast potential as an expression (-> array of expressions)
+##########################################################################
+
+V_fast            = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+V_fast_der_u_expr = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+V_fast_der_t_expr = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nworkers()])
+
+@spawnat first_worker ( 
+
+    # firstind = bdy_indexes[1][1];
+    # lastind  = bdy_indexes[1][end];
+    lastind  = length(bdy_indexes[1]);
+    # du_local = dulocal;
+    # dr_local = drlocal;
+    # dy_local = dylocal;
+
[…]

### 2015-03-09 myDArray.jl

@@ -1,0 +1,27 @@
+## chk_indexes is an array of tuples defining the lengths of the chunks on each process
+## ex:
+## Out [46]: 3-element Array{Any,1}:
+## (1:9,)
+## (10:14,)
+## (15:23,)
+##
+
+function myDArray(init, dims, procs, dist, chk_indexes)
+    np      = prod(dist)
+    procs   = procs[1:np]
+    cuts    = [[chk_indexes[i][1][1] for i = 1:num_workers] for ii = 1:1]
+    idxs    = chk_indexes
+    chunks  = Array(RemoteRef, dist...)
+    for i = 1:np
+        chunks[i] = remotecall(procs[i], init, idxs[i])
+    end
+    p = 0
+    for i = 1:length(procs)
+        if procs[i] == myid()
+            p = i
+        end
+    end
+    p = max(1 , p)
+    A = remotecall_fetch(procs[p], r->typeof(fetch(r)), chunks[p])
+    DArray{eltype(A),length(dims),A}(dims, chunks, procs, idxs, cuts)
+end

### 2015-03-13 TeXFiles/stochInf1D.tex
q variables from the very beginning

@@
 \end{equation}
-Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.
-The scale-invariance of the noise term renders the noise parameter dimensionless.
 Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
 For computational reasons, we work with the {\em Stratonovich} convention.
-Our parametrization is such that, for constant rain input $r(t)=r$ and in the long-time limit, the mean of $S$ converges to the limit of the unperturbed ($\gamma=0$) system $Kr$ (see below).
+Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{eq}=Kr_0$ (see below).
 
-According to \cite{lau_2007_StateDepDiff}, the probability, $P(S,T|S_0,t_0)$, of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as
+The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
+Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
+\begin{equation}
+  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
+  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
+\end{equation}
+W.r.t. these new  variables and parameters, model equation (\ref{sde}) becomes
+\begin{equation}
+  \dot q(t)
+  =
+  \frac{\beta}{T\gamma}e^{-\beta q(t)}
+  -
+  \frac{1}{T}\rho(t)
+  +
+  \frac{1}{\sqrt{T}}\eta(t)\,,
+\end{equation}
+with
+\begin{equation}
+  \rho(t)
+  =
+  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
+  +
+  \frac{(2+\gamma)\beta}{2\gamma}\,.
+\end{equation}
+
+%Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.
+
+According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
 \begin{equation}\label{pathint}
-P(S,T|S_0,t_0) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
+P(q_1,T|q_0,0) 
+= 
+\frac{1}{Z}
+\int 
+e^{-{\mathcal S}[q,\dot q]} 
+\delta(q(T)-q_1)
+\delta(q(0)-q_0)
+\mathcal{D}q \,,
 \end{equation}
-where the integral extends over all paths starting at $S_0$ and ending at $S$.
-The path-measure $\mathcal DS$ is the formal limit
-\begin{equation}\label{pathmeasure}
-{\mathcal D}S
+where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
+The path-measure $\mathcal Dq$ is formally written as an infinite product
+\begin{equation}\label{pathmeasure_q}
+{\mathcal Dq}
 =
-\lim_{\Delta t\rightarrow 0}
-\prod_{i}
-\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,,
+\prod_{t}
+dq(t)\,.
 \end{equation}
-for a discretization $S_i = S(t_i)$ with $t_i-t_{i-1}=\Delta t$.
-The {\em action} is a functional defined on paths $S:[0,T]\longrightarrow \mathbb R_+$ and reads as
-\begin{equation}\label{action}
-  \tilde{\mathcal S}[\dot{S}(t),S(t)]
+The {\em action} is a functional on the space of paths and reads as
+%\begin{equation}\label{action}
+%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
+%  =
+%  \int_0^T
+%  dt
+%  \left\{
+%    \frac{K}{2\gamma S^2(t)}
+%    \left(
+%        \dot S(t)
+%        -
+%        r(t)
+%        +
+%        \frac{1+\gamma}{K}S(t)
+%    \right)^2
+%    -
+%    \frac{2+\gamma}{4K}
+%  \right\}
+% \,.
+%\end{equation}
+\begin{equation}
+{\mathcal S}[{q},\dot q]
 =
-  \int_0^T
-  dt
-  \left\{
-    \frac{K}{2\gamma S^2(t)}
+\frac{1}{T}
+\int_0^T \dt \left\{
+\frac{1}{2}
 \left(
-        \dot S(t)
+    T\dot q(t)
+    +
+    \rho(t)
     -
-        r(t)
+    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
+    -
+    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
+\right\} \,.
+\end{equation}
+Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.
+
+We introduce the time-dependent {\em Hamiltonian}
+\begin{equation}\label{H}
+  \mathcal{H}(q,t)
+  =
+  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
+\end{equation}
+and rewrite action (\ref{action}) as,
+\begin{multline}\label{action}
+{\mathcal S}[{q},\dot q]
+=
+\frac{1}{T}
+\int_0^T dt\left\{
+    \frac{1}{2}
+    T^2\dot q^2(t) +
+    \frac{1}{2}
+    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
+    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
+    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
+     %\frac{(2+\gamma)\beta^2}{4\gamma}
+\right\}
+\\
++ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
+\\
+= \frac{1}{T}
+\int_0^T dt\left\{
+    \frac{1}{2}
+    T^2\dot q^2(t) +
+    \frac{1}{2}
+    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
+    Tq(t)\dot\rho(t) -
+     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
+%    \frac{(2+\gamma)\beta^2}{4\gamma}
+\right\}
+\\
 +
-        \frac{1+\gamma}{K}S(t)
-    \right)^2
-    -
-    \frac{2+\gamma}{4K}
-  \right\}
+    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
+   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
 \,.
-\end{equation}
-Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S,T|S_0,t_0) \mathit{dS}}=1$.
+\end{multline}
 
-It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S,T|S_0,t_0)$ in the simple case of a constant rain input $r_{0}$.
+Next, we calculate the equilibrium distribution $P_{\text{eq}}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$.
 Systems at thermal equilibrium must fulfill detailed balance, i.e.,
 \begin{equation}\label{detailed_balance}
-P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,.
+P(q_1 t_1 | q_0 t_0 ) P_{\text{eq}}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\text{eq}}(q_1) \,.
 \end{equation}
-Assuming a solution of the form
-
+After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
 \begin{equation}\label{Peq}
-  P_{\text{eq}}(S)
+  P_{\text{eq}}(q)
   \propto
-  e^{-\mathcal{H}(S)}\,,
+  e^{-2\mathcal{H}(q)}\,.
 \end{equation}
-equation (\ref{detailed_balance}) is rewritten as
-
-\begin{equation}\label{detailed_balance_2}
-P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 )
-\exp\left\{
-        -\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{d\mathcal{H}}{dS}
-    \right\}
-\,,
-\end{equation}
-where $S(t)$ is an arbitrary path connecting $S_0$ with $S_1$.
-After plugging in (\ref{pathint}) and (\ref{action}), and using the Stratonovich property, which allows us to transform a path from $S_0$ to $S_1$ into a path from $S_1$ to $S_0$ through the transformation $S(t) \rightarrow S(-t)$, we get
-
-\begin{equation}\label{constant_rain_H}
--r_{0}+\frac{1+\gamma}{K}S
-=
-\frac{\gamma S^2}{2K}
-\frac{d \mathcal{H}(S)}{dS}\,.
-\end{equation}
-The solution to (\ref{constant_rain_H}) is, up to an additive constant, given by eq.
-\begin{equation}\label{solution_const_rain}
-\mathcal{H}(S)=\frac{2 K r_0}{\gamma S}+\frac{2(1+\gamma)}{\gamma}\ln(S)\,.
-\end{equation}
-%\begin{figure}
-%    \centering
-%    \includegraphics[width=0.6\textwidth]{Figs/InvGamma.png}
-%    \caption{Inverse gamma distributions with scale parameter $\beta = 5$ and shape parameters $\alpha = 3,4,5,6$.}
-%    \label{fig:inv_gamma}
-%\end{figure}
-Thus, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
-
+Transforming back to the original variables, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
 \begin{equation}\label{inverse_gamma}
@@
 
-Let us consider now the general case of a time dependent rain input $r(t)$ with the following time-series of {\em observed} outputs,
+Let us now consider two time-series of {\em observed} rain inputs and discharge outputs, $r_s=r(t_s)$ and $y_s=S(t_s)/K$, respectively, measured at times $0=t_1<t_2<\dots < t_{n+1}=T$.
+For the measurement error model, we simply set
 \begin{equation}\label{data}
-  y_i=\frac{1}{K}S(t_i)+\sigma\epsilon_i\,,\quad i=1,\dots,n+1\,,
+  \ln\frac{y_s}{r_s}
+  =
+  \beta q_s
+  +
+  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
 \end{equation}
-where the times $t_i$ are chosen so that $0=t_1<t_2<\dots < t_{n+1}=T$ and the $\epsilon_i$ are uncorrelated standard normal random variables.
-The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
-\begin{equation}\label{likelihood}
-  f(\vc y | \bt)
+where $\epsilon_s$ are uncorrelated standard normal random variables.
+
+%The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
+%\begin{equation}\label{likelihood}
+%  f(\vc y | \bt)
+%  \propto
+%  \int
+%  \exp\bigg[
+%    -{\tilde{\mathcal S}}[\dot{S},S]
+%    -\frac{1}{2}
+%    \sum_{i=1}^{n+1}
+%    \frac
+%    {(y_i-S(t_i)/K)^2}
+%    {\sigma^2}
+%  \bigg]
+%  {\mathcal D} S(\bt)\,,
+%\end{equation}
+%with the parameter vector $\bt=(K,\gamma)^T$.
+%Choosing for the parameters $K$ and $\gamma$ the improper prior given by
+%
+%\begin{equation}\label{prior}
+%  f(\bt)
+%  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
+%\end{equation}
+%
+%{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
+%\begin{equation}\label{posterior}
+%  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
+%\end{equation}
+
+The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
+For the time being we assume constant priors.
+Then, the posterior reads as
+\begin{equation}\label{posterior_pathint}
+  f(\bt | \vc y, \vc r)
   \propto
@@
   \exp\bigg[
-    -{\tilde{\mathcal S}}[\dot{S},S]
+    -{{\mathcal S}}[{q},\dot q]
     -\frac{1}{2}
-    \sum_{i=1}^{n+1}
+    \sum_{s=1}^{n+1}
     \frac
-    {(y_i-S(t_i)/K)^2}
-    {\sigma^2}
-  \bigg]
-  {\mathcal D} S(\bt)\,,
-\end{equation}
-with the parameter vector $\bt=(K,\gamma)^T$.
-Choosing for the parameters $K$ and $\gamma$ the improper prior given by
-
-\begin{equation}\label{prior}
-  f(\bt)
-  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
-\end{equation}
-
-{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
-\begin{equation}\label{posterior}
-  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
-\end{equation}
-
-We replace the variable $S(t)$ and the parameter $K$ by the dimensionless quantities, $q(t)$ and $\beta$, respectively, by means of the transformations
-\begin{equation}
-  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
-  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
-\end{equation}
-In terms of these dimensionless variables and parameters the posterior reads (without prior)
-\begin{equation}\label{posterior_pathint}
-  f(\bt | \vc y)
-  \propto
-  \int
-  \exp\bigg[
-    -{{\mathcal S}}[\dot{q},q]
-    -\frac{1}{2}
-    \sum_{i=1}^{n+1}
-    \frac
-    {(y_i-r_ie^{\beta q_i})^2}
+    {(\ln(y_s/r_s)-{\beta q_s})^2}
     {\sigma^2}
@@
   {\mathcal Dq}
-  \,,
+  \,.
 \end{equation}
-%\prod_{t}
-%\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
-where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,
-
-\begin{equation}\label{pathmeasure_q}
-{\mathcal Dq}
-=
-\prod_{t}
-dq(t)\,,
-\end{equation}
-
-while the action is rewritten in the compact form,
-
-\begin{equation}\label{action_new}
-{\mathcal S}[\dot{q},q]
-=
-\frac{1}{T}
-\int_0^T \dt \left\{
-\frac{1}{2}
-\left(
-    T\dot q(t)
-    +
-    \rho(t)
-    -
-    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
-    -
-    \frac{(2+\gamma)\beta^2}{4\gamma}
-\right\} \,,
-\end{equation}
-
-with
-\begin{equation}\label{rho}
-\rho(t)=\frac{T\dot r(t)}{\beta r(t)}+\frac{\beta(\gamma+1)}{\gamma}
-=
-\frac{T}{\beta}\frac{d \left[\ln r(t)\right]}{dt}+\frac{\beta(\gamma+1)}{\gamma}\,.
-\end{equation}
-
-In order to ease the numerical computations, it is useful to introduce the time-dependent Hamiltonian,
-
-\begin{equation}\label{H}
-  \mathcal{H}_q(q,t)
-  =
-  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
-\end{equation}
-
-and rewrite the action (\ref{action_new}) as,
-
-\begin{multline}\label{action_final}
-{\mathcal S}[\dot{q},q]
-=
-\frac{1}{T}
-\int_0^T dt\left\{
-    \frac{1}{2}
-    T^2\dot q^2(t) +
-    \frac{1}{2}
-    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
-    T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
-     \frac{(2+\gamma)\beta^2}{4\gamma}
-\right\}
-\\
-+ \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
-\\
-= \frac{1}{T}
-\int_0^T dt\left\{
-    \frac{1}{2}
-    T^2\dot q^2(t) +
-    \frac{1}{2}
-    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
-    Tq(t)\dot\rho(t) -
-    \frac{(2+\gamma)\beta^2}{4\gamma}
-\right\}
-\\
-+
-    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
-   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
-\,.
-\end{multline}
-
 Since the integral on the r.h.s.
-of (\ref{posterior_pathint}) is very expensive to calculate, we will sample simultaneously parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
-Since this distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
+of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
+Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
 This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system.
 Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
-
 \begin{equation}\label{Hamiltonian}
@@
 \end{equation}
-
-where
-
+with
 \begin{equation}\label{K}
@@
 \end{equation}
+where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
+The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
+\begin{equation}\label{phaseSpacePathInt}
+    f(\bt | \vc y, \vc r)
+  \propto
+  \int
+  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
+  {\mathcal Dp}
+  \wedge
+  {\mathcal Dq}
+  \wedge
+  d{\pmb\pi}
+  \,.
+\end{equation}
 
-where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
 The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps
-
 \begin{enumerate}
@@
   \item
-  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically (leap-frog algorithm).
+  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
   \item
@@
 \end{enumerate}
+The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.
 
-The last step is the standard Metropolis algorithm, while the first two steps allow to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.
-
-Let us first write down the proper discretization of the path-integral (\ref{posterior_pathint}), with the action (\ref{action_final}).
-Therefore, let us assume that the measurement time points $\left\{ y_i \right\}_{i=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
+Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
+Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
 Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
-The discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ read
-
+Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure 
+\begin{equation}
+  \mathcal Dp\wedge\mathcal Dq
+  \approx
+  \prod_i dp_i\wedge dq_i 
+\end{equation}
+and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
 \begin{align}
@@
    +
-   \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
+   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
    \\
@@
    \left\{
-    \frac{1}{2} T^2 \dot q_i^2
+    \frac{1}{2} 
+    T^2 \dot q_i^2
     +
-   \left(
     \frac{1}{2}
      \left(
-            \rho_i-\frac{\beta}{\gamma}e^{-\beta \bar q_i}
+        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
      \right)^2
     -
-        T\bar q_i\dot\rho_i
-    \right)
+    \frac{\beta^2}{2\gamma}
+    e^{-\beta q_i}
+   -
+    T q_i\dot\rho_i
    \right\}
   \\
-  &-
-  \frac{(2+\gamma)\beta^2}{4\gamma}
-  +
+  &+
   \frac{1}{\gamma}
@@
   +
-  q_N \rho_{N-1}
+  q_N \rho_{N}
   -
@@
   \sum_{s=1}^{n+1}
-  \frac{(y_s-r_{(s-1)j+1} e^{\beta q_{(s-1)j+1}})^2}{2\sigma^2}
-  % +\ln \left(K^{1-N/2} \gamma^{1+N/2}\right)\nonumber
+  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
    \,,\label{Vdisc}
@@
 \begin{equation}
-  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,\quad
-  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
+  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
+  %\quad
+%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
 \end{equation}
@@
 \begin{equation}\label{rhodisc}
-\rho_i = K \frac{\ln(r_{i+1}/r_i)}{\Delta t}+\gamma+1\,,\quad
+\rho_i = \frac{T}{\beta} \frac{\ln(r_{i+1}/r_i)}{\Delta t}
++
+\frac{(2+\gamma)\beta}{2\gamma}
+\,,\quad
 \dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
 \end{equation}
-
-Note that we've neglected terms of order $\mathcal O(1/N)$ in the action (\ref{Vdisc}) and that we've assumed $r(t)$ to be at least twice differentiable.
+Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.
 
@@
 
-We choose another approach, the so-called \emph{staging} as described in Tuckerman et al. (1993), and diagonalize the harmonic part only in between the measurement points.
-We first note that the Hamiltonian $\mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})$ (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
+We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
+We first note that Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
 Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
@@
 The discretized harmonic part of the action is rewritten as
-
 \begin{multline}
@@
 \end{multline}
-
 with
-
 \begin{equation}
@@
 \end{equation}
-
-Following Tuckerman et al. (1993) we apply the coordinate transformation
-
+Following \cite{tuckerman1993efficient} we apply the coordinate transformation
 \begin{equation}
@@
 \end{equation}
-
 with its inverse given by
-
 \begin{equation}
@@
 \end{equation}
-
 or equivalently by the recursive relation
-
 \begin{equation}
@@
 \end{equation}
-
 %Moreover, we introduce the following transformation for the parameters,
@@
 %
+
 This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal H_{HMC}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
@@
     -
-    \frac
-    {(y_s-r_{(s-1)j+1}e^{u_{(s-1)j+1}})^2}
-    {\sigma^2}
+    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{2\sigma^2}
    \right\}\,,\\
@@
      \left(
-            \rho_i-\frac{\beta}{\gamma}e^{-\beta \bar q_i}
+        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
      \right)^2
     -
-        T\bar q_i\dot\rho_i
+    \frac{\beta^2}{2\gamma}
+    e^{-\beta q_i}
+   -
+    T q_i\dot\rho_i
    \right\}
   \\
-  &-
-  \frac{(2+\gamma)\beta^2}{4\gamma}
-  +
+  &+
   \frac{1}{\gamma}
@@
 \begin{enumerate}
-  \item[i]
+  \item[\it i.]
   $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$
-  \item[ii]
+  \item[\it ii.]
   $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$
-  \item[iii]
+  \item[\it iii.]
   $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$
@@
 for $\tau =P\Delta \tau$.
-In regime (ii) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
+In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
 However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.

### 2015-03-15 Main_HMC_4_2.jl
in the middle of last parallelization attempt

@@ -25,7 +25,7 @@
 ## simone.ulzega@eawag.ch
 ## carlo.albert@eawag.ch
 ## ============================================================================================
-addprocs(3)
+addprocs(2)
 num_workers  = nworkers()
 first_worker = workers()[1]
 last_worker  = workers()[end]
@@ -212,6 +212,10 @@ du = myDArray(I->[ext_u[i] for i in I[1]],(du_size,),workers(),[nworkers()],chk_
 dr = myDArray(I->[ext_r[i] for i in I[1]],(du_size,),workers(),[nworkers()],chk_indexes)
 dy = myDArray(I->[ext_y[i] for i in I[1]],(dy_size,),workers(),[nworkers()],chk_y_indexes)
 
+du_short = distribute(u)
+dr_short = distribute(r)
+dy_short = distribute(y)
+
 ## --------------------------------------------------------------------------------------------
 ## Share important stuff with all processes
 ## --------------------------------------------------------------------------------------------
@@ -227,13 +231,18 @@ for pid in workers()
     remotecall(pid, x->(global r; r = x; nothing), r)
     remotecall(pid, x->(global dt; dt = x; nothing), dt)
     remotecall(pid, x->(global sigma; sigma = x; nothing), sigma)
-    remotecall(pid, x->(global proc_indexes; proc_indexes = x; nothing), proc_indexes)
+    remotecall(pid, x->(global bdy_indexes; bdy_indexes = x; nothing), bdy_indexes)
     remotecall(pid, x->(global lnr_der; lnr_der = x; nothing), lnr_der)
     remotecall(pid, x->(global num_workers; num_workers = x; nothing), num_workers)
     remotecall(pid, x->(global first_worker; first_worker = x; nothing), first_worker)
     remotecall(pid, x->(global last_worker; last_worker = x; nothing), last_worker)
+    remotecall(pid, x->(global du_indexes; du_indexes = x; nothing), du_indexes)
 end
-
+for pid in workers()
+    remotecall(pid, x->(global du; du = x; nothing), du)
+    remotecall(pid, x->(global dy; dy = x; nothing), dy)
+    remotecall(pid, x->(global dr; dr = x; nothing), dr)
+end
 ## --------------------------------------------------------------------------------------------
 ## Loading...
 ## --------------------------------------------------------------------------------------------

### 2015-03-15 ParInf_Fun_AD_4_2.jl

@@ -40,6 +40,60 @@ end
 ##
 
 ##########################################################################
+## Fast potential as ... many functions
+##########################################################################
+
+@everywhere function V_fast_dat(u, i::Int64)      # i = 0:n
+
+    out = ( y[i+1] - r[i*j+1]*exp(u[i*j+1]) )^2/(2*sigma^2)
+    return out
+
+end
+
+
+@everywhere function V_fast_bdy(theta, u, i::Int64)    # i = 1:n
+
+    K = T*exp(theta[1])
+    g = exp(theta[2])
+    
+    out = K * ( u[(i-1)*j+1] - u[i*j+1] )^2/( j*dt ) / ( 2*g )
+    
+    return out
+end
+
+@everywhere function V_fast_stg(theta, u, i::Int64, k::Int64)  # i = 1:n, k = 2:j
+    
+    K = T*exp(theta[1])
+    g = exp(theta[2])
+
+    out = k*u[(i-1)*j+k]^2/( (k-1)*dt ) * K/(2*g)
+    
+    return out
+end
+
+@everywhere function V_fast(theta, u) 
+    V_fast_sum = V_fast_dat(u, y, r, 0)
+    for i = 1:n
+        V_fast_sum += V_fast_bdy(theta, u, i) + V_fast_dat(u, y, r, i)
+        for k = 2:j 
+            V_fast_sum += V_fast_stg(theta, u, i, k)
+        end
+    end
+    return V_fast_sum
+end
+
+out = dzeros(2)
+for p in workers()
+    @spawnat p localpart(out)[1] = sum([fix(localpart(d),i) for i = 1:3])
+end
+out
+
+@everywhere V_fast_der_dat = rdiff(V_fast_dat,(ones(N), 1            ))
+@everywhere V_fast_der_bdy = rdiff(V_fast_bdy,(ones(2), ones(N), 1   ))
+@everywhere V_fast_der_stg = rdiff(V_fast_stg,(ones(2), ones(N), 1, 1))
+
+
+##########################################################################
 ## Fast potential as an expression (-> array of expressions)
 ##########################################################################
 
@@ -52,23 +106,20 @@ V_fast_der_t_expr = DArray(I->[:($i) for i in I[1]],(nworkers(),),workers(),[nwo
     # firstind = bdy_indexes[1][1];
     # lastind  = bdy_indexes[1][end];
     lastind  = length(bdy_indexes[1]);
[…]

### 2015-03-15 test.jl
simple clean up

@@ -1,0 +1,3143 @@
+
+x=1.3
+typeof(x)
+
+int64('b')
+
+for i=97:110
+	print(string(char(i), "\n"))
+end
+x=(1,2,3)
+typeof(x)
+
+x[1],x[3]
+
+x[4]=44
+
+a,b=x
+
+a=x
+
+a
+
+x,y=a
+
+x
+
+y
+
+
+[1 2 3 4 5; 11 12 13 14 15; 1 2 3 4 5; 7 8 9 1 2; 2 4 6 8 1]$
+
+q=Array(Int64,2)
+
+rand(2,2)
+
+[[1 2], [3 4]]
+
+ Int64[vec([1 2] [3 4])]
+
+ [x*y for x in 1:2, y in 1:4]
+
+[[1 2] [3 4]]
+
+w=vec([1 2 3; 4 5 6; 7 8 9])
+
+[1 2 3; 4 5 6; 7 8 9]
+
+rt=reshape(w,3,3)
+
+sum(rt)
+mean(rt)
+std(rt)
+std(rt)/mean(rt)*100
+
+count(x->x>=8,rt)
+
+rt
+
+a=[x*y for x in 1:2, y in 1, z in 1:3]
+
+a[7]
+
+rt[2,3]
+
+rt[end]
+
+rt[1:2:end]
+
+asa=reshape(1:9,3,3)
+
+asa[1:2:end]
+
+linspace(1,2,5)
+
+a=linspace(0,1)
+
+a[repmat([true, false],50)]
+
+
+x=reshape(1:8,2,4)
+
+x[:,2:3]=repmat([1 3],2)
+
+x
+
+repmat([1,3],2)
+
+x=cell(2)
+
+x[1]=ones(2)
+x[2]=trues(3)
+
+x
+
+a=x
+
+a
+
+b=copy(x)
+c=deepcopy(x)
+
+x[1]="Bang"
+
+x[2][1]=false
+
+x
+a
+
+b
+c
+
+p=Point(0,0.0,"Origin")
+
+typeof(p)
+
+p.x=1
+names(p)
+
+names(Point)
+
+p.x
+
+
+y={"a"=>1,(2,3)=>true}
+
+y["a"]
+y[(2,3)]
+
+haskey(y,"a")
+
+
+keys(y),values(y)
+
+typeof(keys(y))
+
+repr(123.3)
+
+x=123
+asas="$x+3=$(x+3)"
+typeof(asas)
+
+x==123 ? a=23 : a=44
+
+i=0;
[…]

### 2015-06-16 Analysis_HMC_napa.jl
backup trying to fix bug with sin rain

@@ -17,7 +17,7 @@
 ## ============================================================================================
 dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/temp_data"
 dir2  = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data"
-fname = string("_sin301_n10")
+fname = string("_sinr_n10")
 ##
 ##
 ## ============================================================================================
@@ -236,6 +236,7 @@ plt.plot(xy, xy, "r--", linewidth = 3)
 plt.savefig("$dir/figure4$fname.png",transparent=true)
 
 
+#=
 range = (nsample_burnin+2):(nsample+1)
 plt.figure(4)
 plt.subplots_adjust(hspace=0.5)
@@ -261,6 +262,7 @@ plt.plot(res.qy, res.qx, linewidth = 3)
 xy = [x for x in linspace(minimum([res.qx,res.qy]),maximum([res.qx,res.qy]),4)]
 plt.plot(xy, xy, "r--", linewidth = 3)
 plt.savefig("$dir/figure4$fname.png",transparent=true)
+=#
 
 ## --------------------------------------------------------------------------------------------
 ## Plot parameters:

### 2015-06-16 Data_Generator.jl

@@ -20,7 +20,8 @@
 
 dir  = "C:/Users/ulzegasi/Julia_files/ParInf_HMC" 
 dir2 = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data" 
-
+using PyPlot
+pygui(true)
 
 ##
 ## ============================================================================================
@@ -30,27 +31,26 @@ dir2 = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data"
 range = 2:5002
 tdat  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
 
-n_temp = 10               # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
-N_temp = n_temp*1000 + 1  # Total number of points used to generate a system realization
+n = 10          # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
+N = n*1000 + 1  # Total number of points used to generate a system realization
 
-t_temp  = linspace(tdat[1], tdat[end], N_temp)      # Generate N time points in [t[1],t[end]]      
-T_temp  = t_temp[end]-t_temp[1]                     # Total time interval
-dt_temp = T_temp/(N_temp-1)                         # Time step
-ty_temp = iround(linspace(1, N_temp, n_temp + 1))   # Indeces of "boundary" beads (= measurement points)     
+t  = linspace(tdat[1], tdat[end], N) # Generate N time points in [t[1],t[end]]      
+T  = t[end]-t[1]                     # Total time interval
+dt = T/(N-1)                         # Time step
+ty = iround(linspace(1, N, n + 1))   # Indeces of "boundary" beads (= measurement points)     
 
 K     = 50.         # Retention time
 gam   = 0.2         # Dimensionless noise parameter
 sig   = 0.10        # Measurement noise
 
-S = zeros(N_temp)   # System realization (with true parameters)
+S = zeros(N)   # System realization (with true parameters)
 
 ##
 ## ============================================================================================
 ## Rain input
 ## ============================================================================================
 ##
[…]

### 2015-06-16 Main_HMC_napa_sinrain.jl

@@ -54,7 +54,7 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 10000
+const nsample_eff    = 20000
 const nsample        = nsample_eff + nsample_burnin
 
 const dtau           = 0.30                # MD time step
@@ -139,7 +139,7 @@ for s = 1:(n+1)
 end
 =#
 for s = 1:n 								  # ... and their linear interpolation (staging points)
-    q[ty[s]:ty[s+1]] = (1/true_bet)*linspace(bq[s],bq[s+1],ty[s+1]-ty[s]+1)  # +(sigma/bet)*randn(ty[s+1]-ty[s]+1)
+    q[ty[s]:ty[s+1]] = (1/bet)*linspace(bq[s],bq[s+1],ty[s+1]-ty[s]+1)  # +(sigma/bet)*randn(ty[s+1]-ty[s]+1)
 end
 #=
 for s = 1:n+1

### 2015-06-16 ParInf_Fun_AD_napa.jl

@@ -176,10 +176,10 @@ dV_n_u     = rdiff(V_n, outsym=:out_n, u     = ones(Float64, N))
 dV_1_theta = rdiff(V_1, outsym=:out_1, theta = ones(Float64, params)) 
 dV_1_u     = rdiff(V_1, outsym=:out_1, u     = ones(Float64, N)) 
 
-@eval dV_n(theta,u) = [($dV_n_theta)[2][1],($dV_n_u)[2][1]]
-@eval dV_1(theta,u) = [($dV_1_theta)[2][1],($dV_1_u)[2][1]]
+@eval dV_n(theta,u) = [($dV_n_theta)[2],($dV_n_u)[2]]
+@eval dV_1(theta,u) = [($dV_1_theta)[2],($dV_1_u)[2]]
 # or
-@eval dV(theta,u) = [($dV_n_theta)[2][1]+($dV_1_theta)[2][1],($dV_n_u)[2][1]+($dV_1_u)[2][1]]
+@eval dV(theta,u) = [($dV_n_theta)[2]+($dV_1_theta)[2],($dV_n_u)[2]+($dV_1_u)[2]]
 
 ##
 ##

### 2015-06-16 Analysis_HMC_napa.jl
again backup trying to fix bug with sin rain

@@ -17,7 +17,7 @@
 ## ============================================================================================
 dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/temp_data"
 dir2  = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data"
-fname = string("_sinr_n10")
+fname = string("_sinr")
 ##
 ##
 ## ============================================================================================

### 2015-06-16 Data_Generator.jl

@@ -81,6 +81,8 @@ yerr=2*sig*y
 plt.errorbar(t[ty], y, yerr=(yerr,yerr), fmt="o", color="r", markersize = 10, capsize=8, elinewidth=4)
 =#
 
+t[ty]
+ty
 St = zeros(N,2)
 St[:,1] = S
 St[:,2] = t

### 2015-06-16 Main_HMC_napa_sinrain.jl

@@ -18,7 +18,7 @@
 dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC"  # Main project directory, local repository   
 dir2  = "$dir/temp_data"                            # Secondary directory (e.g., output storage)
 dir3  = "$dir/input_data"                           # Input data directory
-fname = string("_sinr_n10")                        # Output files
+fname = string("_n10_K50_g02_s10_sinr")                        # Output files
 using ReverseDiffSource, PyPlot #, ForwardDiff      # Automated differentiation package
 pygui(true)
 srand(18072011)                                     # Seeding the RNG provides reproducibility
@@ -31,14 +31,16 @@ srand(18072011)                                     # Seeding the RNG provides r
 St     = readdlm("$dir3/St$fname.dat")
 S      = St[:,1]
 long_t = St[:,2]   
-long_r = vec(readdlm("$dir3/r$fname.dat"))
+bq     = vec(readdlm("$dir3/bq$fname.dat"))
 y      = vec(readdlm("$dir3/y$fname.dat"))
-bq     = vec(readdlm("$dir3/bq$fname.dat"))  
 ##
 ## ============================================================================================
 ## System parameters
 ## ============================================================================================
 ##
+range = 2:5002
+tdat  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
+
 const params = 2              # Number of system parameters (k, gamma)
 const n = 10                  # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
 const j = 30                  # n(j-1) -> total number of staging beads, j-1 -> staging beads per segment
@@ -48,7 +50,7 @@ if  ((N-1)%j) != 0
     error("Be careful, the number of staging points j must fulfill (N-1)/j = integer !")
 end
 
-const t  = long_t[iround(linspace(1, size(long_t,1), N))]
+const t  = linspace(tdat[1], tdat[end], N) # Generate N time points in [t[1],t[end]]
 const T  = t[end]-t[1]                     # Total time interval
 const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
@@ -120,13 +122,12 @@ for i = 2:N
     end
 end
[…]

### 2015-06-17 TeXFiles/stochInf1D.tex
update of tex files and few other things. Issue with sin rain not fixed yet

@@
 %\end{equation}
-\begin{equation}
+\begin{equation}\label{action1}
 {\mathcal S}[{q},\dot q]
@@
 \end{equation}
-and rewrite action (\ref{action}) as,
+and rewrite action (\ref{action1}) as,
 \begin{multline}\label{action}
@@
    =
-   \int_0^T \frac{ p^2(t)}{2m_q}dt
+   \int_0^T \frac{ p^2(t)}{2m(t)}dt
    + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
@@
     T q_i\dot\rho_i
-   \right\}
+   \right\}\label{Vdisc}
   \\
@@
   \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
-   \,,\label{Vdisc}
+  \nonumber
+   \,,
 \end{align}
@@
 \begin{equation}\label{rhodisc}
-\rho_i = \frac{T}{\beta} \frac{\ln(r_{i+1}/r_i)}{\Delta t}
+\rho_i = \frac{T}{\beta} \frac{\ln(r_{i}/r_{i-1})}{\Delta t}
 +
@@
 We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
-We first note that Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
+We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
 Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
@@
 \begin{equation}
-  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{s+k+1}+ \frac{1}{k}q_{sj+1} \,.
+  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
 \end{equation}
@@
     +
-    \frac{T}{j\Delta t}
+    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{\sigma^2}
+   \right\}
+   \\
+  \nonumber
+  &+
+  \frac{T}{2j\Delta t}
+  \sum_{s=1}^{n}
     (u_{(s-1)j+1} - u_{sj+1})^2
-    -
-    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{2\sigma^2}
-   \right\}\,,\\
+   \,,\\
   \mathcal H_1
@@
   \\
+  \nonumber
   &+
@@
 \end{align}
-for $k=2,\dots,j$.
+for $s=1,\dots,n$ and $k=2,\dots,j$.
 

### 2015-06-17 Data_Generator.jl

@@ -20,7 +20,7 @@
 
 dir  = "C:/Users/ulzegasi/Julia_files/ParInf_HMC" 
 dir2 = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data" 
-using PyPlot
+using PyPlot, Distributions
 pygui(true)
 
 ##
@@ -39,9 +39,10 @@ T  = t[end]-t[1]                     # Total time interval
 dt = T/(N-1)                         # Time step
 ty = iround(linspace(1, N, n + 1))   # Indeces of "boundary" beads (= measurement points)     
 
-K     = 50.         # Retention time
-gam   = 0.2         # Dimensionless noise parameter
-sig   = 0.10        # Measurement noise
+K   = 50.         # Retention time
+gam = 0.2         # Dimensionless noise parameter
+bet = sqrt(T*gam/K)
+sig = 0.10        # Measurement noise
 
 S = zeros(N)   # System realization (with true parameters)
 
@@ -67,8 +68,14 @@ for i = 2:N
         error("Negative volume!")
     end
 end
-bq = log((1/K)*(S[ty]./r[ty])) + sig*randn(n+1)  
-y  = r[ty].*exp(bq)                                   
+
+y  = S[ty]/K.*rand(LogNormal(0,sig),n+1)
+bq = log(y./r[ty])
+
+########################
+# bq = log((1/K)*(S[ty]./r[ty])) + sig*randn(n+1)  
+# y  = r[ty].*exp(bq)
+########################                                   
 
 #=
 plt.figure()
@@ -81,8 +88,6 @@ yerr=2*sig*y
 plt.errorbar(t[ty], y, yerr=(yerr,yerr), fmt="o", color="r", markersize = 10, capsize=8, elinewidth=4)
 =#
 
-t[ty]
-ty
 St = zeros(N,2)
 St[:,1] = S
 St[:,2] = t

### 2015-06-17 Main_HMC_napa_sinrain.jl

@@ -19,7 +19,7 @@ dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC"  # Main project directory, lo
 dir2  = "$dir/temp_data"                            # Secondary directory (e.g., output storage)
 dir3  = "$dir/input_data"                           # Input data directory
 fname = string("_n10_K50_g02_s10_sinr")                        # Output files
-using ReverseDiffSource, PyPlot #, ForwardDiff      # Automated differentiation package
+using ReverseDiffSource, PyPlot, Distributions #, ForwardDiff      # Automated differentiation package
 pygui(true)
 srand(18072011)                                     # Seeding the RNG provides reproducibility
 # srand(time())                                     # Seeding the RNG with actual time provides randomness
@@ -31,7 +31,7 @@ srand(18072011)                                     # Seeding the RNG provides r
 St     = readdlm("$dir3/St$fname.dat")
 S      = St[:,1]
 long_t = St[:,2]   
-bq     = vec(readdlm("$dir3/bq$fname.dat"))
+# bq     = vec(readdlm("$dir3/bq$fname.dat"))
 y      = vec(readdlm("$dir3/y$fname.dat"))
 ##
 ## ============================================================================================
@@ -41,7 +41,7 @@ y      = vec(readdlm("$dir3/y$fname.dat"))
 range = 2:5002
 tdat  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
 
-const params = 2              # Number of system parameters (k, gamma)
+const nparams = 2              # Number of system parameters (k, gamma)
 const n = 10                  # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
 const j = 30                  # n(j-1) -> total number of staging beads, j-1 -> staging beads per segment
 const N = int64(n*j+1)        # Total number of discrete time points = n*j + 1
@@ -96,14 +96,14 @@ lnr_der = Array(Float64,N)      # Log derivative of (smoothed) rain input
 q       = Array(Float64,N)       
 u       = Array(Float64,N)       
 ## Masses and momenta for HMC
-mp      = Array(Float64,params+N)     
-p       = Array(Float64,params+N)
[…]

### 2015-06-17 ParInf_Fun_AD_napa.jl

@@ -166,14 +166,14 @@ end
 ##########################################################################
 ## Derivatives
 ##########################################################################
-# dV_N_theta = rdiff(V_N, outsym=:out_N, theta = ones(Float64, params)) 
+# dV_N_theta = rdiff(V_N, outsym=:out_N, theta = ones(Float64, nparams)) 
 # dV_N_u     = rdiff(V_N, outsym=:out_N, u     = ones(Float64, N)) 
 # @eval dV_N(theta,u) = [($dV_N_theta)[2][1],($dV_N_u)[2][1]]
 
-dV_n_theta = rdiff(V_n, outsym=:out_n, theta = ones(Float64, params)) 
+dV_n_theta = rdiff(V_n, outsym=:out_n, theta = ones(Float64, nparams)) 
 dV_n_u     = rdiff(V_n, outsym=:out_n, u     = ones(Float64, N)) 
 
-dV_1_theta = rdiff(V_1, outsym=:out_1, theta = ones(Float64, params)) 
+dV_1_theta = rdiff(V_1, outsym=:out_1, theta = ones(Float64, nparams)) 
 dV_1_u     = rdiff(V_1, outsym=:out_1, u     = ones(Float64, N)) 
 
 @eval dV_n(theta,u) = [($dV_n_theta)[2],($dV_n_u)[2]]
@@ -201,8 +201,8 @@ function napa(theta, u, counter)     # (Tuckerman et al., JCP 97(3), 1992 )
     for s = 1:n 
         for k = 2:j 
             u_old = u[(s-1)*j+k]
-            u[(s-1)*j+k] = u[(s-1)*j+k]*cos(w_stg(k)*dtau/2.0) + p[params+(s-1)*j+k]*sin(w_stg(k)*dtau/2.0) / (m_stg*w_stg(k))
-            p[params+(s-1)*j+k] = p[params+(s-1)*j+k]*cos(w_stg(k)*dtau/2.0) - m_stg*w_stg(k)*u_old*sin(w_stg(k)*dtau/2.0)
+            u[(s-1)*j+k] = u[(s-1)*j+k]*cos(w_stg(k)*dtau/2.0) + p[nparams+(s-1)*j+k]*sin(w_stg(k)*dtau/2.0) / (m_stg*w_stg(k))
[…]

### 2015-06-18 Analysis_HMC_napa.jl
working version of napa code with sin rain

@@ -17,7 +17,7 @@
 ## ============================================================================================
 dir   = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/temp_data"
 dir2  = "C:/Users/ulzegasi/Julia_files/ParInf_HMC/input_data"
-fname = string("_sinr")
+fname = string("_sinr_680_150")
 ##
 ##
 ## ============================================================================================
@@ -30,7 +30,7 @@ paramslist     = readdlm("$dir/params$fname.dat")
 N              = int64(paramslist[findin(paramslist[:,1],["N"]),2])[1]
 j              = int64(paramslist[findin(paramslist[:,1],["j"]),2])[1]
 n              = int64(paramslist[findin(paramslist[:,1],["n"]),2])[1] 
-nparams		   = int64(paramslist[findin(paramslist[:,1],["params"]),2])[1] 
+nparams		   = int64(paramslist[findin(paramslist[:,1],["nparams"]),2])[1] 
 nsample_burnin = int64(paramslist[findin(paramslist[:,1],["nsample_burnin"]),2])[1]
 nsample_eff    = int64(paramslist[findin(paramslist[:,1],["nsample_eff"]),2])[1]  
 n_napa         = int64(paramslist[findin(paramslist[:,1],["n_napa"]),2])[1] 
@@ -103,12 +103,12 @@ pygui(true)
 
 plt.figure(1,figsize=(12.5, 8.5))
 plt.subplots_adjust(hspace=0.5)
-plt.subplot(211)[:set_xlim]([-5,605])
+plt.subplot(211)[:set_xlim]([-5,845])
 plt.grid("on")
 plt.ylabel("S[t]")
 plt.title("Water volume")
 plt.plot(t, S,"r")
-plt.subplot(212)[:set_xlim]([-5,605])
+plt.subplot(212)[:set_xlim]([-5,845])
 plt.grid("on")
 plt.xlabel("time")
 plt.ylabel("r[t]")
@@ -118,7 +118,7 @@ plt.savefig("$dir/figure1$fname.png",transparent=true)
 
 plt.figure(2,figsize=(12.5, 8.5))
 plt.subplots_adjust(hspace=0.5)
-plt.subplot(211)[:set_xlim]([-5,605])
+plt.subplot(211)[:set_xlim]([-5,845])
[…]

### 2015-06-18 Main_HMC_napa_sinrain.jl

@@ -31,7 +31,6 @@ srand(18072011)                                     # Seeding the RNG provides r
 St     = readdlm("$dir3/St$fname.dat")
 S      = St[:,1]
 long_t = St[:,2]   
-# bq     = vec(readdlm("$dir3/bq$fname.dat"))
 y      = vec(readdlm("$dir3/y$fname.dat"))
 ##
 ## ============================================================================================
@@ -56,12 +55,11 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 20000
+const nsample_eff    = 50000
 const nsample        = nsample_eff + nsample_burnin
 
-const dtau           = 0.30                # MD time step
+const dtau           = 0.25                # MD time step
 const n_napa         = 3                   # Number of NAPA time steps
-# nn = 16  ; deltatau = dtau / nn          # Short time steps in RESPA --> NOT NEEDED IN THE NAPA IMPLEMENTATION
 
 const true_K     = 50.                    # Retention time
 const true_gam   = 0.2                     # Dimensionless noise parameter
@@ -131,25 +129,14 @@ r = Array(Float64,N); r = sin(t/100.).*sin(t/100.) + 0.1
 for i = 1:(N-1)  
 	lnr_der[i] = (log(r[i+1])-log(r[i]))/dt   # Log-derivative of the rain
 end
-#=
-for s = 1:(n+1)
-	y[s] = yfull[ty[s]]						  # Synthetic measurement data
-end           		
-for s = 1:(n+1) 
-    q[ty[s]] = (1/bet)*log(y[s]/r[ty[s]])	  # To dimensionless coordinates (measurement points) ...
-end
-=#
+
 q_init = (1/true_bet)*log(y./r[ty])
 bq     = true_bet*q_init
 for s = 1:n 								  # ... and their linear interpolation (staging points)
     q[ty[s]:ty[s+1]] = linspace(q_init[s],q_init[s+1],ty[s+1]-ty[s]+1)
 end
-#=
-for s = 1:n+1
-	bq[s] = log(y[s]/r[ty[s]])
-end
-=#
 
+############## It could be interesting to plot a few figures to visualize input data ##############
 #=
[…]

### 2015-06-19 Analysis_HMC_napa.jl
working version of napa code with sin rain + good results

@@ -303,9 +303,9 @@ plt.savefig("$dir/figure6$fname.png",transparent=true,dpi=300)
 
 lowlim = maximum([2,nsample_burnin+1])                  # To possibly discard initial 
                                                         # system realizations (burn-in)
-plt.figure(7)
+plt.figure(7,figsize=(12.5, 8.5))
 plt.subplots_adjust(hspace=0.5)
-plt.subplot(211)
+plt.subplot(211)[:set_xlim]([0.0,250])
 plt.xlabel("K")
 plt.ylabel("Probability density")
 plt.title("PDF")
@@ -319,8 +319,7 @@ xx_K   = linspace(0.0001, 400, 500)
 plt.plot(xx_K, pdf(LogNormal(4,1),xx_K), "b", linewidth = 1, label = "LogNormal distribution")
 # plt.axvline(x=mu_K, linewidth=2, color="r")
 # plt.axvspan(mu_K-sig_K, mu_K+sig_K, facecolor="y", alpha=0.3)
-
-plt.subplot(212)
+plt.subplot(212)[:set_xlim]([0.0,1.6])
 plt.xlabel("Gamma")
 plt.ylabel("Probability density")
 plt.title("PDF")

### 2015-07-16 TeXFiles/Manuscript_Albert_Ulzega.tex
tex files updated to NewJPhys template

--- 2015-06-17 TeXFiles/stochInf1D.tex
@@
-\documentclass[11pt, a4paper]{article}
-
-\usepackage{a4wide}
-\usepackage{amsmath,amssymb,amsthm,amscd}
-\usepackage{natbib}
-\usepackage[pdftex]{graphicx}
-\usepackage[affil-it]{authblk}
-\usepackage{sidecap}
-\usepackage[font=small,labelfont=bf]{caption}
-
-%\usepackage{showkeys}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%    INSTITUTE OF PHYSICS PUBLISHING                                   %
+%                                                                      %
+%   `Preparing an article for publication in an Institute of Physics   %
+%    Publishing journal using LaTeX'                                   %
+%                                                                      %
+%    LaTeX source code `ioplau2e.tex' used to generate `author         %
+%    guidelines', the documentation explaining and demonstrating use   %
+%    of the Institute of Physics Publishing LaTeX preprint files       %
+%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
+%                                                                      %
+%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
+%                                                                      %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%
+% First we have a character check
+%
+% ! exclamation mark    " double quote
+% # hash                ` opening quote (grave)
+% & ampersand           ' closing quote (acute)
+% $ dollar              % percent
+% ( open parenthesis    ) close paren.
+% - hyphen              = equals sign
+% | vertical bar        ~ tilde
+% @ at sign             _ underscore
+% { open curly brace    } close curly
+% [ open square         ] close square bracket
+% + plus sign           ; semi-colon
+% * asterisk            : colon
+% < open angle bracket  > close angle
+% , comma               . full stop
+% ? question mark       / forward slash
+% \ backslash           ^ circumflex
+%
+% ABCDEFGHIJKLMNOPQRSTUVWXYZ
+% abcdefghijklmnopqrstuvwxyz
+% 1234567890
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+\documentclass[12pt,a4paper,final]{iopart}
+\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
+%Uncomment next line if AMS fonts required
+\usepackage{iopams}
+\usepackage{graphicx}
+\usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
+%\usepackage{amsmath,amssymb,amsthm,amscd}
+%\usepackage{natbib}
 
 \usepackage[all]{xy}
-
-\newcommand{\bb}[1]{\begin{equation}\label{#1}}
-\newcommand{\ee}{\end{equation}}
-\newcommand{\vc}[1]{{\bf #1 }}
-\renewcommand{\vec}[1]{{\bf #1}}
-\newcommand{\Sum}[1]{\sum\limits_{#1}}
-\newcommand{\Int}[1]{\int\limits_{#1}}
-\newcommand{\Prod}[1]{\prod\limits_{#1}}
-\newcommand{\const}{\operatorname{const}}
-\newcommand{\ul}{\underline}
-\newcommand{\ol}{\overline}
-\newcommand{\tr}{\operatorname{tr}}
-\newcommand{\bbA}{\mathbb A}
-\newcommand{\bbR}{\mathbb R}
-\newcommand{\bbN}{\mathbb N}
-\newcommand{\bbZ}{\mathbb Z}
-\newcommand{\bbC}{\mathbb C}
-\newcommand{\bbT}{\mathbb T}
-\newcommand{\QB}{\mathcal H}
-\newcommand{\PQB}{\tilde{\mathcal H}}
-\newcommand{\CQB}{\mathcal H'}
-\newcommand{\Ad}{\operatorname{Ad}}
-\newcommand{\ad}{\operatorname{ad}}
-\newcommand{\gh}{\operatorname{gh}}
-\newcommand{\Der}{\operatorname{Der}}
-\newcommand{\Aut}{\operatorname{Aut}}
-\newcommand{\aut}{\operatorname{aut}}
-\newcommand{\Diff}{\operatorname{Diff}}
-\renewcommand{\dim}{\operatorname{dim}}
-\newcommand{\Vol}{\operatorname{Vol}}
-\newcommand{\Pf}{\operatorname{Pf}}
-\newcommand{\Cov}{\operatorname{Cov}}
-
-\newcommand{\g}{{\bf g}}
-\newcommand{\h}{{\bf h}}
-\renewcommand{\d}{{\bf d}}
-\newcommand{\e}{{\bf e}}
-\renewcommand{\t}{{\bf t}}
-\newcommand{\lpartial}{\overset{\rightarrow}{\partial}}
-\newcommand{\rpartial}{\overset{\leftarrow}{\partial}}
-\newcommand{\lder}[1]{\overset{\rightarrow}{\vc #1}}
-\newcommand{\rder}[1]{\overset{\leftarrow}{\vc #1}}
-\newcommand{\supp}{\operatorname{supp}}
-\newcommand{\im}{\operatorname{Im}}
-\renewcommand{\ker}{\operatorname{Ker}}
-\newcommand{\Var}{\operatorname{Var}}
-
-%define the \identy operator
-\def\litID{{\sf id}}
-\def\identy{{\mathsurround0pt\mathchoice{\textidenty}{\textidenty}{\scptidenty}{\scptidenty}}}
-\def\scptidenty{\setbox0\hbox{$\scriptstyle1$}\bothidenty}
-\def\textidenty{\setbox0\hbox{$1$}\bothidenty}
-\def\bothidenty{\rlap{\hbox to.97\wd0{\hss\vrule height.06\ht0 width.82\wd0}}
- \copy0\rlap{\kern-.36\wd0\vrule height1.05\ht0 width.05\ht0}\kern.14\wd0}
-%end of definition \identy
-
+\newcommand{\eq}{\mbox{eq}}
+\newcommand{\dt}{\mathit{dt}}
 \newcommand{\bbM}{\mathbb M}
@@
 \newcommand{\Bt}{\pmb\Theta}
-\newcommand{\dt}{\mathit{dt}}
-\newcommand{\eq}{\text{eq}}
+\newcommand{\vc}[1]{{\bf #1 }}
 
 \begin{document}
-\title{Parameter Inference for a Simple Stochastic Hydrological Model}
 
-\author{C. Albert\thanks{\noindent Email: carlo.albert@eawag.ch} \,,
-S. Ulzega}
-\affil{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
-\maketitle
+\title[Parameter inference with stochastic models]{Parameter Inference for a Simple Stochastic Hydrological Model}
+
+\author[cor1]{Carlo Albert}%$^{1,2}$
+\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
+%\address{$^2$Address Two, Neverland}
+\ead{carlo.albert@eawag.ch}
+
+\author{Simone Ulzega}
+\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
+\ead{simone.ulzega@eawag.ch}
+
+%\author[cor1]{Author Three}
+%\address{Address Four, Neverland}
+%\eads{\mailto{author.three@mail.com}, \mailto{author.three@gmail.com}}
+
+\begin{abstract}
+\textbf{[TO BE WRITTEN]} This document describes the  preparation of an article using \LaTeXe\ and \verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
+This class file is designed to help
+authors produce preprints in a form suitable for submission to any of the
+journals published by IOP Publishing.
+Authors submitting to any IOP journal, i.e.\
+both single- and double-column ones, should follow the guidelines set out here.
+On acceptance, their TeX code will be converted to
+the appropriate format for the journal concerned.
+
+\end{abstract}
+
+%Uncomment for PACS numbers title message
+\pacs{00.00, 20.00, 42.10}
+% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
+\vspace{2pc}
+\noindent{\it Keywords}: Article preparation, IOP journals
+% Uncomment for Submitted to journal title message
+\submitto{\NJP}
+% Comment out if separate title page not required
+%\maketitle
+
+\section{Introduction}
 
@@
 \end{equation}
-Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
+Due to the state-dependence of the noise term, Eq.~(\ref{sde}) is ill-defined unless we specify the discretization convention.
 For computational reasons, we work with the {\em Stratonovich} convention.
-Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{eq}=Kr_0$ (see below).
+Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see below).
 
@@
 \end{equation}
-W.r.t. these new  variables and parameters, model equation (\ref{sde}) becomes
+W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
 \begin{equation}
@@
 
-%Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.
-
-According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
+According to Ref.~\cite{lau_2007}, the probability $P(q_1,T|q_0,0)$ of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
 \begin{equation}\label{pathint}
@@
 The {\em action} is a functional on the space of paths and reads as
-%\begin{equation}\label{action}
-%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
-%  =
-%  \int_0^T
-%  dt
-%  \left\{
-%    \frac{K}{2\gamma S^2(t)}
-%    \left(
-%        \dot S(t)
-%        -
-%        r(t)
-%        +
-%        \frac{1+\gamma}{K}S(t)
-%    \right)^2
-%    -
-%    \frac{2+\gamma}{4K}
-%  \right\}
-% \,.
-%\end{equation}
 \begin{equation}\label{action1}
@@
 \end{equation}
-and rewrite action (\ref{action1}) as,
-\begin{multline}\label{action}
-{\mathcal S}[{q},\dot q]
-=
-\frac{1}{T}
+and rewrite the action~(\ref{action1}) as,
+\begin{eqnarray}\label{action}
+{\mathcal S}[{q},\dot q] \nonumber
+\\
+= \frac{1}{T}
 \int_0^T dt\left\{
@@
      %\frac{(2+\gamma)\beta^2}{4\gamma}
-\right\}
+\right\} \nonumber
 \\
-+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
++ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
 \\
@@
 %    \frac{(2+\gamma)\beta^2}{4\gamma}
-\right\}
+\right\}  \nonumber
 \\
@@
 \,.
-\end{multline}
+\end{eqnarray}
 
-Next, we calculate the equilibrium distribution $P_{\text{eq}}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$.
+Next, we calculate the equilibrium distribution $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$.
 Systems at thermal equilibrium must fulfill detailed balance, i.e.,
 \begin{equation}\label{detailed_balance}
-P(q_1 t_1 | q_0 t_0 ) P_{\text{eq}}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\text{eq}}(q_1) \,.
+P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,.
 \end{equation}
@@
 \begin{equation}\label{Peq}
-  P_{\text{eq}}(q)
+  P_{\eq}(q)
   \propto
@@
 \end{equation}
-Transforming back to the original variables, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
+Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
 \begin{equation}\label{inverse_gamma}
-  P_{\text{eq}}(S)
+  P_{\eq}(S)
   \propto
@@
 
-%The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
-%\begin{equation}\label{likelihood}
-%  f(\vc y | \bt)
-%  \propto
-%  \int
-%  \exp\bigg[
-%    -{\tilde{\mathcal S}}[\dot{S},S]
-%    -\frac{1}{2}
-%    \sum_{i=1}^{n+1}
-%    \frac
-%    {(y_i-S(t_i)/K)^2}
-%    {\sigma^2}
-%  \bigg]
-%  {\mathcal D} S(\bt)\,,
-%\end{equation}
-%with the parameter vector $\bt=(K,\gamma)^T$.
-%Choosing for the parameters $K$ and $\gamma$ the improper prior given by
-%
-%\begin{equation}\label{prior}
-%  f(\bt)
-%  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
-%\end{equation}
-%
-%{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
-%\begin{equation}\label{posterior}
-%  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
-%\end{equation}
-
 The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
@@
 of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
-Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
+Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by Ref.~\cite{duane_1987}.
 This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system.
@@
 \begin{equation}\label{Hamiltonian}
-    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
+    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
     =
@@
   \int
-  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
+  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
   {\mathcal Dp}
@@
 
-The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps
+The HMC method, which is a combination of the {\em Metropolis algorithm} of Ref.~\cite{metropolis_1953} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps,
 \begin{enumerate}
@@
 and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
-\begin{align}
-   K( p,{\pmb\pi})
-   &\approx
+\begin{eqnarray}
+   K( p,{\pmb\pi}) \approx
    \sum_{i=1}^N
@@
    \\
-   V(q,\bt)
-   &\approx
-   \frac{\Delta t}{T}
-   \sum_{i=2}^{N}
-   \left\{
-    \frac{1}{2}
-    T^2 \dot q_i^2
+   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
+   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
+     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
+    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
+  \\
   +
-    \frac{1}{2}
-     \left(
-        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
-     \right)^2
-    -
-    \frac{\beta^2}{2\gamma}
-    e^{-\beta q_i}
-   -
-    T q_i\dot\rho_i
-   \right\}\label{Vdisc}
-  \\
-  &+
   \frac{1}{\gamma}
@@
   \\
-  & +
+  +
   \sum_{s=1}^{n+1}
   \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
-  \nonumber
+  \label{Vdisc}
    \,,
-\end{align}
+\end{eqnarray}
 with
@@
 
-%\begin{equation}
-%  \dot{q}_i:=\frac{ q_i -  q_{i-1}}{\Delta t}\,,\quad
-%  \bar{ q}_i:=\frac{q_i + q_{i-1}}{2}\,,
-%\end{equation}
-%and the analogous definition of $\dot r_i$ and $\bar r_i$.
-
 For strongly varying rain input or high-frequency output data, $\Delta t$ needs to be chosen small.
@@
 Calculating the potential requires then an inversion of a Toeplitz matrix.
-This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes (see \cite{doll_1985_fourier}).
+This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes as outlined in Ref.~\cite{doll_1985}.
 
-We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
+We choose however another approach, the so-called \emph{staging} as described in Ref.~\cite{tuckerman_1992}, and diagonalize the harmonic part only in between the measurement points.
 We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
@@
 The discretized harmonic part of the action is rewritten as
-\begin{multline}
+\begin{eqnarray}
   \sum_{i=2}^{N}
   \frac{T}{2\Delta t}
-  (q_i-q_{i-1})^2
+  (q_i-q_{i-1})^2 \nonumber
   \\ =
@@
     (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
-  \right\}\,,
-\end{multline}
+  \right\}\,,\nonumber
+  \\
+\end{eqnarray}
 with
@@
 \end{equation}
-Following \cite{tuckerman1993efficient} we apply the coordinate transformation
+Following Ref.~\cite{tuckerman_1993} we apply the coordinate transformation
 \begin{equation}
@@
 \end{equation}
-%Moreover, we introduce the following transformation for the parameters,
-%
-%\begin{equation}\label{params_transform}
-%\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
-%\end{equation}
-%
-%and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
-%
 
-This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal H_{HMC}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
+This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal{H}_{\mbox{\tiny HMC}}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
 We write
 \begin{equation}
-  \mathcal H_{HMC}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
+  \mathcal{H}_{\mbox{\tiny HMC}}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
 \end{equation}
 where
-\begin{align}
-  \mathcal H_N
-  &=
+\begin{eqnarray}
+  \mathcal H_N =
   \frac{1}{2}
@@
   \\
-  \mathcal H_n
-  &=
+  \mathcal H_n =
   \frac{1}{2}
@@
     \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{\sigma^2}
-   \right\}
+   \right\} \nonumber
    \\
-  \nonumber
-  &+
+  +
   \frac{T}{2j\Delta t}
@@
     (u_{(s-1)j+1} - u_{sj+1})^2
-   \,,\\
-  \mathcal H_1
-  &=
+   \,, \label{H_n} \\
+  \mathcal H_1=
    \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}
@@
     T q_i\dot\rho_i
-   \right\}
+   \right\} \nonumber
   \\
-  \nonumber
-  &+
+  +
   \frac{1}{\gamma}
@@
   -
-  q_1 \rho_{1}
-\,.
-\end{align}
-The harmonic part $\mathcal H_N$, for the staging beads, scales linearly with $N$.
-The harmonic term $\mathcal H_n$, for the boundary beads, as well as the measurement term scale linearly with $n$.
-Finally, the potential does not scale with neither $n$ nor $N$.
+  q_1 \rho_{1} \,. \label{H_1}
+\end{eqnarray}
+The harmonic part for the staging beads given in Eq.~(\ref{H_N}) scales linearly with $N$.
+The term of Eq.~(\ref{H_n}), including both the harmonic part for the boundary beads and the measurement term, scales linearly with $n$.
+Finally, Eq.~(\ref{H_1}) does not scale with neither $n$ nor $N$.
 Thus, depending on the values of $N$ and $n$, the dynamics happens on very different time scales.
-Therefore, we use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
+Therefore, we use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
 In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes.
@@
 \begin{equation}
-  \mathcal H_{HMC}=\mathcal H_N + \mathcal H'\,.
+  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
 \end{equation}
@@
 \end{equation}
-The fast outer propagator is then given by eqs.
-\begin{align}
-  u_{(s-1)j+k}(\Delta\tau/2)
-  &=
-  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
+The fast outer propagator is then given by the equations,
+\begin{eqnarray}
+  u_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
+  = u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
   +
   \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\\
-  p_{(s-1)j+k}(\Delta\tau/2)
-  &=
-  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
+  p_{(s-1)j+k}(\Delta\tau/2) \nonumber \\
+  = p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
   -
   m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
-\end{align}
+\end{eqnarray}
 for $s=1,\dots,n$ and $k=2,\dots,j$.
@@
 For the boundary beads, it reads
-\begin{align}
-  u_{(s-1)j+1}(\Delta\tau)
-  &=
-  u_{(s-1)j+1}(0)
+\begin{eqnarray}
+  u_{(s-1)j+1}(\Delta\tau) \nonumber \\
+  = u_{(s-1)j+1}(0)
   +
@@
   F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
-  p_{(s-1)j+1}(\Delta\tau)
-  &=
-  p_{(s-1)j+1}(0)
+  p_{(s-1)j+1}(\Delta\tau) \nonumber \\
+  = p_{(s-1)j+1}(0)
   +
@@
   )\,,
-\end{align}
+\end{eqnarray}
 where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
@@
 Thus,
-\begin{equation}
-  p_{(s-1)j+k}(\Delta\tau)
+\begin{eqnarray}
+  p_{(s-1)j+k}(\Delta\tau) \nonumber \\
   =
@@
   \,.
-\end{equation}
+\end{eqnarray}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results and discussions}
 
-%We can now fully exploit the two different time-scales that are involved in the problem.
-First, the harmonic part of the fast component ${\mathcal V}_f[\dot{\vc u},\vc u]$ can be seen as describing the dynamics of heavy coupled \emph{boundary} data-point beads and lighter uncoupled \emph{staging} beads.
-The different weights of boundary and staging beads are defined by adjusting their \emph{effective masses}, $m_{\text{bdy}}$ and $m_{\text{st}}$, respectively.
-These values are used in (\ref{Kdisc}) for the masses $m_{q}$.
-%Second, the two different ranges of the forces associated with ${\mathcal V}_f(\vc u,\tilde{\bt})$ and ${\mathcal V}_s(\vc q,\tilde{\bt})$ can be efficiently taken into account by applying a reversible reference system propagator algorithm (RESPA), as outlined in Tuckerman et al.
-(1992).
-The RESPA algorithm, based on appropriate reversible molecular dynamics integrators, can significantly accelerate simulations of systems that are simultaneously subject to short- and long-range forces by separating the two different time-scales.
-
-%The basic idea of the RESPA algorithm is to apply a Trotter decomposition to the system Liouville propagator $\mathcal{L}$ to generate a discrete time propagator factorized into fast and slow contributions.
-Indeed, as shown in Tuckerman et al.
-(1992), the propagator $\exp[i \mathcal{L} \Delta \tau ]$ can be written as
-
-%\begin{equation}\label{G}
-%  \mathcal{G}(\Delta \tau)=\exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\} \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\} \exp\left\{ \frac{\Delta \tau}{2} F_s(x) \frac{\partial}{\partial p} \right\}\,,
-%\end{equation}
-%
-%where for the sake of clarity we have assumed a one-dimensional system.
-The argument can be easily extended to our multi-dimensional case.
-In (\ref{G}) $F_s$ is the force associated with the slowly varying long-range potential ${\mathcal V}_s$ and $\mathcal{L}_f = \dot{x}(\partial/\partial x) + F_f(x)(\partial/\partial p)$ is the Liouville operator of the fast short-range potential ${\mathcal V}_f$.
-The propagator of the slow force $F_s$ can be well approximated with a large time step $\Delta \tau$.
-On the other hand, in order to ensure stable dynamics, the short-range middle propagator is further factorized into $n_f$ smaller time steps of duration $\delta \tau = \Delta \tau / n_f$, according to the Trotter expansion
-%
-%\begin{equation}\label{G_fast}
-%  \exp\left\{ i \mathcal{L}_f \Delta \tau  \right\}
-%  =\left[
-%  \exp\left\{ \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
-%  \exp\left\{  \delta\tau \dot{x} \frac{\partial}{\partial x}  \right\}
-%  \exp\left\{  \frac{\delta \tau}{2} F_f(x) \frac{\partial}{\partial p} \right\}
-%  \right]^{n_f} \,.
-%\end{equation}
-%
-%Coordinates and conjugate momenta, i.e., $(q,\bt; p,{\pmb\pi})$ in our case, are propagated using (\ref{G}) and (\ref{G_fast}) iteratively by applying a velocity Verlet algorithm.
-%
-\textbf{Some preliminary results}.
 The results shown in the figures refer to synthetic data obtained assuming a simple sinusoidal model for the rain input.
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
-    \caption{Rain input (dashed), system realization (solid line) and synthetic experimental data. The system output and the observed data were generated using $K=200$ and $\gamma = 0.2$.}
-    \label{fig:rain_data_S}
+    \includegraphics[width=0.4\textwidth]{Figs/FigRainData.png}
+    \caption{\label{fig:rain_data_S}Rain input (dashed), system realization (solid line) and synthetic experimental data.
+The system output and the observed data were generated using $K=200$ and $\gamma = 0.2$.}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.9\textwidth]{Figs/FigSpaghetti.png}
+    \includegraphics[width=0.5\textwidth]{Figs/FigSpaghetti.png}
     \caption{Simulated system realizations and synthetic experimental data obtained assuming a simple sinusoidal model for the rain input.
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.6\textwidth]{Figs/FigFinalState.png}
+    \includegraphics[width=0.3\textwidth]{Figs/FigFinalState.png}
     \caption{PDF for the system final state expressed via the coordinate set $\left\{ q_N \right\}$.
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.7\textwidth]{Figs/FigChains.png}
+    \includegraphics[width=0.4\textwidth]{Figs/FigChains.png}
     \caption{Markov chains for the two parameters $K$ (top) and $\gamma$ (bottom).}
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.9\textwidth]{Figs/FigPhaseSpaceEvol.png}
+    \includegraphics[width=0.5\textwidth]{Figs/FigPhaseSpaceEvol.png}
     \caption{System dynamics in the phase space $K-\gamma$.
@@
 
-\begin{figure}
+\begin{figure}[htb!]
     \centering
-    \includegraphics[width=0.9\textwidth]{Figs/FigKg.png}
+    \includegraphics[width=0.5\textwidth]{Figs/FigKg.png}
     \caption{PDF for the parameters $K$ (left) and $\gamma$ (right).
@@
 
-\bibliographystyle{apalike} % chicago \shortcite(N)
-\bibliography{C:/Users/albertca/SWITCHdrive/refs}
+\subsection{Sample subsection}
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\subsubsection{Sample subsubsection}
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+\begin{figure}[htb!]
+ \centering
+ \includegraphics{lion.png}
+ \caption{\label{figureone}This is sample picture -- ShareLaTeX lion. }
+\end{figure}
+
+\subsection{Simplified coding and extra features for tables}
+The basic coding format can be simplified using extra commands provided in
+the \verb"iopart" class file. The commands up to and including
+the start of the tabular environment
+can be replaced by
+\small\begin{verbatim}
+\Table{\label{label}Table caption}
+\end{verbatim}\normalsize
+this also activates the definitions within \verb"\lineup".
+The final three lines can also be reduced to \verb"\endTable" or
+\verb"\endtab". Similarly for a table which does not fit in when indented
+\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable" or \verb"\endtab"
+can be used. \LaTeX\ optional positional parameters can, if desired, be added after
+\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".
+
+\verb"\centre{#1}{#2}" can be used to centre a heading
+\verb"#2" over \verb"#1"
+columns and \verb"\crule{#1}" puts a rule across
+\verb"#1" columns. A negative
+space \verb"\ns" is usually useful to reduce the space between a centred
+heading and a centred rule. \verb"\ns" should occur immediately after the
+\verb"\\" of the row containing the centred heading (see code for
+\tref{tabl1}). A small space can be
+inserted between rows of the table
+with \verb"\ms" and a half line space with \verb"\bs"
+(both must follow a \verb"\\" but should not have a
+\verb"\\" following them).
+
+\Table{\label{tabl1}A table with headings spanning two columns and containing notes.
+To improve the
+visual effect a negative skip ($\backslash${\tt ns})
+has been put in between the lines of the
+headings. Commands set-up by $\backslash${\tt lineup} are used to aid
+alignment in columns. $\backslash${\tt lineup} is defined within
+the $\backslash${\tt Table} definition.}
+\br
+&&&\centre{2}{Separation energies}\\
+\ns
+&Thickness&&\crule{2}\\
+Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
+\mr
+$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
+$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
+$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
+\br
+\end{tabular}
+\item[] $^{\rm a}$ Self-supporting.
+\item[] $^{\rm b}$ Deposited over Al backing.
+\end{indented}
+\end{table}
+
+\section*{References}
+\bibliographystyle{unsrt}
+\bibliography{refs}
+
+%\section*{References}
+%\begin{thebibliography}{10}
+%\bibitem{ref1} J.~Doe, Article name, \textit{Phys. Rev. Lett.}
+%\bibitem{ref2} J.~Doe, J. Smith, Other article name, \textit{Phys. Rev. Lett.}
+%\bibitem{web} \href{http://www.google.pl}{www.google.pl}
+%\end{thebibliography}
 

### 2015-07-16 Analysis_HMC_napa.jl

@@ -289,6 +289,8 @@ plt.figure(6)
 plt.xlabel("K")
 plt.ylabel("Gamma")
 plt.title("Evolution in K-Gamma space")
+axes()[:set_ylim]([0,6])
+axes()[:set_xlim]([0,450])
 grid = iround(linspace(1, size(thetas)[1], 1000))       # Number of points in K-gamma space to be plotted
 grid2 = grid[length(grid)-49:length(grid)]              # Last points in K-gamma space to be highlighted
 plt.plot(thetas[grid,1], thetas[grid,2], linewidth = 0.5)

### 2015-07-16 Main_HMC_napa_Maimai.jl

@@ -53,10 +53,10 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 2000
+const nsample_eff    = 20000
 const nsample        = nsample_eff + nsample_burnin
 
-const dtau           = 0.02                # MD time step
+const dtau           = 0.025               # MD time step
 const n_napa         = 5                   # Number of NAPA time steps
 # nn = 16  ; deltatau = dtau / nn          # Short time steps in RESPA --> NOT NEEDED IN THE NAPA IMPLEMENTATION
 
@@ -81,7 +81,7 @@ theta = [bet, gam]
 K_min   = 0.0
 gam_min = 0.0
 K_max   = 1000.0
-gam_max = 5.0
+gam_max = 10.0
 ##                 
 ## ============================================================================================
 ## Initialization of arrays
@@ -291,7 +291,7 @@ m_theta_burnin = m_theta
 
 m_bdy    = 700                      # m = m_q / dt
 m_stg    = 300                      # we assume m_q prop. to dt ==> m = costant     
-m_theta  = [150, 150]
+m_theta  = [250, 150]
 
 mp[1:nparams] = m_theta
 for s = 1:n  

### 2015-08-13 TeXFiles/Manuscript_Albert_Ulzega.tex
paper with intro etc.

@@
 
-\title[Parameter inference with stochastic models]{Parameter Inference for a Simple Stochastic Hydrological Model}
+\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models}
 
 \author[cor1]{Carlo Albert}%$^{1,2}$
-\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
+\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
 %\address{$^2$Address Two, Neverland}
@@
 \author{Simone Ulzega}
-\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 Dübendorf, Switzerland}
+\address{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}
 \ead{simone.ulzega@eawag.ch}
@@
 \begin{abstract}
-\textbf{[TO BE WRITTEN]} This document describes the  preparation of an article using \LaTeXe\ and \verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
-This class file is designed to help
-authors produce preprints in a form suitable for submission to any of the
-journals published by IOP Publishing.
-Authors submitting to any IOP journal, i.e.\
-both single- and double-column ones, should follow the guidelines set out here.
-On acceptance, their TeX code will be converted to
-the appropriate format for the journal concerned.
+Bayesian statistics has become an indispensable tool in many applied sciences, for the purpose of uncertainty analysis.  
+Inferring parametric uncertainty, for stochastic differential equation models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
+Here, we consider the generic problem of calibrating a 1D SDE model to time series and quantifying the ensuing parametric uncertainty.
+We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
+Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
+Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
+While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems.
+Furthermore the algorithm is highly parallelizable.
 
@@
 \vspace{2pc}
-\noindent{\it Keywords}: Article preparation, IOP journals
+\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations
+
 % Uncomment for Submitted to journal title message
@@
 
-In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
-We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
+In many applied areas of research phenomenological models are employed to predict the behaviour of complex systems of all sorts.
+Parameters of these models often don't have a direct physical meaning and need to be inferred from measurements.
+In order to make reliable predictions with such models it is important to describe the dominant errors in the model and to quantify the parametric uncertainty resulting from the inference process.
+
+{\em Bayesian statistics} describes knowledge about parameters through probability distributions and model-based learning through a consistent update rule.
+It is thus well suited to quantify parametric uncertainty resulting from a model based inference process.
+Bayesian statistics is commonly used in many applied sciences and grows in importance in the physics community as well \cite{vonToussaint2011bayesian}.
+
+A faithful description of the dominant errors in a model naturally leads to {\em stochastic differential equations} (SDEs).
+The kind of problems we consider here are the calibration of ordinary 1D SDE models to noisy time series and the quantification of the resulting parametric uncertainty.
+Whilst the techniques we use are generic they will be presented by means of a simple yet non trivial SDE model from hydrology.
+
+Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
+Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data.
+Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}.
+For the case of linear SDEs that are coupled to nonlinear deterministic ODEs, more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
+The problem with these simulation-based methods is their inefficiency in the presence of many data points.
+One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
+Such {\em Approximate Bayes Computations} \cite{Albert_2013_ABC} are relatively easy to apply as they only require us to run the simulator.
+However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.
+
+Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and simulation of the dynamics of the latter.
+Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987_HMC} incorporate the data points for the suggestion of new parameters whereby much higher acceptance rates are achieved.
+The drawback of these methods is that the model equations need to be known and derivatives have to be calculated.
+The latter problem, however, is largely remedied by the use of automated differentiation algorithms.
+
+Tuning of HMC algorithms is a non-trivial matter.
+Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
+It has been shown that efficiency is gained when the kinetic term is made dependent of the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
+Here, we explore a different, computationally simper route.
+Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
+Thus, we employ a multiple time scale integration technique, for the simulation of the statistical mechanical system \cite{tuckerman1993efficient}.
+At lest for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.
+
+The resulting algorithm is very efficient and, through the use of automated differentiation, easily applicable to a wide range of inference problems with SDEs.
+Furthermore, it is easily parallelizable.
+
+\section{An Exemplary SDE Model}
+
+The methods we present are generic and applicable to a wide range of SDE models.
+However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
+Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
+Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
+The model equation is thus given by the SDE
 \begin{equation}\label{sde}
@@
 \end{equation}
-where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
+where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
 \begin{equation}\label{whitenoise}
@@
 \end{equation}
-Due to the state-dependence of the noise term, Eq.~(\ref{sde}) is ill-defined unless we specify the discretization convention.
-For computational reasons, we work with the {\em Stratonovich} convention.
-Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see below).
+Eq.~(\ref{sde}) is to be understood in the {\em Stratonovich} sense.
+Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see eq.
+(\ref{equilibrium_mean}) below).
+Scale-invariance of the error term in (\ref{sde}) leads to power law tails in the system's distribution (see eq.
+(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref).
 
-The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
-Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
+Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
+Furthermore, by means of the transformation $S(t)=1/n(t)$, eq.
+(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors and extensively studied \cite{dutre_1977_SDE, fujisaka_1986_intermittency}.
+
+Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs will be studied elsewhere.
+Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with SDE models that are calibrated to observed time-series.
+Here, we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, which is observed, at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
+\begin{equation}\label{data}
+  \ln y_s
+  =
+  \ln\frac{S(t_s)}{K}
+  +
+  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,.
+\end{equation}
+For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data (\ref{data}).
+Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
+In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{prior}(\bt)$, and the posterior, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
+\begin{equation}\label{Bayes}
+  f_{post}(\bt|\vc y)
+  =
+  \frac{f_{prior}(\bt)L(\vc y|\bt)}
+  {f_{prior}(\bt')L(\vc y|\bt')d\bt'}\,,
+\end{equation}
+where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution, for model outputs given model parameters, evaluated at the measured data.
+Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior to be flat on an area large enough to encompass the area where the likelihood function is significantly different from zero.
+That is, the posterior is data-driven.
+
+Before we set out to derive the likelihood function from the model equations (\ref{sde}) through (\ref{data}) let us express the parameters and state variable with dimensionless quantities.
+The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
+We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
 \begin{equation}
@@
 \end{equation}
+For our algorithm it is important to have the model equation in a form where the noise term does not depend on any of the variables that have to be inferred, i.e., on neither the state variable nor the parameters that have to be inferred.
+At least in 1D this can always be achieved through re-parametrization.
 
-According to Ref.~\cite{lau_2007}, the probability $P(q_1,T|q_0,0)$ of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
+The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$, at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
 \begin{equation}\label{pathint}
@@
 where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
-The path-measure $\mathcal Dq$ is formally written as an infinite product
+The path-measure $\mathcal Dq$ is formally written as the infinite product
 \begin{equation}\label{pathmeasure_q}
@@
 \end{equation}
-Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.
+Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$ \cite{lau_2007_StateDepDiff}.
 
@@
 \end{equation}
-and rewrite the action~(\ref{action1}) as,
+and rewrite action~(\ref{action1}) as,
 \begin{eqnarray}\label{action}
@@
 
-Next, we calculate the equilibrium distribution $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$.
-Systems at thermal equilibrium must fulfill detailed balance, i.e.,
+Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model.
+However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$.
+After plugging (\ref{pathint}) and (\ref{action}) into the detailed balance condition,
 \begin{equation}\label{detailed_balance}
-P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,.
+P(q_1 t_1 | q_0 t_0 ) P_{\eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\eq}(q_1) \,,
 \end{equation}
-After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
+and using transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ onto paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
 \begin{equation}\label{Peq}
@@
 \end{equation}
-Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
+Transforming back to the original variables, it turns out that $P_{\eq}(S)$ is an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
 \begin{equation}\label{inverse_gamma}
@@
 whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
-\begin{equation}
+\begin{equation}\label{equilibrium_mean}
   \langle S\rangle_{\eq}=Kr_{0},
@@
   K^2r_{0}^2
-  \frac{\gamma}{2-\gamma}\,.
+  \frac{\gamma}{2-\gamma}\,
 \end{equation}
-Note that the variance diverges, for $\gamma\geq 2$.
+and diverges, for $\gamma\geq 2$.
 The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.
 
-Let us now consider two time-series of {\em observed} rain inputs and discharge outputs, $r_s=r(t_s)$ and $y_s=S(t_s)/K$, respectively, measured at times $0=t_1<t_2<\dots < t_{n+1}=T$.
-For the measurement error model, we simply set
-\begin{equation}\label{data}
-  \ln\frac{y_s}{r_s}
-  =
-  \beta q_s
-  +
-  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
-\end{equation}
-where $\epsilon_s$ are uncorrelated standard normal random variables.
-
-The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
-For the time being we assume constant priors.
-Then, the posterior reads as
+If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to likelihood function, which is expressed as a path-integral as follows
 \begin{equation}\label{posterior_pathint}
-  f(\bt | \vc y, \vc r)
+  f(\bt | \vc y)
   \propto
@@
     \frac
-    {(\ln(y_s/r_s)-{\beta q_s})^2}
+    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
     {\sigma^2}
@@
 \end{equation}
-Since the integral on the r.h.s.
-of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
-Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by Ref.~\cite{duane_1987}.
-This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Inference Algorithm}
+
+In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987_HMC}.
+The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the field $q(t)$.
 Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
@@
 \end{equation}
-with
+where
 \begin{equation}\label{K}
@@
 \end{equation}
-where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
+and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
 The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
 \begin{equation}\label{phaseSpacePathInt}
-    f(\bt | \vc y, \vc r)
+    f(\bt | \vc y)
   \propto
@@
   {\mathcal Dp}
-  \wedge
    {\mathcal Dq}
-  \wedge
    d{\pmb\pi}
@@
 
-The HMC method, which is a combination of the {\em Metropolis algorithm} of Ref.~\cite{metropolis_1953} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps,
+The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \cite{Alder_1959_MD, Rahman_1964_MD}, iterates the following steps:
 \begin{enumerate}
@@
   \item
-  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
+  The system is integrated by means of a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
   \item
@@
 \end{enumerate}
-The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.
+The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate.
 
-Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
+In order to simulate the dynamics of Hamiltonian (\ref{Hamiltonian}), we need to discretize the path-integral (\ref{phaseSpacePathInt}).
 Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
 Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
-Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure
+Path-integral (\ref{phaseSpacePathInt}) is approximated by an ordinary integral, with the approximate path-measure
 \begin{equation}
-  \mathcal Dp\wedge\mathcal Dq
+  \mathcal Dp\mathcal Dq
   \approx
-  \prod_i dp_i\wedge dq_i
+  \prod_i dp_i dq_i
 \end{equation}
@@
 \begin{equation}\label{rhodisc}
-\rho_i = \frac{T}{\beta} \frac{\ln(r_{i}/r_{i-1})}{\Delta t}
+\rho_i = \frac{T}{\beta} \frac{\ln(r(t_{i})/r(t_{i-1}))}{\Delta t}
 +
@@
 \end{equation}
-Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.
+Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}).
 
-For strongly varying rain input or high-frequency output data, $\Delta t$ needs to be chosen small.
-Thus, the harmonic part of the Hamiltonian becomes stiff and, consequently, the time-step of MD needs to be chosen small, which leads to slow convergence.
-To solve this problem, we change coordinates, such that the harmonic part of the Hamiltonian becomes (partly) diagonal and the particle interaction moves to the potential.
-One option would be to use the Fourier modes of ${\eta}(t)$ as configurational variables.
-Calculating the potential requires then an inversion of a Toeplitz matrix.
-This approach is interesting because, in our case, it allows for an approximate analytical integration of high-frequency modes as outlined in Ref.~\cite{doll_1985}.
+The discretized Hamiltonian describes a {\em classical polymer chain} with harmonic bonds between neighbouring beads in an external field.
+The latter consists of two parts, a field that results from the dynamics of the original equations (\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.
 
-We choose however another approach, the so-called \emph{staging} as described in Ref.~\cite{tuckerman_1992}, and diagonalize the harmonic part only in between the measurement points.
-We first note that the Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
-Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
-Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.
+The potential (\ref{Vdisc}) contains terms with different scaling in the potentially large numbers $N$ and $n$, which describe dynamics on different time-scales.
+In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (iii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
+An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
+We choose an exact approach and employ a multiple time scale integration based on Trotter's formula \cite{tuckerman1993efficient}.
 
-The discretized harmonic part of the action is rewritten as
+For this purpose it proves to be useful to introduce so-called \emph{staging} variables \cite{tuckerman1992reversible}, and diagonalize the harmonic part in between the measurement points.
+Therefore, we rewrite the discretized harmonic part of the action as
 \begin{eqnarray}
@@
 \end{equation}
-Following Ref.~\cite{tuckerman_1993} we apply the coordinate transformation
+We apply the coordinate transformation
 \begin{equation}
@@
 \end{equation}
-or equivalently by the recursive relation
+or, equivalently, by the recursive relation
 \begin{equation}
@@
 
-This transformation decouples fast degrees of freedom from slow ones and allows us to split the Hamiltonian $\mathcal{H}_{\mbox{\tiny HMC}}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$.
-We write
+We split the Hamiltonian $\mathcal{H}_{\mbox{\tiny HMC}}$ into components with different scaling behaviour in the potentially large numbers $n$ and $N$ and write
 \begin{equation}
@@
 The harmonic part for the staging beads given in Eq.~(\ref{H_N}) scales linearly with $N$.
-The term of Eq.~(\ref{H_n}), including both the harmonic part for the boundary beads and the measurement term, scales linearly with $n$.
+The terms of Eq.~(\ref{H_n}), including both the harmonic part for the boundary beads and the measurement term, scale linearly with $n$.
 Finally, Eq.~(\ref{H_1}) does not scale with neither $n$ nor $N$.
-Thus, depending on the values of $N$ and $n$, the dynamics happens on very different time scales.
-Therefore, we use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
-In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes.
+Thanks to the staging variables $\mathcal H_N$ decouples completely from $\mathcal H_n$.
+
+We use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
+In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
 \begin{enumerate}
   \item[\it i.]
-  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$
+  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
   \item[\it ii.]
-  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$
+  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
   \item[\it iii.]
-  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$
+  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
 \end{enumerate}
-
+Here, we restrict ourselves to regime ($ii$), i.e., we assume that we do not have too many measurements and/or the measurement error is not too small.
+The generalization of the method to the other schemes is straightforward.
 In regime ($ii$) we simply separate the harmonic part of the action, for the staging beads, from the rest and write
@@
 where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
-Trotter's formula allows us to write the Hamiltonian propagator as
+Trotter's formula \cite{Trotter_1959_product} allows us to write the Hamiltonian propagator as
 \begin{equation}
@@
   +
-  \mathcal O(P^{-2})\,,
+  \mathcal O(\tau^3/P^{2})\,,
 \end{equation}
@@
 
-For the inner, slow propagator, we employ the velocity Verlet algorithm.
+For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{Swope_1982_Verlet}.
 For the boundary beads, it reads
@@
 
-\section{Results and discussions}
+\section{Implementation}
+
+AD
+
+parallelization
+
+\section{Numerical Results}
 
@@
 
-\subsection{Sample subsection}
+\section{Conclusions}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-\subsubsection{Sample subsubsection}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-\begin{figure}[htb!]
- \centering
- \includegraphics{lion.png}
- \caption{\label{figureone}This is sample picture -- ShareLaTeX lion. }
-\end{figure}
-
-\subsection{Simplified coding and extra features for tables}
-The basic coding format can be simplified using extra commands provided in
-the \verb"iopart" class file. The commands up to and including
-the start of the tabular environment
-can be replaced by
-\small\begin{verbatim}
-\Table{\label{label}Table caption}
-\end{verbatim}\normalsize
-this also activates the definitions within \verb"\lineup".
-The final three lines can also be reduced to \verb"\endTable" or
-\verb"\endtab". Similarly for a table which does not fit in when indented
-\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable" or \verb"\endtab"
-can be used. \LaTeX\ optional positional parameters can, if desired, be added after
-\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".
-
-\verb"\centre{#1}{#2}" can be used to centre a heading
-\verb"#2" over \verb"#1"
-columns and \verb"\crule{#1}" puts a rule across
-\verb"#1" columns. A negative
-space \verb"\ns" is usually useful to reduce the space between a centred
-heading and a centred rule. \verb"\ns" should occur immediately after the
-\verb"\\" of the row containing the centred heading (see code for
-\tref{tabl1}). A small space can be
-inserted between rows of the table
-with \verb"\ms" and a half line space with \verb"\bs"
-(both must follow a \verb"\\" but should not have a
-\verb"\\" following them).
-
-\Table{\label{tabl1}A table with headings spanning two columns and containing notes.
-To improve the
-visual effect a negative skip ($\backslash${\tt ns})
-has been put in between the lines of the
-headings. Commands set-up by $\backslash${\tt lineup} are used to aid
-alignment in columns. $\backslash${\tt lineup} is defined within
-the $\backslash${\tt Table} definition.}
-\br
-&&&\centre{2}{Separation energies}\\
-\ns
-&Thickness&&\crule{2}\\
-Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
-\mr
-$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
-$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
-$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
-\br
-\end{tabular}
-\item[] $^{\rm a}$ Self-supporting.
-\item[] $^{\rm b}$ Deposited over Al backing.
-\end{indented}
-\end{table}
+We've presented a new algorithm, for the generation of posterior parameter samples, for ordinary 1D SDE models that are calibrated to time-series.
+The algorithm is derived from a re-interpretation of the posterior distribution as a partition function of a 1D statistical mechanical system and employs a Hamiltonian Monte Carlo approach combined with a multiple time-scale integration.
+Furthermore, a generic re-parametrization is suggested, which decouples harmonic modes in between measurement points from both the measurement potential and the model parameters, and allows for an efficient analytical integration of these modes.
+The algorithm can be implemented in an efficient and generic fashion using automated differentiation and parallelization.
+Furthermore, it can easily be adapted to other inference problems.
+In particular, it can be adapted to higher dimensional SDEs and SDEs coupled to ODEs.
+We explore these adaptations in our future work.
 
@@
 \bibliographystyle{unsrt}
-\bibliography{refs}
+\bibliography{C:/Users/albertca/SWITCHdrive/refs}
 

### 2015-08-17 Main_HMC_napa_Maimai.jl
temp_data contains first attempt with real rain

@@ -53,10 +53,10 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 20000
+const nsample_eff    = 50000
 const nsample        = nsample_eff + nsample_burnin
 
-const dtau           = 0.025               # MD time step
+const dtau           = 0.02                # MD time step
 const n_napa         = 5                   # Number of NAPA time steps
 # nn = 16  ; deltatau = dtau / nn          # Short time steps in RESPA --> NOT NEEDED IN THE NAPA IMPLEMENTATION
 
@@ -81,7 +81,7 @@ theta = [bet, gam]
 K_min   = 0.0
 gam_min = 0.0
 K_max   = 1000.0
-gam_max = 10.0
+gam_max = 1.0
 ##                 
 ## ============================================================================================
 ## Initialization of arrays

### 2015-08-17 TeXFiles/Manuscript_Albert_Ulzega.tex
some work (Simone) on the paper

@@
 \begin{abstract}
-Bayesian statistics has become an indispensable tool in many applied sciences, for the purpose of uncertainty analysis.  
-Inferring parametric uncertainty, for stochastic differential equation models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
-Here, we consider the generic problem of calibrating a 1D SDE model to time series and quantifying the ensuing parametric uncertainty.
-We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
-Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
+Bayesian statistics has become an indispensable tool in many applied sciences for the purpose of uncertainty analysis.
+Inferring parametric uncertainty for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
+Here we present a Monte Carlo-based method for Bayesian parameter inference with one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
+%Here, we consider the generic problem of calibrating a one dimensional SDE model to %time series and quantifying the ensuing parametric uncertainty.
+We re-interpret the Bayesian posterior distribution for model parameters as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
+Depending on the number of discretization points and the number of measurement points the dynamics of this system happen on very different time scales.
 Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
-While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems.
-Furthermore the algorithm is highly parallelizable.
+While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems. Furthermore the algorithm is highly parallelizable.
 
@@
 
-In many applied areas of research phenomenological models are employed to predict the behaviour of complex systems of all sorts.
+Phenomenological models are employed in many applied areas of research to predict the behaviour of complex systems of all sorts.
 Parameters of these models often don't have a direct physical meaning and need to be inferred from measurements.
@@
 It is thus well suited to quantify parametric uncertainty resulting from a model based inference process.
-Bayesian statistics is commonly used in many applied sciences and grows in importance in the physics community as well \cite{vonToussaint2011bayesian}.
+Bayesian statistics is commonly used in many applied sciences and is growing in importance in the physics community as well \cite{vonToussaint_2011}.
 
@@
 Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}.
-For the case of linear SDEs that are coupled to nonlinear deterministic ODEs, more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
+For the case of linear SDEs that are coupled to nonlinear deterministic ordinary differential equations (ODEs), more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
 The problem with these simulation-based methods is their inefficiency in the presence of many data points.
 One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
-Such {\em Approximate Bayes Computations} \cite{Albert_2013_ABC} are relatively easy to apply as they only require us to run the simulator.
+Such {\em Approximate Bayes Computations} \cite{Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator.
 However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.
 
-Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and simulation of the dynamics of the latter.
-Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987_HMC} incorporate the data points for the suggestion of new parameters whereby much higher acceptance rates are achieved.
+Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
+Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987} incorporate the data points for the suggestion of new parameters, whereby much higher acceptance rates are achieved.
 The drawback of these methods is that the model equations need to be known and derivatives have to be calculated.
@@
 Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
-It has been shown that efficiency is gained when the kinetic term is made dependent of the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
-Here, we explore a different, computationally simper route.
-Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
-Thus, we employ a multiple time scale integration technique, for the simulation of the statistical mechanical system \cite{tuckerman1993efficient}.
-At lest for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.
+It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
+Here we explore a different, computationally simper route.
+Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales.
+Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
+At least for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.
 
@@
 \end{equation}
-Eq.~(\ref{sde}) is to be understood in the {\em Stratonovich} sense.
+Eq.~(\ref{sde}) is to be understood in the {\em Stratonovich} sense (ref \textbf{Stratonovich}).
 Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see eq.
 (\ref{equilibrium_mean}) below).
-Scale-invariance of the error term in (\ref{sde}) leads to power law tails in the system's distribution (see eq.
-(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref).
+Scale-invariance of the model of eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref \textbf{FF}).
 
@@
 
-Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs will be studied elsewhere.
+Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs \cite{tessier_1996} will be studied elsewhere.
 Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with SDE models that are calibrated to observed time-series.
-Here, we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, which is observed, at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
+Here we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
 \begin{equation}\label{data}
-  \ln y_s
+  \ln \left( y_s \right)
   =
-  \ln\frac{S(t_s)}{K}
+  \ln \left( \frac{S(t_s)}{K} \right)
   +
@@
 \end{equation}
-For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data (\ref{data}).
+For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by eq.~(\ref{data}).
 Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
-In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{prior}(\bt)$, and the posterior, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
+In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{\mbox{\small prior}}(\bt)$, while the posterior distribution, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
 \begin{equation}\label{Bayes}
-  f_{post}(\bt|\vc y)
+  f_{\mbox{\small post}}(\bt|\vc y)
   =
-  \frac{f_{prior}(\bt)L(\vc y|\bt)}
-  {f_{prior}(\bt')L(\vc y|\bt')d\bt'}\,,
+  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
+  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
 \end{equation}
-where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution, for model outputs given model parameters, evaluated at the measured data.
+where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution for model outputs given model parameters, evaluated at the measured data.
 Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior to be flat on an area large enough to encompass the area where the likelihood function is significantly different from zero.
@@
   =
-  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
+  \frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
   +
@@
 \end{equation}
-For our algorithm it is important to have the model equation in a form where the noise term does not depend on any of the variables that have to be inferred, i.e., on neither the state variable nor the parameters that have to be inferred.
-At least in 1D this can always be achieved through re-parametrization.
+For our algorithm it is important to have the model equation in a form where the noise term does not depend on neither the state variables nor the parameters that have to be inferred.
+In a one dimensional model this can always be achieved through re-parametrization.
 
-The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$, at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
+The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
 \begin{equation}\label{pathint}
@@
 \end{equation}
-The {\em action} is a functional on the space of paths and reads as
+The {\em action} is a functional on the space of paths and reads as \cite{lau_2007}
 \begin{equation}\label{action1}
@@
 \end{equation}
-Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$ \cite{lau_2007_StateDepDiff}.
+Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.
 
@@
 
-Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model.
+Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{dutre_1977_SDE, fujisaka_1986_intermittency} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model.
 However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$.
@@
 
-If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to likelihood function, which is expressed as a path-integral as follows
+If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as a path-integral as follows
 \begin{equation}\label{posterior_pathint}
-  f(\bt | \vc y)
+  f_{\mbox{\small post}}(\bt | \vc y)
   \propto
@@
   \exp\bigg[
-    -{{\mathcal S}}[{q},\dot q]
     -\frac{1}{2}
@@
     {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
-    {\sigma^2}
+    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
     %-\ln(K\gamma)
@@
 \end{equation}
+On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$.
+The second term is the probability of a system realization $q$ and the path-integral extends over all possible system realizations.
 
@@
 
-In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987_HMC}.
+In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
 The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the field $q(t)$.
@@
 \begin{equation}\label{phaseSpacePathInt}
-    f(\bt | \vc y)
+    f_{\mbox{\small post}}(\bt | \vc y)
   \propto
@@
 
-The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \cite{Alder_1959_MD, Rahman_1964_MD}, iterates the following steps:
+The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
 \begin{enumerate}
@@
 An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
-We choose an exact approach and employ a multiple time scale integration based on Trotter's formula \cite{tuckerman1993efficient}.
+We choose an exact approach and employ a multiple time scale integration based on Trotter's formula \cite{tuckerman_1993}.
 
-For this purpose it proves to be useful to introduce so-called \emph{staging} variables \cite{tuckerman1992reversible}, and diagonalize the harmonic part in between the measurement points.
+For this purpose it proves to be useful to introduce so-called \emph{staging} variables \cite{tuckerman_1992}, and diagonalize the harmonic part in between the measurement points.
 Therefore, we rewrite the discretized harmonic part of the action as
@@
 
-We use the Trotter formula according to \cite{tuckerman1992reversible} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
+We use the Trotter formula according to \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes these different time scales into account.
 In order to design an appropriate partition of the Hamiltonian, we need to distinguish different regimes, such as
@@
 where $\{\cdot\,,\,\cdot\}$ denotes the Poisson brackets that are defined on functions on the phase space.
-Trotter's formula \cite{Trotter_1959_product} allows us to write the Hamiltonian propagator as
+Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
 \begin{equation}
@@
 
-For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{Swope_1982_Verlet}.
+For the inner, slow propagator, we employ the velocity Verlet algorithm \cite{swope_1982_verlet}.
 For the boundary beads, it reads
@@
 \bibliographystyle{unsrt}
-\bibliography{C:/Users/albertca/SWITCHdrive/refs}
-
-%\section*{References}
-%\begin{thebibliography}{10}
-%\bibitem{ref1} J.~Doe, Article name, \textit{Phys. Rev. Lett.}
-%\bibitem{ref2} J.~Doe, J. Smith, Other article name, \textit{Phys. Rev. Lett.}
-%\bibitem{web} \href{http://www.google.pl}{www.google.pl}
-%\end{thebibliography}
+\bibliography{refs}
 

### 2015-08-19 TeXFiles/Manuscript_Albert_Ulzega.tex
incomplete new paper version (Simone)

@@
 \usepackage[breaklinks=true,colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
+\usepackage{url}
+\usepackage{caption}
 %\usepackage{amsmath,amssymb,amsthm,amscd}
@@
 On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$.
-The second term is the probability of a system realization $q$ and the path-integral extends over all possible system realizations.
+The second term is the probability of the given system realization $q$ and the path-integral extends over all possible system realizations.
+Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
 
@@
 In order to derive an efficient algorithm to draw parameter samples from (\ref{posterior_pathint}) we interpret it as the partition function of a 1D statistical mechanical system and simulate the dynamics of the latter, employing the so-called {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}.
-The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the field $q(t)$.
+The model parameters $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
 Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
@@
 and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
-The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
+The posterior (\ref{posterior_pathint}) can then be expressed by the phase space path integral
 \begin{equation}\label{phaseSpacePathInt}
@@
   \item
-  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
+  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by eq.~(\ref{K}).
   \item
-  The system is integrated by means of a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
+  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
+  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
   \item
@@
 The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate.
+Each new phase space configuration corresponds to a well-defined combination of model parameters $\bt$.
+Thus, the simulated succession of configurations represent a Markov chain of sampled parameter vectors that naturally leads to the sought probability distribution for the system parameters.
 
-In order to simulate the dynamics of Hamiltonian (\ref{Hamiltonian}), we need to discretize the path-integral (\ref{phaseSpacePathInt}).
+In order to simulate the dynamics of the Hamiltonian (\ref{Hamiltonian}), we first need to discretize the path-integral (\ref{phaseSpacePathInt}).
 Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
 Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
-Path-integral (\ref{phaseSpacePathInt}) is approximated by an ordinary integral, with the approximate path-measure
+The path-integral (\ref{phaseSpacePathInt}) is then approximated by an ordinary integral, with the approximate path-measure
 \begin{equation}
@@
 
-The discretized Hamiltonian describes a {\em classical polymer chain} with harmonic bonds between neighbouring beads in an external field.
+The discretized Hamiltonian describes a {\em classical polymer chain} of $N$ beads with harmonic bonds between neighbouring beads in an external field.
 The latter consists of two parts, a field that results from the dynamics of the original equations (\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.
+The masses $m_q$ and $m_{\alpha}$ are tunable parameters of the algorithm.
+A good strategy consists in assigning larger weights to the beads associated with measurement points than to the intermediate beads.
+The dynamics of the system will be therefore mostly dependent on the lighter intermediate particles, while the heavy measurement beads will be constrained in the vicinity of the measured data points.
+We thus reduce the original Bayesian inference problem to simulating the dynamics of a linear polymer, as sketched in Fig.~(\ref{fig:polymer}).
+In this context the physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced (vertical axis).
+Each \emph{state} of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$.
+
+\begin{figure}[htb!]
+    \centering
+    \captionsetup{justification=justified}
+    \includegraphics[width=0.7\textwidth]{Figs/FigPolymer.png}
+    \caption{Simulated dynamics of a polymer chain with $n+1=11$ data points (large filled circles) and $j-1=9$ intermediate beads (small empty circles).
+All other parameters are discussed in Section~\ref{numerical_results}.
+Top: initial state.
+The intermediate beads represent a linear interpolation to the data points.
+Bottom: polymer configuration after 1000 iterations of the propagation algorithm.
+The dotted line represents the initial configuration.
+It is evident that the new configuration is mostly determined by the dynamics of the low-mass intermediate beads, while the heavy-mass data points barely move from their initial positions.}
+    \label{fig:polymer}
+\end{figure}
 
 The potential (\ref{Vdisc}) contains terms with different scaling in the potentially large numbers $N$ and $n$, which describe dynamics on different time-scales.
-In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (iii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
+In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (ii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
 An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
@@
 \end{equation}
-We apply the coordinate transformation
+We apply the coordinate transformations
 \begin{equation}
-  u_{sj+1} = q_{sj+1}\,,\quad
-  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
+  u_{sj+1} = q_{sj+1}\,, \qquad\qquad s=0,\dots,n\, ,
 \end{equation}
-with its inverse given by
+for the boundary beads, corresponding to the original measurement points, and
 \begin{equation}
-  q_{sj+1} = u_{sj+1}\,,\quad
+  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,, \quad s=0,\dots,n-1\, , \quad k=2,\dots,j\, ,
+\end{equation}
+for the intermediate staging beads.
+The inverse transformations are given by
+\begin{eqnarray}
+  q_{sj+1} = u_{sj+1}\,,
+  \\
   q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
-  +\frac{j-k+1}{j}u_{sj+1}\,,
-\end{equation}
-or, equivalently, by the recursive relation
+  +\frac{j-k+1}{j}u_{sj+1}\,.
+\end{eqnarray}
+The latter can be equivalently expressed by the recursive relation
 \begin{equation}
@@
 \end{enumerate}
-Here, we restrict ourselves to regime ($ii$), i.e., we assume that we do not have too many measurements and/or the measurement error is not too small.
+Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is small and/or the measurement error $\sigma$ is large.
 The generalization of the method to the other schemes is straightforward.
@@
 \end{equation}
-In order to design a reversible integrator, for the associated Hamiltonian equations, we define the Liouville operators
+In order to design a reversible integrator for the associated Hamilton equations, we define the Liouville operators
 \begin{equation}
@@
 Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
-\begin{equation}
+\begin{equation}\label{propagator}
   e^{i(L_N+L')\tau}
@@
 for $\tau =P\Delta \tau$.
-In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes a much faster dynamics than the inner one.
+In regime ($ii$) the outer propagator $\exp[iL_N(\Delta \tau/2)]$ describes much faster dynamics than the inner one.
 However, it is the dynamics of uncoupled harmonic oscillators, which we can readily solve.
@@
 \end{eqnarray}
-where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
-For the model parameters, analogous equations apply.
+with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
+For the model parameters $\bt$ and their momenta ${\pmb\pi}$, analogous equations apply.
 When it comes to the staging beads, however, only the momenta have to be updated, because the associated kinetic term is not part of $\mathcal H'$ but of $\mathcal H_N$.
@@
   )
-  \,.
+  \,,
 \end{eqnarray}
+with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over the time $\tau$.
 
-\section{Implementation}
+\section{Numerical Results}\label{numerical_results}
 
-AD
+The algorithm described above was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9)~\cite{julia_web}.
+One of the attractive features of the language is its (fast growing) library of over 600 packages for numerical and scientific applications.
+In our case, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions.
+Our algorithm benefits greatly from the use of AD.
+Indeed, it gives us the possibility to modify the model of eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered.
+This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
 
-parallelization
+We have run preliminary simulations on a 64-bit laptop computer equipped with a Windows 7 operating system, a double-core 1.6~GHz CPU (Intel i5-4200U) and 8~GB of RAM.
+We have initially used a serial implementation of the algorithm.
+We assumed $n+1 = 11$ measurement points and $j = 10, 20,\dots, 50$ intermediate staging beads, with a total number of discretization points $N = 101, 201, \dots, 501$.
+We have set $\Delta\tau = 0.25$ (arbitrary units of time) and $P = 3$ in the Hamiltonian propagator~(\ref{propagator}), with a constant total observation time $T = 833$ (arbitrary units of time).
+The "true" values of the parameters to be inferred were set to $K_{\mbox{\small true}}=50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
+Their initial values in our simulations were set to $K=200$~(arbitrary units of time) and $\gamma = 0.5$.
 
-\section{Numerical Results}
-
-The results shown in the figures refer to synthetic data obtained assuming a simple sinusoidal model for the rain input.
-The experimental data were then generated using the "true" parameter values $K=200$ and $\gamma = 0.2$.
-We considered $n=51$ data points, $j=10$ staging points, and therefore $N=501$ discretization points.
-The total time $T$ is $\approx 833$ [units of time], so that the discretization points are separated by a time interval $\Delta t \approx 1.6 $ [units of time].
-Following the RESPA method described above, we used a long time step $\Delta \tau = 0.02$ [units of time] for the evolution of the slow (long-range) potential, further subdivided in 16 smaller steps for the evolution of the fast (short-range) potential.
+The results shown here were obtained assuming a simple sinusoidal input $r(t) = \sin^2 \left( 0.01 t \right) + 0.1$.
+A system realization was first obtained from eq.~(\ref{sde}) using $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
+Such system realization was then used to generate a synthetic time series of observed data according to eq.~(\ref{data}).
+The error $\sigma$ was set to $0.1$.
+The input signal, the "true" system realization and the data time series are shown in Fig.~(\ref{fig:rain_data_S})
 
@@
     \centering
-    \includegraphics[width=0.4\textwidth]{Figs/FigRainData.png}
-    \caption{\label{fig:rain_data_S}Rain input (dashed), system realization (solid line) and synthetic experimental data.
-The system output and the observed data were generated using $K=200$ and $\gamma = 0.2$.}
+    \includegraphics[width=0.6\textwidth]{Figs/FigRainData.png}
+    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data.
+One may notice how the system response follows the oscillations of the input signal.}
 \end{figure}
+
+We observed a reasonably linear dependence of the total computing time on the number of discretization points $N$.
+In a typical case with $j=30$ (i.e., $N = 301$ discretization points) a complete run with 20000 iterations required about 13 minutes.
+A set of 200 simulated system realizations is shown in Fig.~(\ref{fig:spaghetti}) together with the synthetic data generated as described above.
+In the simulation the masses were set to 720 for the boundary beads, 130 for the staging beads and 150 for both the dimensionless parameters $\beta$ and $\gamma$.
+The different dynamics of the heavy boundary beads and the light staging beads can be appreciated in Fig.~(\ref{fig:spaghetti}).
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{Figs/FigSpaghetti.png}
-    \caption{Simulated system realizations and synthetic experimental data obtained assuming a simple sinusoidal model for the rain input.
-In the inset one may appreciate the different dynamics of the heavy data-point beads and the light staging beads.}
+    \includegraphics[width=0.7\textwidth]{Figs/FigSpaghetti.png}
+    \caption{Simulated system realizations and corresponding synthetic data generated as described in the text.
+In the inset one may appreciate the different dynamics of heavy data-point and light staging beads.}
     \label{fig:spaghetti}
@@
     \centering
-    \includegraphics[width=0.3\textwidth]{Figs/FigFinalState.png}
-    \caption{PDF for the system final state expressed via the coordinate set $\left\{ q_N \right\}$.
-The dashed curve represents a normal distribution with the same mean and variance.
-The final state is basically normally distributed.}
-    \label{fig:final_state}
+    \includegraphics[width=0.7\textwidth]{Figs/FigChainK.png}
+    \caption{Markov chain for the inferred parameter $K$.}
+    \label{fig:chainK}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.4\textwidth]{Figs/FigChains.png}
-    \caption{Markov chains for the two parameters $K$ (top) and $\gamma$ (bottom).}
-    \label{fig:chains}
+    \includegraphics[width=0.7\textwidth]{Figs/FigChainG.png}
+    \caption{Markov chain for the inferred parameter $\gamma$.}
+    \label{fig:chainG}
 \end{figure}
+
+The Markov chains for 50000 iterations for the parameters $K$ and $\gamma$ are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively.
+Both of them are clearly compatible with the parameter values $K_{\mbox{\small true}} = 50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
+The system evolution in phase space $K-\gamma$ is shown in Fig.~(\ref{fig:phase_space_evol}).
+
+%\begin{figure}[htb!]
+%    \centering
+%    \includegraphics[width=0.3\textwidth]{Figs/FigFinalState.png}
+%    \caption{PDF for the system final state expressed via the coordinate set $\left\{ q_N \right\}$.
+The dashed curve represents a normal distribution with the same mean and variance.
+The final state is basically normally distributed.}
+%    \label{fig:final_state}
+%\end{figure}
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{Figs/FigPhaseSpaceEvol.png}
-    \caption{System dynamics in the phase space $K-\gamma$.
-The green dot is the initial state, the red dot is the final state, the yellow dots are the last 50 states of the HMC chain, and the orange square corresponds to the true parameter values used to generate the data.}
+    \includegraphics[width=0.7\textwidth]{Figs/FigPhaseSpaceEvol.png}
+    \caption{System dynamics in the phase space $K-\gamma$. The empty circle is the initial state, while the square corresponds to the true parameter values used to generate the data.}
     \label{fig:phase_space_evol}
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.5\textwidth]{Figs/FigKg.png}
-    \caption{PDF for the parameters $K$ (left) and $\gamma$ (right).
+%\begin{figure}[htb!]
+%    \centering
+%    \includegraphics[width=0.5\textwidth]{Figs/FigKg.png}
+%    \caption{PDF for the parameters $K$ (left) and $\gamma$ (right).
 The true values used to generate the data are $K=200$ and $\gamma = 0.2$.
 The initial values used in the simulations were $K=50$ and $\gamma = 0.4$.}
-    \label{fig:K_distr}
-\end{figure}
+%    \label{fig:K_distr}
+%\end{figure}
 

### 2015-08-19 Main_HMC_napa_sinrain.jl

@@ -55,7 +55,7 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 50000
+const nsample_eff    = 20000
 const nsample        = nsample_eff + nsample_burnin
 
 const dtau           = 0.25                # MD time step
@@ -289,9 +289,9 @@ m_bdy_burnin   = m_bdy
 m_stg_burnin   = m_stg
 m_theta_burnin = m_theta
 
-m_bdy    = 680                      # m = m_q / dt
-m_stg    = 300                      # we assume m_q prop. to dt ==> m = costant     
-m_theta  = [150, 150]
+m_bdy    = 800                      # m = m_q / dt
+m_stg    = 120                      # we assume m_q prop. to dt ==> m = costant     
+m_theta  = [130, 130]
 
 mp[1:nparams] = m_theta
 for s = 1:n  
@@ -414,7 +414,7 @@ end
 ## Save parameters and results
 ## ============================================================================================
 ##
-fname = string("_sinr_680_150") 
+fname = string("_sinr800_120") 
 
 param_names  = vcat("N", "j", "n", "t[1]", "dt", "nparams", "true_K", "true_gam", "sigma", "K", "gam", 
 	"nsample_burnin", "nsample_eff", "m_bdy_burnin", "m_bdy", "m_theta_burnin", "m_theta_bet", 

### 2015-08-20 TeXFiles/Manuscript_Albert_Ulzega.tex
completed Simone's version of the paper, going back to Carlo

@@
 \end{equation}
-Eq.~(\ref{sde}) is to be understood in the {\em Stratonovich} sense (ref \textbf{Stratonovich}).
-Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see eq.
-(\ref{equilibrium_mean}) below).
-Scale-invariance of the model of eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref \textbf{FF}).
+Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
+Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
+Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref \textbf{FF}).
 
@@
 \end{equation}
-For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by eq.~(\ref{data}).
+For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}).
 Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
@@
   \item
-  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by eq.~(\ref{K}).
+  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
   \item
@@
 Our algorithm benefits greatly from the use of AD.
-Indeed, it gives us the possibility to modify the model of eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered.
+Indeed, it gives us the possibility to modify the model of Eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered.
 This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
+%
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
+    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data.
+One may notice how the system response follows the oscillations of the input signal.}
+\end{figure}
 
@@
 The "true" values of the parameters to be inferred were set to $K_{\mbox{\small true}}=50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
-Their initial values in our simulations were set to $K=200$~(arbitrary units of time) and $\gamma = 0.5$.
-
-The results shown here were obtained assuming a simple sinusoidal input $r(t) = \sin^2 \left( 0.01 t \right) + 0.1$.
-A system realization was first obtained from eq.~(\ref{sde}) using $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
-Such system realization was then used to generate a synthetic time series of observed data according to eq.~(\ref{data}).
-The error $\sigma$ was set to $0.1$.
-The input signal, the "true" system realization and the data time series are shown in Fig.~(\ref{fig:rain_data_S})
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.6\textwidth]{Figs/FigRainData.png}
-    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data.
-One may notice how the system response follows the oscillations of the input signal.}
-\end{figure}
-
-We observed a reasonably linear dependence of the total computing time on the number of discretization points $N$.
-In a typical case with $j=30$ (i.e., $N = 301$ discretization points) a complete run with 20000 iterations required about 13 minutes.
-A set of 200 simulated system realizations is shown in Fig.~(\ref{fig:spaghetti}) together with the synthetic data generated as described above.
-In the simulation the masses were set to 720 for the boundary beads, 130 for the staging beads and 150 for both the dimensionless parameters $\beta$ and $\gamma$.
-The different dynamics of the heavy boundary beads and the light staging beads can be appreciated in Fig.~(\ref{fig:spaghetti}).
-
+Their initial values in the simulations were set to $K=200$~(arbitrary units of time) and $\gamma = 0.5$.
+%
 \begin{figure}[htb!]
@@
     \includegraphics[width=0.7\textwidth]{Figs/FigSpaghetti.png}
-    \caption{Simulated system realizations and corresponding synthetic data generated as described in the text.
-In the inset one may appreciate the different dynamics of heavy data-point and light staging beads.}
+    \caption{Simulated system realizations and corresponding synthetic data generated as described in the text with $j=30$ staging beads and therefore $N=301$ discretization points.
+In the inset one may appreciate the different dynamics of heavy data points and light staging beads.}
     \label{fig:spaghetti}
@@
 
+We have assumed a simple sinusoidal input $r(t) = \sin^2 \left( 0.01 t \right) + 0.1$.
+A system realization was first obtained from Eq.~(\ref{sde}) using $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
+Such system realization was then used to generate a synthetic time series of observed data according to Eq.~(\ref{data}).
+The error $\sigma$ was set to $0.1$.
+The input signal, the "true" system realization and the corresponding data time series are shown in Fig.~(\ref{fig:rain_data_S})
+
+Within the investigated range of parameters we observed a reasonably linear dependence of the total computing time on the number of discretization points $N$.
+For example, with $j=30$ (i.e., $N = 301$ discretization points) a complete run with 20000 iterations required about 13 minutes.
+A set of 200 simulated system realizations with $j=30$ is shown in Fig.~(\ref{fig:spaghetti}) together with the synthetic data generated as described above.
+In the simulation the masses were set (in arbitrary units) to 720 for the boundary beads, 130 for the staging beads and 150 for both the dimensionless parameters $\beta$ and $\gamma$.
+The different dynamics of the heavy boundary beads and the light staging beads can be appreciated in Fig.~(\ref{fig:spaghetti}).
+%
 \begin{figure}[htb!]
@@
 \end{figure}
-
+%
 \begin{figure}[htb!]
@@
 
-The Markov chains for 50000 iterations for the parameters $K$ and $\gamma$ are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively.
-Both of them are clearly compatible with the parameter values $K_{\mbox{\small true}} = 50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
-The system evolution in phase space $K-\gamma$ is shown in Fig.~(\ref{fig:phase_space_evol}).
-
+The Markov chains for the parameters $K$ and $\gamma$ generated after 50000 iterations of the HMC algorithm are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively.
+Both of them are clearly compatible with the true parameter values $K_{\mbox{\small true}} = 50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}}=0.2$.
+The efficiency of the algorithm can be best appreciated by inspecting the system evolution in the phase space $K-\gamma$, as shown in Fig.~(\ref{fig:phase_space_evol}).
+The very first step of the algorithm takes already the system in the vicinity of the true parameter values, where most of its dynamics then occur.
+A few excursions away from such values are immediately compensated by the algorithm.
 %\begin{figure}[htb!]
@@
 %\end{figure}
-
+%
 \begin{figure}[htb!]
@@
     \includegraphics[width=0.7\textwidth]{Figs/FigPhaseSpaceEvol.png}
-    \caption{System dynamics in the phase space $K-\gamma$. The empty circle is the initial state, while the square corresponds to the true parameter values used to generate the data.}
+    \caption{System dynamics in the phase space $K-\gamma$. The empty circle represents the initial state, while the square corresponds to the true parameter values used to generate the data.}
     \label{fig:phase_space_evol}
@@
 %    \centering
-%    \includegraphics[width=0.5\textwidth]{Figs/FigKg.png}
-%    \caption{PDF for the parameters $K$ (left) and $\gamma$ (right).
-The true values used to generate the data are $K=200$ and $\gamma = 0.2$.
-The initial values used in the simulations were $K=50$ and $\gamma = 0.4$.}
+%    \includegraphics[width=0.7\textwidth]{Figs/FigK.png}
+%    \caption{Probability density function for the inferred parameter $K$. The true value used to generate the data, $K_{\mbox{\small true}}=50$, is represented by the dotted vertical line.}
 %    \label{fig:K_distr}
 %\end{figure}
+%
+%\begin{figure}[htb!]
+%    \centering
+%    \includegraphics[width=0.65\textwidth]{Figs/FigG.png}
+%    \caption{Probability density function for the inferred parameter $\gamma$. The true value used to generate the data, $K_{\mbox{\small true}}=50$, is represented by the dotted vertical line.}
+%    \label{fig:G_distr}
+%\end{figure}
+The Markov chains of Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}) naturally lead to the probability density functions (PDF) for the two parameters of interest, $K$ and $\gamma$, respectively.
+These functions were calculated using the built-in kernel density estimator provided by Mathematica (version 10) and are shown in Fig.~(\ref{fig:KG_distr}).
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=1.0\textwidth]{Figs/FigKg.png}
+    \caption{Probability density functions for the inferred parameters $K$ (left) and $\gamma$ (right).
+The true values used to generate the data, $K_{\mbox{\small true}}=50$ and $\gamma_{\mbox{\small true}}=0.2$, are represented by the dotted vertical lines.}
+    \label{fig:KG_distr}
+\end{figure}
+
+The structure of the HMC algorithm is well suited to be parallelized.
+We tested the built-in Julia multiprocessing environment to run the program on two separate processes, but we observed a much slower performance compared to the serial implementation.
+This should not be surprising since with the small number of discretization points used so far, not more than a few hundreds, the overheads associated with the communication between the parallel processes significantly exceeded their advantages.
+Nevertheless, the method might be applied to systems requiring a much larger number of points.
+In hydrology, for instance, typical time series corresponding to hourly observations of precipitations and river discharges may contain several tens of thousands of measurements.
+In such cases a parallel version of the program would definitely improve the performance of the method.
+However, it is interesting to observe that Julia might not be the best option for parallelization.
+Indeed, the multiprocessing environment provided by Julia is strongly task oriented, while our problem is fundamentally data oriented.
+In view of a future parallelization of the HMC algorithm it is therefore advisable to investigate also alternative programming languages, such as C++.
+This issue will be discussed in a forthcoming publication.
 

### 2015-08-20 Main_HMC_napa_sinrain.jl

@@ -42,7 +42,7 @@ tdat  = float64(readdlm("$dir/t.dat")[range,2]) # Time points t.dat
 
 const nparams = 2              # Number of system parameters (k, gamma)
 const n = 10                  # n+1 -> "end point" beads (see Tuckerman '93), n -> number of segments
-const j = 30                  # n(j-1) -> total number of staging beads, j-1 -> staging beads per segment
+const j = 10                  # n(j-1) -> total number of staging beads, j-1 -> staging beads per segment
 const N = int64(n*j+1)        # Total number of discrete time points = n*j + 1
                               # IMPORTANT: (N-1)/j = integer = n (measurement points)
 if  ((N-1)%j) != 0 
@@ -55,7 +55,7 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 20000
+const nsample_eff    = 1000
 const nsample        = nsample_eff + nsample_burnin
 
 const dtau           = 0.25                # MD time step

### 2015-08-20 TeXFiles/Manuscript_Albert_Ulzega.tex
FF ref added. ready for Carlo

@@
 Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
-Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref \textbf{FF}).
+Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.
 
@@
     \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
-    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data.
+    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ (in some arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data.
 One may notice how the system response follows the oscillations of the input signal.}

### 2015-08-24 TeXFiles/Manuscript_Albert_Ulzega.tex
draft for internal review

@@
 
-\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models}
+\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for 1D Nonlinear Stochastic Differential Equation Models}
 
@@
 \begin{abstract}
-Bayesian statistics has become an indispensable tool in many applied sciences for the purpose of uncertainty analysis.
-Inferring parametric uncertainty for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
-Here we present a Monte Carlo-based method for Bayesian parameter inference with one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
-%Here, we consider the generic problem of calibrating a one dimensional SDE model to %time series and quantifying the ensuing parametric uncertainty.
-We re-interpret the Bayesian posterior distribution for model parameters as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
-Depending on the number of discretization points and the number of measurement points the dynamics of this system happen on very different time scales.
+Bayesian statistics has become an indispensable tool in many applied sciences, in particular, for the purpose of uncertainty analysis.
+Inferring parametric uncertainty, for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
+%Here we present a Monte Carlo-based method for Bayesian parameter inference with generic one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
+Here, we consider the generic problem of calibrating a one dimensional SDE model to time series and quantifying the ensuing parametric uncertainty.
+We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to sample from it.
+Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
 Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
@@
 \vspace{2pc}
-\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations
+\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations, path-integrals
 
@@
 One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
-Such {\em Approximate Bayes Computations} \cite{Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator.
+Such {\em Approximate Bayes Computations} \cite{marin_2012_ABC, Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator.
 However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.
@@
 Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
-Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987} incorporate the data points for the suggestion of new parameters, whereby much higher acceptance rates are achieved.
+Upon discretizing time of the original problem we are led to the statistical mechanics of a polymer with harmonic bonds in an exterior potential.
+The measurements are interpreted as an additional exterior potential acting on the measurement beads of the polymer only.
+Furthermore, model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer.
+To simulate the dynamics of this system we apply the {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}, which combines Molecular Dynamics (MD) \cite{alder_1959_MD, rahman_1964_MD} with the Metropolis algorithm \cite{metropolis_1953}.
+Compared to filtering methods, we achieve much higher acceptance rates since data points are already used for the suggestion of new parameters, and thus model realizations that are incompatible with the data are never even considered.
 The drawback of these methods is that the model equations need to be known and derivatives have to be calculated.
@@
 Tuning of HMC algorithms is a non-trivial matter.
-Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
+Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the {\em numerical integration} scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
 It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
 Here we explore a different, computationally simper route.
-Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales.
+Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
 Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
-At least for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.
+At least for 1D SDEs we always find a parametrization, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.
 
@@
 
-\section{An Exemplary SDE Model}
+\section{An Exemplary Inference Problem}
 
@@
 However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
-Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
+Consider, for simplicity, a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
 Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
@@
 Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
-Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
-Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.
+Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S(t)$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
+Scale-invariance of eq.
+(\ref{sde}), for large $S$, leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.
 
@@
 Furthermore, by means of the transformation $S(t)=1/n(t)$, eq.
-(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors and extensively studied \cite{dutre_1977_SDE, fujisaka_1986_intermittency}.
+(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{fujisaka_1986_intermittency}.
 
 Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs \cite{tessier_1996} will be studied elsewhere.
-Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with SDE models that are calibrated to observed time-series.
+Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with nonlinear SDE models that are calibrated to observed time-series.
 Here we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
@@
   +
-  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,.
+  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
 \end{equation}
+where the $\epsilon_i$ are uncorrelated standard normal errors.
+
 For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}).
-Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
-In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{\mbox{\small prior}}(\bt)$, while the posterior distribution, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
+Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters, $\bt$, is expressed in terms of probability distributions.
+Assume we have prior knowledge in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, which is believed to be a realization of the model.
+The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of eq.
 \begin{equation}\label{Bayes}
@@
 \end{equation}
-where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution for model outputs given model parameters, evaluated at the measured data.
-Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior to be flat on an area large enough to encompass the area where the likelihood function is significantly different from zero.
-That is, the posterior is data-driven.
+where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution, for model outputs given model parameters, evaluated at the measured data.
+Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior influence to be negligible. That is, we assume the posterior to be data-driven.
 
-Before we set out to derive the likelihood function from the model equations (\ref{sde}) through (\ref{data}) let us express the parameters and state variable with dimensionless quantities.
+Before we set out to derive the likelihood function, from the model equations (\ref{sde}) through (\ref{data}), let us express the parameters and state variable with dimensionless quantities.
 The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
@@
 \end{equation}
-For our algorithm it is important to have the model equation in a form where the noise term does not depend on neither the state variables nor the parameters that have to be inferred.
+For our algorithm it is important to have the model equation in a form where the noise term does neither depend on the state variables nor the parameters that have to be inferred.
 In a one dimensional model this can always be achieved through re-parametrization.
@@
 Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.
-
 We introduce the time-dependent {\em Hamiltonian}
@@
 
-If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as a path-integral as follows
+If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as the path-integral
 \begin{equation}\label{posterior_pathint}
@@
 \end{equation}
-On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$.
-The second term is the probability of the given system realization $q$ and the path-integral extends over all possible system realizations.
+The first term in the exponent describes the log-probability distribution of model outputs, given model parameters, inputs and a system realization $q(t_s)$.
+The second term is the log-probability of a system realization $q(t)$ and the path-integral extends over all possible system realizations.
 Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
@@
   \item
-  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
+  Momenta $p(t)$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
   \item
@@
 The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate.
-Each new phase space configuration corresponds to a well-defined combination of model parameters $\bt$.
-Thus, the simulated succession of configurations represent a Markov chain of sampled parameter vectors that naturally leads to the sought probability distribution for the system parameters.
+Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense.
+Thus, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.
 
-In order to simulate the dynamics of the Hamiltonian (\ref{Hamiltonian}), we first need to discretize the path-integral (\ref{phaseSpacePathInt}).
+In order to simulate the dynamics of Hamiltonian (\ref{Hamiltonian}), we first need to discretize path-integral (\ref{phaseSpacePathInt}).
 Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
@@
 \end{equation}
-and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
+and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ are given by eqs.
 \begin{eqnarray}
@@
    \sum_{i=1}^N
-   \frac{ p_i^2}{2m_q}\Delta t
+   \frac{ p_i^2}{2m_i}\Delta t
    +
@@
 The discretized Hamiltonian describes a {\em classical polymer chain} of $N$ beads with harmonic bonds between neighbouring beads in an external field.
-The latter consists of two parts, a field that results from the dynamics of the original equations (\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.
-The masses $m_q$ and $m_{\alpha}$ are tunable parameters of the algorithm.
-A good strategy consists in assigning larger weights to the beads associated with measurement points than to the intermediate beads.
-The dynamics of the system will be therefore mostly dependent on the lighter intermediate particles, while the heavy measurement beads will be constrained in the vicinity of the measured data points.
+The latter consists of two parts, a field that results from the dynamics of the original Eq.
+(\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.
+The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm.
+Since measurement beads are constrained more than intermediate beads we will assign larger masses to the former.
+Fig. (\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer. 
+The physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced (vertical axis).
+Measurements beads only move within the measurement uncertainty whilst the intermediate beads explore much larger regions of phase space.
+
+%The dynamics of the system will be therefore mostly dependent on the lighter intermediate particles, while the heavy measurement beads will be constrained in the vicinity of the measured data points.
 We thus reduce the original Bayesian inference problem to simulating the dynamics of a linear polymer, as sketched in Fig.~(\ref{fig:polymer}).
-In this context the physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced (vertical axis).
 Each \emph{state} of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$.
@@
 The potential (\ref{Vdisc}) contains terms with different scaling in the potentially large numbers $N$ and $n$, which describe dynamics on different time-scales.
-In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step (ii) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
+In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part and a brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step to sufficiently resolve its dynamics.
 An interesting approximative approach would be to employ a partial averaging of the fast Fourier modes as described in \cite{doll_1985_fourier}.
@@
   \left\{
-    \frac{\Delta t}{m_q}p_{(s-1)j+k}^2
+    \frac{\Delta t}{m'}p_{(s-1)j+k}^2
     +
@@
   \left\{
-   \frac{\Delta t }{m_q}p_{(s-1)j+1}^2
+   \frac{\Delta t }{M}p_{(s-1)j+1}^2
     +
@@
 \end{eqnarray}
+We have introduced two different masses, $M$ and $m'$, for the boundary and staging beads, respectively. 
 The harmonic part for the staging beads given in Eq.~(\ref{H_N}) scales linearly with $N$.
@@
 Finally, Eq.~(\ref{H_1}) does not scale with neither $n$ nor $N$.
-Thanks to the staging variables $\mathcal H_N$ decouples completely from $\mathcal H_n$.
+Thanks to the staging variables, $\mathcal H_N$ decouples completely from $\mathcal H_n$.
 
@@
 \end{enumerate}
-Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is small and/or the measurement error $\sigma$ is large.
+Here, we restrict ourselves to regime ($ii$), i.e., we assume that the number of measurements $n$ is not too large and/or the measurement error $\sigma$ is not too small.
 The generalization of the method to the other schemes is straightforward.
@@
 \begin{equation}
-  m=m_q/\Delta t\,,\quad
+  m=m'/\Delta t\,,\quad
   \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
@@
   +
-  \frac{\Delta\tau}{m} p_{(s-1)j+1}(0)
+  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
   +
-  \frac{\Delta \tau^2}{2m}
+  \frac{\Delta \tau^2}{2M}
   F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
@@
 with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over the time $\tau$.
+After $P$ steps, the proposed configuration, $(\vc u',\bt';\vc p',{\pmb\pi}')$ is accepted with the Metropolis probability
+\begin{equation}
+  \min
+  \left(
+    1,
+    e^{\mathcal{H}_{\mbox{\tiny HMC}}(\vc u,\bt;\vc p,{\pmb\pi}) - \mathcal{H}_{\mbox{\tiny HMC}}(\vc u',\bt';\vc p',{\pmb\pi}')}
+  \right)\,,
+\end{equation}
+and the next iteration starts with sampling an new momentum vector $(\vc p,{\pmb\pi})$.
 
@@
 Our algorithm benefits greatly from the use of AD.
-Indeed, it gives us the possibility to modify the model of Eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered.
+Indeed, it gives us the possibility to modify Eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered.
 This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
@@
 We have run preliminary simulations on a 64-bit laptop computer equipped with a Windows 7 operating system, a double-core 1.6~GHz CPU (Intel i5-4200U) and 8~GB of RAM.
-We have initially used a serial implementation of the algorithm.
+We have used a serial implementation of the algorithm.
 We assumed $n+1 = 11$ measurement points and $j = 10, 20,\dots, 50$ intermediate staging beads, with a total number of discretization points $N = 101, 201, \dots, 501$.
-We have set $\Delta\tau = 0.25$ (arbitrary units of time) and $P = 3$ in the Hamiltonian propagator~(\ref{propagator}), with a constant total observation time $T = 833$ (arbitrary units of time).
-The "true" values of the parameters to be inferred were set to $K_{\mbox{\small true}}=50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
-Their initial values in the simulations were set to $K=200$~(arbitrary units of time) and $\gamma = 0.5$.
+We have set $\Delta\tau = 0.25$ and $P = 3$ in the Hamiltonian propagator~(\ref{propagator}), with a constant total observation time $T = 833$ (arbitrary units of time).
+The "true" values of the parameters to be inferred were set to $K_{\mbox{\small true}}=50$ (arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$.
+Their initial values in the simulations were set to $K=200$ and $\gamma = 0.5$.
 %
@@
 A set of 200 simulated system realizations with $j=30$ is shown in Fig.~(\ref{fig:spaghetti}) together with the synthetic data generated as described above.
-In the simulation the masses were set (in arbitrary units) to 720 for the boundary beads, 130 for the staging beads and 150 for both the dimensionless parameters $\beta$ and $\gamma$.
+In the simulation the masses were set (in arbitrary units) to $M=720$, for the boundary beads, $m=130$, for the staging beads and $m_\alpha=150$, for both the dimensionless parameters $\beta$ and $\gamma$.
 The different dynamics of the heavy boundary beads and the light staging beads can be appreciated in Fig.~(\ref{fig:spaghetti}).
@@
 The Markov chains for the parameters $K$ and $\gamma$ generated after 50000 iterations of the HMC algorithm are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively.
-Both of them are clearly compatible with the true parameter values $K_{\mbox{\small true}} = 50$~(arbitrary units of time) and $\gamma_{\mbox{\small true}}=0.2$.
+Both of them are clearly compatible with the true parameter values $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}}=0.2$.
 The efficiency of the algorithm can be best appreciated by inspecting the system evolution in the phase space $K-\gamma$, as shown in Fig.~(\ref{fig:phase_space_evol}).
-The very first step of the algorithm takes already the system in the vicinity of the true parameter values, where most of its dynamics then occur.
-A few excursions away from such values are immediately compensated by the algorithm.
+The very first step of the algorithm already takes the system to the vicinity of the true parameter values, where most of its dynamics then occurs.
+Few excursions far away from the true parameter values explore the heavy tails of the posterior parameter distribution.
 %\begin{figure}[htb!]
@@
 The structure of the HMC algorithm is well suited to be parallelized. 
-We tested the built-in Julia multiprocessing environment to run the program on two separate processes, but we observed a much slower performance compared to the serial implementation.
+For hydrological applications, where the number of measurement points are often in the tens of thousands, this is an important advantage.
+
+%We tested the built-in Julia multiprocessing environment to run the program on two separate processes, but we observed a much slower performance compared to the serial implementation.
 This should not be surprising since with the small number of discretization points used so far, not more than a few hundreds, the overheads associated with the communication between the parallel processes significantly exceeded their advantages.
@@
 The algorithm is derived from a re-interpretation of the posterior distribution as a partition function of a 1D statistical mechanical system and employs a Hamiltonian Monte Carlo approach combined with a multiple time-scale integration.
-Furthermore, a generic re-parametrization is suggested, which decouples harmonic modes in between measurement points from both the measurement potential and the model parameters, and allows for an efficient analytical integration of these modes.
+Furthermore, a generic re-parametrization is suggested, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.
 The algorithm can be implemented in an efficient and generic fashion using automated differentiation and parallelization.

### 2015-09-16 TeXFiles/Manuscript_Albert_Ulzega.tex
Version submitted to NJP

@@
 Inferring parametric uncertainty, for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
-%Here we present a Monte Carlo-based method for Bayesian parameter inference with generic one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
 Here, we consider the generic problem of calibrating a one dimensional SDE model to time series and quantifying the ensuing parametric uncertainty.
@@
 
-Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
+Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density, for the parameters, is a {\em path-integral}.
 Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data.
@@
 Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
-Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
+Thus, we employ a multiple time scale integration technique, for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
 At least for 1D SDEs we always find a parametrization, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.
@@
 Furthermore, by means of the transformation $S(t)=1/n(t)$, eq.
-(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{fujisaka_1986_intermittency}.
+(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
 
@@
 \end{equation}
-For our algorithm it is important to have the model equation in a form where the noise term does neither depend on the state variables nor the parameters that have to be inferred.
-In a one dimensional model this can always be achieved through re-parametrization.
+A one dimensional model always allows for such a parametrization, in which the noise term does neither depend on the state variables nor the parameters that have to be inferred.
+This will allow us to do part of the integration later on analytically.
 
@@
 
-Properties of (a transformed version of) (\ref{sde}) have been derived in \cite{dutre_1977_SDE, fujisaka_1986_intermittency} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model.
+Properties of (a transformed version of) (\ref{sde}), for constant input, have been derived in \cite{dutre_1977_SDE, schenzle_1979_multStochProc, fujisaka_1986_intermittency} and we do not repeat them here, as the focus of this paper is not on the properties of this particular model.
 However, in order to prove two claims made earlier, we calculate the {\em equilibrium distribution} $P_{\eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$.
@@
 The second term is the log-probability of a system realization $q(t)$ and the path-integral extends over all possible system realizations.
-Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
+In the next section, we apply a Hamiltonian Monte Carlo method to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
 
@@
 and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
-The posterior (\ref{posterior_pathint}) can then be expressed by the phase space path integral
+The posterior (\ref{posterior_pathint}) is rewritten as the phase space path integral
 \begin{equation}\label{phaseSpacePathInt}
@@
   \item
-  Momenta $p(t)$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
+  Momenta $p(t)$ and ${\pmb\pi}$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
   \item
@@
 Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense.
-Thus, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.
+Thus, after cutting off a possible burn-in, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.
 
@@
 \end{equation}
-In order to design a reversible integrator for the associated Hamilton equations, we define the Liouville operators
+In order to design a reversible integrator, for the associated Hamilton equations, we define the Liouville operators
 \begin{equation}
@@
 \end{eqnarray}
-with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over the time $\tau$.
+with $s=1,\dots,n$ and $k=2,\dots,j$. The above propagators are applied sequentially $P$ times to calculate the system evolution over time $\tau$.
 After $P$ steps, the proposed configuration, $(\vc u',\bt';\vc p',{\pmb\pi}')$ is accepted with the Metropolis probability
@@
 \end{equation}
-and the next iteration starts with sampling an new momentum vector $(\vc p,{\pmb\pi})$.
+and the next iteration starts with sampling a new momentum vector $(\vc p,{\pmb\pi})$.
 
@@
 
+\section*{Acknowledgement}
+This work was partly financed by the Eawag Discretionary Fund.
+
 \section*{References}

### 2015-10-11 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
Ruedi's contributions

%\do\gammaumentclass[reqno]{amsart}
\documentclass[11pt]{article}

\usepackage{a4wide}
\usepackage{amsmath,amssymb,amsthm,amscd}
\usepackage[pdftex]{graphicx}

%\usepackage{showkeys}

\usepackage[all]{xy}

\newcommand{\bb}[1]{\begin{equation}\label{#1}}
\newcommand{\ee}{\end{equation}}
\newcommand{\vc}[1]{{\bf #1 }}
\renewcommand{\vec}[1]{{\bf #1}}
\newcommand{\Sum}[1]{\sum\limits_{#1}}
\newcommand{\Int}[1]{\int\limits_{#1}}
\newcommand{\Prod}[1]{\prod\limits_{#1}}
\newcommand{\const}{\operatorname{const}}
\newcommand{\ul}{\underline}
\newcommand{\ol}{\overline}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\bbA}{\mathbb A}
\newcommand{\bbR}{\mathbb R}
\newcommand{\bbN}{\mathbb N}
\newcommand{\bbZ}{\mathbb Z}
\newcommand{\bbC}{\mathbb C}
\newcommand{\bbT}{\mathbb T}
\newcommand{\QB}{\mathcal H}
\newcommand{\PQB}{\tilde{\mathcal H}}
\newcommand{\CQB}{\mathcal H'}
\newcommand{\Ad}{\operatorname{Ad}}
\newcommand{\ad}{\operatorname{ad}}
\newcommand{\gh}{\operatorname{gh}}
\newcommand{\Der}{\operatorname{Der}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\aut}{\operatorname{aut}}
\newcommand{\Diff}{\operatorname{Diff}}
\renewcommand{\dim}{\operatorname{dim}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\Pf}{\operatorname{Pf}}
\newcommand{\Cov}{\operatorname{Cov}}

\newcommand{\g}{{\bf g}}
\newcommand{\h}{{\bf h}}
\renewcommand{\d}{{\bf d}}
\newcommand{\e}{{\bf e}}
\renewcommand{\t}{{\bf t}}
\newcommand{\lpartial}{\overset{\rightarrow}{\partial}}
\newcommand{\rpartial}{\overset{\leftarrow}{\partial}}
\newcommand{\lder}[1]{\overset{\rightarrow}{\vc #1}}
\newcommand{\rder}[1]{\overset{\leftarrow}{\vc #1}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\im}{\operatorname{Im}}
\renewcommand{\ker}{\operatorname{Ker}}
\newcommand{\Var}{\operatorname{Var}}

%define the \identy operator
\def\litID{{\sf id}}
\def\identy{{\mathsurround0pt\mathchoice{\textidenty}{\textidenty}{\scptidenty}{\scptidenty}}}
\def\scptidenty{\setbox0\hbox{$\scriptstyle1$}\bothidenty}
\def\textidenty{\setbox0\hbox{$1$}\bothidenty}
\def\bothidenty{\rlap{\hbox to.97\wd0{\hss\vrule height.06\ht0 width.82\wd0}}
 \copy0\rlap{\kern-.36\wd0\vrule height1.05\ht0 width.05\ht0}\kern.14\wd0}
%end of definition \identy
\newtheorem{theorem}{Theorem}[section]

\newtheorem{corollary}[theorem]{Corollary}

\newtheorem{definition}[theorem]{Definition}

\newtheorem{lemma}[theorem]{Lemma}

\newtheorem{lemmadefinition}[theorem]{Lemma and Definition}

\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}

\newtheorem{example}[theorem]{Example}

\newtheorem{remark}[theorem]{Remark}

\newcommand{\bbM}{\mathbb M}
\newcommand{\bt}{\pmb\theta}
\newcommand{\Bt}{\pmb\Theta}

\begin{document}

\title{Boosting Bayesian Parameter Inference of Nonlinear Stochastic Differential Equation Models by  Hamilton Equation Scale Separation }

\author{Carlo Albert\footnote{Eawag, Swiss Federal Institute of Aquatic Science and Technology, 8600 D\"ubendorf, Switzerland}, Simone Ulzega\footnotemark[1], Ruedi Stoop\footnote{Institute of Neuroinformatics and Institute of Computational Science UZH/ETHZ, Irchel Campus 8057 Zurich, Switzerland}}

\maketitle

\begin{abstract}
For the modeling of dynamical systems characterized by time-series, we propose a novel very efficient approach for generating posterior parameter distributions. The algorithm is inspired by re-interpreting the posterior distribution as a partition function of a 
%1D 
statistical mechanics system. To arrive at distribution samples, we employ a Hamiltonian Monte Carlo approach combined with multiple time-scale integration.  
At least for 1D problems, generic re-parametrization allows us to decouple the harmonic modes between measurement points and solve their dynamics analytically. 
%Generic re-parametrization helps us to decouple the harmonic modes between measurement points, which leads to an analytical solution of the node dynamics. 
Our approach is applicable to a wide range of inference problems and is highly parallelizable.

\end{abstract}

\section{Introduction}

Modeling a dynamical process starts with a basic model of the process, which usually is obtained from more or less deep insight into the nature of the process. The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy. 
A minimal example is a perceptron [XXX], the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time. While this task can easily be achieved for clean data using gradient descent learning, for noisy input data, this is largely impossible, as noise cannot be learned. The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
%\begin{figure}[h]
%    \centering
%    \includegraphics[width=1.0\textwidth]{neuron.pdf}
%    \caption{General problem setting: Parameter estimation (synaptic weights $w_1$,$w_2$) of a perceptron, from a) noiseless, b) noisy data (noise sampled from a flat distribution over the  interval $[-0.2,0.2]$). Whereas for noiseless data the estimates perfectly converge, for noisy data the estimates  the system attempts to also include the noise, leading to a nontrivial distribution of the parameter estimates, and rendering the extraction of the optimal parameters a nontrivial task. Red circles: location of the optimal parameters in the noiseless case.}
%    \label{neural}
%\end{figure}
%If the physics underlying the evolution of the system follows a differential equation,
%the search for the optimal parameters and their typical distribution is called Bayesian inference of a stochastic differential equation (SDE).

In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions [REF]. 
If a constant noise term is added to the output of a (reasonably fast) deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
If noise enters the formulation of the model equations, however, Bayesian inference quickly becomes computationally very expensive due to the high dimensional integrals that have to be evaluated.

In our paper, we demonstrate the calibration of ordinary 1D stochastic differential equation (SDE) models based on noisy time series, and the quantification of the resulting parametric uncertainty. The generic approach that we use is exemplified by a simple SDE model from hydrology.

%Bayesian parameter inference for SDE models is computationally very expensive, as the posterior parameter probability density is a path-integral. 
Problems of this kind are commonly solved by Monte Carlo (MC) methods that are based on simulating model realizations and comparing them to the data. Popular methods are particle filters \cite{chopin_2013_SMC2}, Metropolis-within-Gibbs algorithms \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters} or Approximate Bayes Computations (ABC) \cite{marin_2012_ABC, Albert_2014_ABC}. 
A major problem with these simulation-based methods is, however, their inefficiency in the presence of many data points or high dimensions. One solution is to map the output space to a smaller dimensional space of summary statistics, and accept/reject proposed model parameters depending on how well associated model runs conform with the data in terms of these summary statistics [Ref.]. However, how to choose the summary statistics to achieve a significant representation of the posterior parameter distribution is a largely unsolved problem.

%In our approach we remedy these difficulties. 
%We start by re-interpreting the posterior distribution as the partition function of a statistical mechanical system, of which we simulate the time evolution.
These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
After discretizing the time of the original problem, we are led to the statistical mechanics of a polymer %-like problem 
with harmonic bonds in an exterior potential. In this framework, the measurements are interpreted as an additional exterior potential that acts only on the polymer's `measurement beads', and model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer. 
To simulate the dynamics of this system, we apply the Hamiltonian Monte Carlo (HMC) algorithm \cite{duane_1987}, which combines Molecular Dynamics (MD) \cite{alder_1959_MD, rahman_1964_MD} with the Metropolis algorithm  \cite{metropolis_1953}. 
Compared to traditional methods, the use of the Hamiltonian approach achieves much higher acceptance rates since data points are already used for the suggestion of new parameters, and thus model realizations incompatible with the data are never considered. The drawback is that the model equations need to be known and derivatives have to be calculated.

HMC requires two sets of parameters to be tuned: (i) the parameters that define the kinetic energy of the statistical mechanical system and (ii) those parameters that define the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm. Efficiency of HMC algorithms can be gained if the kinetic term is made dependent on the configuration geometry of the statistical mechanical system. If the Riemann geometry of the parameter space of statistical models is taken into account, then the simulated search of paths across this manifold samples the target density in an utmost efficient way
    \cite{girolami_2011_HMC}. Unfortunately, this procedure is both demanding and computationally costly,  depending strongly on the quality of the space's extracted geometry.

%Riemann geometry of the parameter space of statistical models and thus automatically adapts to the local structure when simulating paths across this manifold, providing highly efficient convergence and exploration of the target density

Here we 
%therefore 
explore a computationally simpler approach, which, to our knowledge, has never been applied in the context of Bayesian inference before. 
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales. This suggests a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}. 
We will show that for 1D SDEs we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics. 
Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series. 
In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points. 
%Our approach can be combined with the approach in \cite{girolami_2011_HMC}.

\section{Inference problem setting}

Consider, for simplicity and concreteness, a reservoir dynamics S(t) that on the observation time-scale is linear, with other (inflow and outflow) processes happening at much shorter time scales, so that they can be described by white noise. Furthermore, assume that this noise scales linearly with the system state S(t). The model equation is thus given by the SDE

\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input, $\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
For constant input $r(t) = r_0$ and in the long-time limit, the mean of $S(t)$ converges to the equilibrium solution of the unperturbed $(\gamma = 0)$ system, $S_{eq} = Kr_{0}$ (cf. end of section).

%Scale-invariance of eq. (\ref{sde}), for large $S$, leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.

While this setting was motivated by a popular hydrological model \cite{breinholt_2011_SDE}, it is by no means restricted to this context. By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
In this setting, the input $r(t)$ is a smooth and nowhere vanishing function. We assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors,

\begin{equation}\label{data}
  \ln \left( y_s \right)
  =
  \ln \left( \frac{S(t_s)}{K} \right)
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where the $\epsilon_i$ are uncorrelated standard normal errors.
For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters, $\bt$, is expressed in terms of probability distributions.
We start our treatise by assuming that we have prior knowledge in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of 
\begin{equation}\label{Bayes}
  f_{\mbox{\small post}}(\bt|\vc y)
  =
  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the probability distribution for model outputs given model parameters, evaluated at the measured data (the so-called {\em likelihood function}).
We posit that parameters must be non-negative; otherwise we assume no additional prior knowledge (i.e., the posterior is ‘data-driven’ [XX]).

Before we set out to derive from Eqs. (\ref{sde}) - (\ref{data}) the likelihood function, we express the parameters and state variables by dimensionless quantities.
Due to scale-invariance of the noise term, $\gamma$ is already dimensionless. State variable $S(t)$ and parameter $K$ are made dimensionless by means of the transformations $K=T\gamma/\beta^2$ and $S(t)=(T\gamma r(t)/\beta^2)e^{\beta q(t)}$.
%\beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
%  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
%\end{equation}
In these new variables and parameters, Eq.~(\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with $\rho(t)  =  (T/\beta)(\mbox d/\mbox{dt})\left[ \ln \left( r(t) \right) \right]
  +
  (2+\gamma)\beta/(2\gamma)$.
%\begin{equation}
 % \rho(t)
  %=
  %\frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
 % +
 % \frac{(2+\gamma)\beta}{2\gamma}\,.
%\end{equation}
Scale-invariance of Eq.~(\ref{sde}) leads to power law tails in the probability distribution of $S(t)$. In real-world hydrology, for which Eq.~(\ref{sde}) is a model, indeed often fat-tailed error distributions are observed ~\cite{thyer_2009_fattails}.

Because this will permit to do part of the integration to be performed later analytically, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred. In a one dimensional model this can always be achieved through re-parametrization [XXX]. The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ if it was in an initial state $q_0$ at time $t = 0$, is expressed as a {\em path-integral} as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$ and where the path-measure $\mathcal Dq$ is formally written as the infinite product
${\mathcal Dq}=\prod_{t}dq(t)$.
The {\em action} is a functional on the space of paths and reads  \cite{lau_2007}
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
This action includes the Jacobian that is introduced when changing coordinates from
 ${\eta(t)}$ to $q(t)$.
We introduce now the time-dependent {\em Hamiltonian}
$  \mathcal{H}(q,t)= \frac{1}{\gamma}e^{-\beta q}+q\rho(t)$
and rewrite action~(\ref{action1}) as
\begin{multline}\label{action}
{\mathcal S}[{q},\dot q]
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}
Properties of a transformed version of (\ref{sde}) have been derived in \cite{dutre_1977_SDE, schenzle_1979_multStochProc, fujisaka_1986_intermittency}, for constant input. 
To prove two claims made earlier, we calculate the equilibrium distribution $P_{eq}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant input $r(t)\equiv r_{0}$. After plugging (\ref{pathint}) and (\ref{action}) into the detailed balance condition
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{eq}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{eq}(q_1) \,,
\end{equation}
and using the transformation $q(t) \rightarrow q(-t)$, we get, since $\dot\rho(t)= 0$,
$P_{eq}(q)
  \propto
  e^{-2\mathcal{H}(q)}$.
Transforming back to the original variables, it turns out that $P_{eq}(S)$ is an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.
\begin{equation}\label{inverse_gamma}
  P_{eq}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,.
\end{equation}
The mean of this expression equals the equilibrium solution of the unperturbed system ($\gamma=0$)
$\langle S\rangle_{eq}=Kr_{0}$
and its variance for $\gamma< 2$ is given by $ \langle (S - \langle S\rangle_{eq})^2\rangle_{eq}
  =
  K^2r_{0}^2
  \gamma/(2-\gamma)$, which for $\gamma\geq 2$ is seen to diverge.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model. If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior, the posterior (\ref{Bayes}) is, as a function of $\bt$, proportional to the likelihood function 
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Whereas the first term in the exponent describes the log-probability distribution of model outputs, for given model parameters, inputs and a system realization $q(t_s)$, the second term is the log-probability of a system realization $q(t)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Inference algorithm}

Instead of undertaking an often prohibitive numerical computation of the path integral, we now apply HMC to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
For efficiently drawing parameter samples from (\ref{posterior_pathint}), we interpret the latter as the partition function of a 1D statistical mechanical system and simulate its dynamics employing the HMC algorithm \cite{duane_1987}. (This is always possible and does not depend on the nature of the SDE we consider.) The model parameters  $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$. Each degree of freedom, $q(t)$ and $\bt$, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$ respectively, so that the system is defined by the Hamiltonian

\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
where
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m(t)}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
and $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) can then
be expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  e^{-\mathcal {H}_{\mbox{\tiny HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
   {\mathcal Dq}
   d{\pmb\pi}
  \,.
\end{equation}

The HMC method, as a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p(t)$ and ${\pmb\pi}$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretized set of Hamilton equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps permit arbitrarily large jumps in phase space, while maintaining an arbitrarily large acceptance rate. Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense. Thus, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.

In order to simulate the dynamics of the Hamiltonian (\ref{Hamiltonian}), we first need to discretize the path-integral (\ref{phaseSpacePathInt}).
Let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
The path-integral (\ref{phaseSpacePathInt}) is then approximated by an ordinary integral, with the approximate path-measure $  \mathcal Dp\mathcal Dq  \approx  \prod_i dp_i dq_i$.
The discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ are now
\begin{align}
   K( p,{\pmb\pi}) &\approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_i}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
\end{align}
\begin{align} \label{Vdisc}
  V(q,\bt)  &\approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\} 
  \\ \nonumber 
  &+
  \frac{1}{\gamma} e^{-\beta q_N} + q_N \rho_{N} - \frac{1}{\gamma} e^{-\beta q_1} -  q_1 \rho_{2} +
  \sum_{s=1}^{n+1} \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}\,,
\end{align}
with $\dot q_i = (q_i-q_{i-1})/\Delta t$, $\rho_i = T \ln(r(t_{i})/r(t_{i-1}))/(\beta\Delta t)+(2+\gamma)\beta/(2\gamma)$ and $\dot\rho_i = (\rho_i-\rho_{i-1})/\Delta t$, and where terms of order $\mathcal O(N^{-1/2})$ were neglected.

Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field. The latter consists of two parts, a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only. The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm. Since measurement beads are constrained more than intermediate beads, we will assign larger masses to the former. Fig. (\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer. The physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced as the vertical axis. Measurement beads only move within the measurement uncertainty, whilst the intermediate beads explore much larger regions of phase space.

%The dynamics of the system will be therefore mostly dependent on the lighter intermediate particles, while the heavy measurement beads will be constrained in the vicinity of the measured data points.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{FigPolymer2.pdf}
    \caption{Simulated polymer chain dynamics, with $n+1=11$ data points (large circles) and $j-1=9$ intermediate beads (small circles). For other parameters see Section~\ref{numerical_results}. Bottom: initial state, where intermediate beads are on a linear interpolation between data points. Top: polymer after $T=1000$ iterations of the propagation algorithm (dotted line: initial configuration). Clearly the new configuration is mostly determined by the dynamics of the light-mass intermediate beads, while the heavy-mass data points move to a much lesser extent.}
    \label{fig:polymer}
\end{figure}

We have thus reduced the original Bayesian inference problem to simulating the dynamics of a linear polymer (cf. Fig.~(\ref{fig:polymer})). Each state of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$. It is now essential to note that potential (\ref{Vdisc}) contains terms of distinct scaling in the potentially large numbers $N$ and $n$, that refer to dynamics on distinct time-scales. In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and for resolving its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.

Whereas an interesting {\em approximative} approach would have been to employ a partial averaging of the fast Fourier modes \cite{doll_1985_fourier}, we will use an {\em exact multiple time scale integration} based on Trotter's formula \cite{tuckerman_1993}. For this, we introduce so-called {\it staging variables}, and diagonalize the harmonic part between the measurement points. To this end, we rewrite the discretized harmonic part of the action as
\begin{equation}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2
  =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{equation}
with $  q^*_{(s-1)j+k}  =  ((k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} )/k$.
For the boundary beads, corresponding to the original measurement points, we apply the coordinate transformations
$  u_{sj+1} = q_{sj+1}$, $s=0,\dots,n$
and for the intermediate staging beads
$u_{sj+k} = q_{sj+k} - q^*_{sj+k}$, $s=0,\dots,n-1$, $k=2,\dots,j$. 
Their inverse transformations are given by
\begin{align}
  q_{sj+1} &= u_{sj+1}\,,
  \\
  q_{sj+k} &= \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{align}
which can be captured by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{sj+k+1}+ \frac{1}{k}u_{sj+1} \,.
\end{equation}
We now split the Hamiltonian $\mathcal{H}_{\mbox{\tiny HMC}}$ into components according to their scaling behavior in $n$ and $N$, and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}= \mathcal H_N+\mathcal H_n+\mathcal H_1\,,
\end{equation}
where
\begin{align}
  \mathcal H_N &=
  \frac{1}{2}
  \sum_{s=1}^{n}
  \sum_{k=2}^j
  \left\{
    \frac{\Delta t}{m'}p_{(s-1)j+k}^2
    +
    \frac{Tk}{\Delta t(k-1)}
    u_{(s-1)j+k}^2
  \right\}\,,\label{H_N}
  \\\label{H_n}
  \mathcal H_n &=
  \frac{1}{2}
  \sum_{s=1}^{n+1}
  \left\{
   \frac{\Delta t }{M}p_{(s-1)j+1}^2
    +
    \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta u_{(s-1)j+1}})^2}{\sigma^2}
   \right\}
   \\\nonumber
  &+
  \frac{T}{2j\Delta t}
  \sum_{s=1}^{n}
    (u_{(s-1)j+1} - u_{sj+1})^2
   \,,
  \\ \label{H_1}
  \mathcal H_1 &=
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}
   +
  \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}
  \\\nonumber
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{2} \,.
\end{align}

Here, we have introduced two masses, $M$ and $m'$, for the boundary and staging beads, respectively.
For the staging beads, the harmonic part of Eq.~(\ref{H_N}) scales linearly with $N$. The terms of Eq.~(\ref{H_n}), including both the harmonic part for the boundary beads and the measurement term, scale linearly with $n$. Finally, Eq.~(\ref{H_1}) neither scales with $n$ nor $N$.
Thanks to the staging variables, $\mathcal H_N$ and $\mathcal H_n$ have become fully decoupled.
We use Trotter's formula \cite{tuckerman_1992} in order to design a reversible molecular dynamics integrator that takes the presence of the different time scales into account.
For an appropriate partition of the Hamiltonian, three distinct regimes can be distinguished:

 \begin{enumerate}
   \item[\it (i)]
  $\mathcal H_N \sim \mathcal H_n >> \mathcal H_1$,
  \item[\it (ii)]
  $\mathcal H_N >> \mathcal H_n \sim \mathcal H_1$,
  \item[\it (iii)]
  $\mathcal H_N >> \mathcal H_n >> \mathcal H_1$.
\end{enumerate}

In the following we restrict ourselves to regime $(ii)$, where the number of measurements $n$ is assumed to be not too large and/or the measurement error $\sigma$ to be not too small (the generalization of the method to the other regimes would, however, be straightforward). In this regime we may simply separate the harmonic part of the action for the staging beads from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
For obtaining reversible integrators, we define the Liouville operators
$ iL_N=\{\cdot\,,\,\mathcal H_N\}$, $iL'=\{\cdot\,,\,\mathcal H'\}$,
where $\{\cdot\,,\,\cdot\}$ denote the Poisson brackets that apply to functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
Here, the outer propagator $\exp[iL_N(\Delta \tau/2)]$ reflects much faster dynamics than the inner one. 
However, thanks to our re-parametrization, it describes the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m'/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The outer propagator becomes
\begin{align}
  u_{(s-1)j+k}(\Delta\tau/2)
  &=
  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,
  \\
  p_{(s-1)j+k}(\Delta\tau/2)
  &=
  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{align}
for $s=1,\dots,n$ and $k=2,\dots,j$. 
This elegant `analytical' solution is one main boosting part of our algorithm. 
For the inner propagator, we employ the time-reversible and symplecticity preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
\begin{align}
  u_{(s-1)j+1}(\Delta\tau)
  &= u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2M}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau)
  &= p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{align}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
Analogous equations emerge for the model parameters $\bt$ and their momenta ${\pmb\pi}$, by exchanging $u_{…}$ by $\theta_{…}$ and $p_{…}$ by $\pi_{…}$, along with the corresponding masses, respectively. 
For the staging beads
%, as yet another benefit of the approach, 
only the momenta need to be updated (because the associated kinetic term is not part of $\mathcal H'$, but of $\mathcal H_N$).
Thus, with $s=1,\dots,n$ and $k=2,\dots,j$ as before,
\begin{align}
  p_{(s-1)j+k}(\Delta\tau)
  &=
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,.
\end{align}

The propagators (28)-(32) are applied sequentially $P$ times to calculate the system evolution over time $\tau$. The proposed configuration, $(\vc u',\bt';\vc p',{\pmb\pi}')$ is accepted with Metropolis probability $  \min
  \left(
    1,
    e^{\mathcal{H}_{\mbox{\tiny HMC}}(\vc u,\bt;\vc p,{\pmb\pi}) - \mathcal{H}_{\mbox{\tiny HMC}}(\vc u',\bt';\vc p',{\pmb\pi}')}
  \right)$.
The next iteration then starts with sampling a new momentum vector $(\vc p,{\pmb\pi})$.

\section{Results}\label{numerical_results}

For our toy system, we have considered a simple sinusoidal input  $r(t) = \sin^2 \left( 0.01 t \right) + 0.1$. A system realization was first obtained from Eq.~(\ref{sde}) using $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$. 
Such system realization was then used to generate a synthetic time series of observed data according to Eq.~(\ref{data}). The error $\sigma$ was set to $0.1$. The input signal, the "true" system realization and the corresponding data time series are shown in Fig.~(\ref{fig:rain_data_S}). 
%Since we are working with rescaled equations, all results are shown in arbitrary units of time.
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Fig2.pdf}
    \caption{\label{fig:rain_data_S} 
    System realization (solid line) with synthetic observations (filled circles, with error bars indicating the assumed measurement uncertainty).
    The system response closely follows time-delayed the oscillations of the sinusoidal input (dashed).
    Parameters for this figure and all following figures: $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$.
%    Measured system realization (filled circles) and synthetic observed data (solid line and deviation bars attached to the circles evaluated across a longer time interval.). The system response closely follows time-delayed the oscillations of the sinusoidal input (dashed), with larger error bars where the prediction is less faithful. Parameters for this figure and all following figures: $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$.
}
\end{figure}
Within the investigated range of parameters, we observed a reasonably linear dependence of the total computing time on the number of discretization points $N$. For example, a complete run with 20000 iterations with $j=30$ (i.e., $N = 301$ discretization points) required about 13 minutes (for implementation details see the next section). A set of 200 simulated system realizations based on $j=30$ is shown in Fig.~(\ref{fig:spaghetti}), together with the generated synthetic data. In the simulation, the masses were set (in arbitrary units) to $M=720$ for the boundary beads, to $m=130$ for the staging beads, and to $m_\alpha=150$ for both the dimensionless parameters $\beta$ and $\gamma$. The different dynamics of the heavy boundary beads and the light staging beads can be appreciated in Fig.~(\ref{fig:spaghetti}).
%
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Fig3.pdf}
    \caption{Simulated system realizations associated with synthetic data, based on $n+1 = 11$ measurement points, $j=30$ staging beads and  $N=301$ discretization points. 
    %In the inset one may appreciate the different dynamics of heavy data points vs. the lighter staging beads. 
}
    \label{fig:spaghetti}
\end{figure}
The Markov chains for parameters $K$ and $\gamma$, obtained after 50000 iterations of the HMC algorithm, are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively; they are fully compatible with the true, to be inferred, parameter values $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}}=0.2$.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Fig4.pdf}
    \caption{Markov chain evolution of the inferred parameter $K$.}
    \label{fig:chainK}
\end{figure}
%
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Fig5.pdf}
    \caption{Markov chain evolution of the inferred parameter $\gamma$.}
    \label{fig:chainG}
\end{figure}

%\begin{figure}[htb!]
%    \centering
%    \includegraphics[width=0.3\textwidth]{FigFinalState.png}
%    \caption{PDF for the system final state expressed via the coordinate set $\left\{ q_N \right\}$. The dashed curve represents a normal distribution with the same mean and variance. The final state is basically normally distributed.}
%    \label{fig:final_state}
%\end{figure}
%
The efficiency of the algorithm can be appreciated best by inspecting the system evolution in the phase space $K-\gamma$ (Fig.~(\ref{fig:phase_space_evol})). The very first step of the algorithm already takes the system to the vicinity of the true parameter values, where most of its dynamics then occurs. Few excursions lead far away from the true parameter values. These explore the heavy tails of the posterior parameter distribution.
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Fig6.pdf}
    \caption{System dynamics in the phase space $K-\gamma$. The green circle represents the initial state, while the red square corresponds to the true parameter values used to generate the data.}
    \label{fig:phase_space_evol}
\end{figure}

The Markov chains (Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG})) determine the probability density functions (PDF) for $K$ and $\gamma$, respectively. The result obtained by using the built-in kernel density estimator provided by Mathematica (version 10) is exhibited in  Fig.~(\ref{fig:KG_distr}).

\begin{figure}[htb!]
    \centering
    \includegraphics[width=1.0\textwidth]{Fig77.pdf}
    \caption{Probability density functions for the inferred parameters, (a) $K$ and  (b) $\gamma$. The true values used to generate the data are represented by the dashed vertical lines.}
    \label{fig:KG_distr}
\end{figure}

\section{Physical computational implementation}

The algorithm was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9, http:\/\/julialang.org\/). One of the attractive features of this language is its large and dynamic library for numerical and scientific applications.
From this, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions. Our algorithm benefits greatly from the use of AD. This gives us the possibility to modify Eq.~(\ref{sde}) and therefore the action~(\ref{action}), while leaving the implementation of the algorithm unaltered. This makes our program extremely flexible and suitable for a much broader range of applications than the simple exemplary SDE model described here.
%

The simulations were run on a serial implementation of the algorithm on a 64-bit laptop computer equipped with a Windows 7 operating system, a double-core 1.6~GHz CPU (Intel i5-4200U) and 8~GB of RAM.  $n+1 = 11$ measurement points and $j = 10, 20,\dots, 50$ intermediate staging beads, with a total number of discretization points $N = 101, 201, \dots, 501$ were taken. In the Hamiltonian propagator~(\ref{propagator}), we set $\Delta\tau = 0.25$ and $P = 3$, with a constant total observation time $T = 833$ (arbitrary units of time).  Their initial values in the simulations were set to $K=200$ and $\gamma = 0.5$.
%

\section{Conclusions}

We presented a novel, extremely efficient and versatile approach for SDE parameter estimation, based on observed data.
Our algorithm derives its strenght from a translation of the problem of generating posterior parameter samples to the problem of simulating the dynamics of a statistical mechanical system.
The main efficiency gain of our algorithm, compared to the one suggested in Ref. \cite{girolami_2011_HMC}, stems from an exploitation of the fact that this dynamics generically happens on very different time scales. 
Furthermore, at least for 1D systems, our approach also allows for an analytical, and therefore computationally efficient, integration of the fastest part of the dynamics. 
The algorithm suggested in Ref. \cite{girolami_2011_HMC} locally adapts the mass matrix to the curvature of the potential, whereas we use a constant diagonal mass matrix, with reasonable choices of masses, heavier for the measurement beads, lighter for the discretization (staging) beads.
Whilst the former approach leads to some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved, it is likely to be beneficial in cases with strongly varying curvature. 
Our approach, however, is to a large extent complementary to and could be combined with the one in \cite{girolami_2011_HMC}.

%We presented a novel, extremely efficient and versatile approach for SDE parameter estimation, based on observed data. Compared to Ref. \cite{girolami_2011_HMC} that proposed an approach designed to remedy similar shortcomings of the traditional approaches, our approach is conceptually and implementation-wise much simpler and faster. Whereas for the former algorithm, locally the solution space metric needs to be evaluated, we only need to make reasonable choices of weights, heavier for the measurement beads, light for the staging beads. Although these choices are uncritical, they could be adapted to and optimized for the specific problem, if necessary. A particular advantage is that a large number of the updates can be done semi-analytically and do not involve the numerical calculation of derivatives. This saves a considerable computational effort. 
%Our approach is much simpler to implement and to monitor, and should, according to the arguments provided, generally be faster.

The structure of our algorithm is well suited to parallelization, which is important in particular if we deal with a high number of measurements (e.g., some tens of thousands data points). The algorithm can easily be adapted to other inference problems, such as higher dimensional SDE, and SDE coupled to ODE. These adaptations and extensions will, however, be addressed in future works.

%\begin{figure}[htb!]
%    \centering
%    \includegraphics[width=0.7\textwidth]{FigK.png}
%    \caption{Probability density function for the inferred parameter $K$. The true value used to generate the data, $K_{\mbox{\small true}}=50$, is represented by the dotted vertical line.}
%    \label{fig:K_distr}
%\end{figure}
%
%\begin{figure}[htb!]
%    \centering
%    \includegraphics[width=0.65\textwidth]{FigG.png}
%    \caption{Probability density function for the inferred parameter $\gamma$. The true value used to generate the data, $K_{\mbox{\small true}}=50$, is represented by the dotted vertical line.}
%    \label{fig:G_distr}
%\end{figure}

%We tested the built-in Julia multiprocessing environment to run the program on two separate processes, but we observed a much slower performance compared to the serial implementation. This should not be surprising since with the small number of discretization points used so far, not more than a few hundreds, the overheads associated with the communication between the parallel processes significantly exceeded their advantages. Nevertheless, the method might be applied to systems requiring a much larger number of points. In hydrology, for instance, typical time series corresponding to hourly observations of precipitations and river discharges may contain several tens of thousands of measurements. In such cases a parallel version of the program would definitely improve the performance of the method. However, it is interesting to observe that Julia might not be the best option for parallelization. Indeed, the multiprocessing environment provided by Julia is strongly task oriented, while our problem is fundamentally data oriented. In view of a future parallelization of the HMC algorithm it is therefore advisable to investigate also alternative programming languages, such as C++. This issue will be discussed in a forthcoming publication.

\section*{Acknowledgement}
This work was partly financed by the Eawag Discretionary Fund.

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2015-10-22 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
minor changes in tex file

@@
 The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
-%\begin{figure}[h]
-%    \centering
-%    \includegraphics[width=1.0\textwidth]{neuron.pdf}
-%    \caption{General problem setting: Parameter estimation (synaptic weights $w_1$,$w_2$) of a perceptron, from a) noiseless, b) noisy data (noise sampled from a flat distribution over the  interval $[-0.2,0.2]$).
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=1.0\textwidth]{neuron.pdf}
+    \caption{General problem setting: Parameter estimation (synaptic weights $w_1$,$w_2$) of a perceptron, from a) noiseless, b) noisy data (noise sampled from a flat distribution over the  interval $[-0.2,0.2]$).
 Whereas for noiseless data the estimates perfectly converge, for noisy data the estimates  the system attempts to also include the noise, leading to a nontrivial distribution of the parameter estimates, and rendering the extraction of the optimal parameters a nontrivial task.
 Red circles: location of the optimal parameters in the noiseless case.}
-%    \label{neural}
-%\end{figure}
-%If the physics underlying the evolution of the system follows a differential equation,
-%the search for the optimal parameters and their typical distribution is called Bayesian inference of a stochastic differential equation (SDE).
+    \label{neural}
+\end{figure}
+If the physics underlying the evolution of the system follows a differential equation,
+the search for the optimal parameters and their typical distribution is called Bayesian inference of a stochastic differential equation (SDE).
 
@@
 Bottom: initial state, where intermediate beads are on a linear interpolation between data points.
-Top: polymer after $T=1000$ iterations of the propagation algorithm (dotted line: initial configuration).
+Top: polymer after $N_{MC}=1000$ iterations of the propagation algorithm (dotted line: initial configuration).
 Clearly the new configuration is mostly determined by the dynamics of the light-mass intermediate beads, while the heavy-mass data points move to a much lesser extent.}

### 2015-10-23 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
improvement of tex file, new citations added

@@
 
-\title{Boosting Bayesian Parameter Inference of Nonlinear Stochastic Differential Equation Models by  Hamilton Equation Scale Separation }
+\title{Boosting Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models by Hamiltonian Scale Separation}
 
@@
 \begin{abstract}
-For the modeling of dynamical systems characterized by time-series, we propose a novel very efficient approach for generating posterior parameter distributions.
+We propose a novel very efficient approach for generating posterior parameter distributions, for the modeling of dynamical systems characterized by time-series.
 The algorithm is inspired by re-interpreting the posterior distribution as a partition function of a
@@
 
-Modeling a dynamical process starts with a basic model of the process, which usually is obtained from more or less deep insight into the nature of the process.
+Modeling a dynamical process starts with a basic model that is usually obtained from more or less deep insight into the nature of the process.
 The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy.
@@
 Whereas for noiseless data the estimates perfectly converge, for noisy data the estimates  the system attempts to also include the noise, leading to a nontrivial distribution of the parameter estimates, and rendering the extraction of the optimal parameters a nontrivial task.
-Red circles: location of the optimal parameters in the noiseless case.}
+Open circles: location of the optimal parameters in the noiseless case.}
     \label{neural}
 \end{figure}
-If the physics underlying the evolution of the system follows a differential equation,
-the search for the optimal parameters and their typical distribution is called Bayesian inference of a stochastic differential equation (SDE).
 
 In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions [REF].
-If a constant noise term is added to the output of a (reasonably fast) deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
-If noise enters the formulation of the model equations, however, Bayesian inference quickly becomes computationally very expensive due to the high dimensional integrals that have to be evaluated.
+If a constant noise term is added to the output of a deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
+If noise enters the formulation of the model equations, however, Bayesian inference all of a sudden becomes computationally very expensive.
 
@@
 %We start by re-interpreting the posterior distribution as the partition function of a statistical mechanical system, of which we simulate the time evolution.
-These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
+These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanics system and by simulating the dynamics of the latter.
 After discretizing the time of the original problem, we are led to the statistical mechanics of a polymer %-like problem
@@
 
-HMC requires two sets of parameters to be tuned: (i) the parameters that define the kinetic energy of the statistical mechanical system and (ii) those parameters that define the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
-Efficiency of HMC algorithms can be gained if the kinetic term is made dependent on the configuration geometry of the statistical mechanical system.
+HMC requires two sets of parameters to be tuned: (i) the parameters that define the kinetic energy of the statistical mechanics system and (ii) the parameters that define the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
+Efficiency of HMC algorithms can be gained if the kinetic term is made dependent on the configuration geometry of the statistical mechanics system.
 If the Riemann geometry of the parameter space of statistical models is taken into account, then the simulated search of paths across this manifold samples the target density in an utmost efficient way
@@
 explore a computationally simpler approach, which, to our knowledge, has never been applied in the context of Bayesian inference before.
-Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales.
-This suggests a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
+Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanics system happen on very different time scales.
+This suggests a multiple time scale integration technique for the simulation of the statistical mechanics system \cite{tuckerman_1993}.
 We will show that for 1D SDEs we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics.
@@
 \end{equation}
-where $r(t)$ denotes the time varying rain input, $\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
+where $r(t)$ denotes the time varying rain input, $K$ denotes the retention time,$\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
 \begin{equation}\label{whitenoise}
@@
 
-While this setting was motivated by a popular hydrological model \cite{breinholt_2011_SDE}, it is by no means restricted to this context.
-By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
+While this setting was motivated by a popular hydrological model \cite{breinholt_2011_SDE, livina_2003_dischargeFluctuations}, it is by no means restricted to this context.
+By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested, e.g., as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
 In this setting, the input $r(t)$ is a smooth and nowhere vanishing function.
@@
 where the $\epsilon_i$ are uncorrelated standard normal errors.
-For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters, $\bt$, is expressed in terms of probability distributions.
-We start our treatise by assuming that we have prior knowledge in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
+For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters is expressed in terms of probability distributions.
+We start our treatise by assuming that we have prior knowledge about a parameter vector, $\bt$, in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
 The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of
@@
 
-Because this will permit to do part of the integration to be performed later analytically, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred.
-In a one dimensional model this can always be achieved through re-parametrization [XXX].
+At this point, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred.
+This will permit us to perform an integration that is needed in a later step analytically.
+In a one dimensional model this can always be achieved through re-parametrization (see, e.g., chapter 5 in Ref.
+\cite{risken_1989_FockerPlanck}).
 The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ if it was in an initial state $q_0$ at time $t = 0$, is expressed as a {\em path-integral} as
@@
 Instead of undertaking an often prohibitive numerical computation of the path integral, we now apply HMC to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
-For efficiently drawing parameter samples from (\ref{posterior_pathint}), we interpret the latter as the partition function of a 1D statistical mechanical system and simulate its dynamics employing the HMC algorithm \cite{duane_1987}.
-(This is always possible and does not depend on the nature of the SDE we consider.) The model parameters  $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
+For efficiently drawing parameter samples from (\ref{posterior_pathint}), we interpret the latter as the partition function of a 1D statistical mechanics system and simulate its dynamics employing the HMC algorithm \cite{duane_1987}.
+The model parameters  $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$.
 Each degree of freedom, $q(t)$ and $\bt$, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$ respectively, so that the system is defined by the Hamiltonian
@@
 Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field.
-The latter consists of two parts, a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only.
+The latter consists of two parts, a field that results from the measurements and is felt by the measurement beads only (last term on the r.h.s.
+of eq.
+\ref{Vdisc}), and a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads.
 The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm.
@@
 It is now essential to note that potential (\ref{Vdisc}) contains terms of distinct scaling in the potentially large numbers $N$ and $n$, that refer to dynamics on distinct time-scales.
-In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and for resolving its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.
+In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and to resolve its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.
 
@@
 This elegant `analytical' solution is one main boosting part of our algorithm.
-For the inner propagator, we employ the time-reversible and symplecticity preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
+For the inner propagator, we employ the time-reversible and volume preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
 \begin{align}
@@
 For our toy system, we have considered a simple sinusoidal input  $r(t) = \sin^2 \left( 0.01 t \right) + 0.1$.
-A system realization was first obtained from Eq.~(\ref{sde}) using $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
+A system realization was first obtained from Eq.~(\ref{sde}) using $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}}=0.2$.
 Such system realization was then used to generate a synthetic time series of observed data according to Eq.~(\ref{data}).
@@
     System realization (solid line) with synthetic observations (filled circles, with error bars indicating the assumed measurement uncertainty).
-    The system response closely follows time-delayed the oscillations of the sinusoidal input (dashed).
+    The system response closely follows the oscillations of the sinusoidal input (dashed) in a time-delayed manner.
     Parameters for this figure and all following figures: $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$.
@@
 \end{figure}
-The Markov chains for parameters $K$ and $\gamma$, obtained after 50000 iterations of the HMC algorithm, are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively; they are fully compatible with the true, to be inferred, parameter values $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}}=0.2$.
+The Markov chains for parameters $K$ and $\gamma$, obtained after 50000 iterations of the HMC algorithm, are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively; they are fully compatible with the true, to be inferred, parameter values $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
 
@@
 
-We presented a novel, extremely efficient and versatile approach for SDE parameter estimation, based on observed data.
-Our algorithm derives its strenght from a translation of the problem of generating posterior parameter samples to the problem of simulating the dynamics of a statistical mechanical system.
-The main efficiency gain of our algorithm, compared to the one suggested in Ref.
-\cite{girolami_2011_HMC}, stems from an exploitation of the fact that this dynamics generically happens on very different time scales.
+We presented a novel, extremely efficient and versatile approach for data-based SDE parameter estimation.
+Our algorithm obtains its strength from translating the problem of generating posterior parameter samples into the problem of simulating the dynamics of a statistical mechanics system.
+The main novelty in our algorithm is the exploitation of the fact that this dynamics generically happens on very different time scales.
 Furthermore, at least for 1D systems, our approach also allows for an analytical, and therefore computationally efficient, integration of the fastest part of the dynamics.
-The algorithm suggested in Ref.
-\cite{girolami_2011_HMC} locally adapts the mass matrix to the curvature of the potential, whereas we use a constant diagonal mass matrix, with reasonable choices of masses, heavier for the measurement beads, lighter for the discretization (staging) beads.
-Whilst the former approach leads to some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved, it is likely to be beneficial in cases with strongly varying curvature.
-Our approach, however, is to a large extent complementary to and could be combined with the one in \cite{girolami_2011_HMC}.
+
+Our algorithms works with a constant diagonal mass matrix, with reasonable choices of masses - heavier for the measurement beads, lighter for the discretization (staging) beads.
+We expect this to work well in most application cases.
+However, if the curvature of the potential varies strongly, it might be beneficial to adapt the mass matrix to the local curvature as suggested in Ref. \cite{girolami_2011_HMC}.
+This, in turn, comes with some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved.
+Furthermore, we will no longer be able to solve part of the dynamics analytically.
+Nonetheless, for certain applications, a combination of the scale separation method proposed in this paper with a local adaptation of the mass matrix from Ref.
+\cite{girolami_2011_HMC} might  be the most efficient approach.
 
@@
 
-The structure of our algorithm is well suited to parallelization, which is important in particular if we deal with a high number of measurements (e.g., some tens of thousands data points).
+The structure of our algorithm is well suited to parallelization, which is important in particular if we deal with a high number of measurements.
 The algorithm can easily be adapted to other inference problems, such as higher dimensional SDE, and SDE coupled to ODE.

### 2015-10-26 Main_HMC_napa_sinrain.jl
minor updates (tex files untouched), cleaning up a bit

@@ -55,7 +55,7 @@ const dt = T/(N-1)                         # Time step
 const ty = iround(linspace(1, N, n+1))     # Indeces of "boundary" beads (= measurement points)     
 
 const nsample_burnin = 0                   # Number of points in the MCMC
-const nsample_eff    = 1000
+const nsample_eff    = 20000
 const nsample        = nsample_eff + nsample_burnin
 
 const dtau           = 0.25                # MD time step
@@ -186,7 +186,11 @@ plt.plot(t[ty], q[ty], "go",markersize=12)
 for s = 0:(n-1)
     u[s*j+1] = q[s*j+1]
     for k = 2:j
-        u[s*j+k] = q[s*j+k] - ( (k-1)*q[s*j+k+1] + q[s*j+1] )/k
+        # IMPORTANT: the transformation that should be applied is:
+        # u[s*j+k] = q[s*j+k] - ( (k-1)*q[s*j+k+1] + q[s*j+1] )/k
+        # BUT: it is always = 0 when the {q} between data points are linearly distributed
+        # For staging beads as linear interpolation of data points, therefore
+        u[s*j+k] = 0.0
     end
 end
 u[N] = q[N]
@@ -387,7 +391,7 @@ println(string("Fast RESPA in ", round(sum(time_respa_f),6), " seconds \n"))
 ## Back transformations (u -> q):
 ## (Tuckerman et al., JCP 99 (1993), eq. 2.19)
 ## --------------------------------------------------------------------------------------------
-
+#=back_transform_t = time()
 for sample_ind = 1:(nsample+1)
     for s = 1:n
         qs[sample_ind, (s-1)*j+1] = u_sample[sample_ind, (s-1)*j+1]
@@ -399,8 +403,26 @@ for sample_ind = 1:(nsample+1)
         end
     end
     qs[sample_ind, N] = u_sample[sample_ind, N]
-end
+end=#
+# ABOVE is very slow (cache thrashing problem?), BELOW is a much faster version
 # q = (1/bet)*log(S/(K*r))
+
+back_transform_t = time()
+for sample_ind = 1:(nsample+1)
+    for s = 1:n
+        qs[sample_ind, (s-1)*j+1] = u_sample[sample_ind, (s-1)*j+1]
+    end
+    qs[sample_ind, N] = u_sample[sample_ind, N]
+end
+
+for sample_ind = 1:(nsample+1)
+    for s = n:-1:1
+        for k = j:-1:2 
[…]

### 2015-10-26 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
computational details updated in tex files

@@
 A minimal example is a perceptron [XXX], the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time.
-While this task can easily be achieved for clean data using gradient descent learning, for noisy input data, this is largely impossible, as noise cannot be learned.
+While this task can easily be achieved for clean data using gradient descent learning, for noisy input data this is largely impossible, as noise cannot be learned.
 The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
@@
 Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series.
-In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points.
+In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretization points.
 %Our approach can be combined with the approach in \cite{girolami_2011_HMC}.
@@
 
-Consider, for simplicity and concreteness, a reservoir dynamics S(t) that on the observation time-scale is linear, with other (inflow and outflow) processes happening at much shorter time scales, so that they can be described by white noise.
-Furthermore, assume that this noise scales linearly with the system state S(t).
+Consider, for simplicity and concreteness, a reservoir dynamics $S(t)$ that on the observation time-scale is linear, with other (inflow and outflow) processes happening at much shorter time scales, so that they can be described by white noise.
+Furthermore, assume that this noise scales linearly with the system state $S(t)$.
 The model equation is thus given by the SDE
@@
 \end{figure}
-Within the investigated range of parameters, we observed a reasonably linear dependence of the total computing time on the number of discretization points $N$.
-For example, a complete run with 20000 iterations with $j=30$ (i.e., $N = 301$ discretization points) required about 13 minutes (for implementation details see the next section).
+A complete run with 20000 iterations with $n=10$ and $j=30$ (i.e., $N = 301$ discretization points) required only about 40 seconds (for implementation details see the next section).
 A set of 200 simulated system realizations based on $j=30$ is shown in Fig.~(\ref{fig:spaghetti}), together with the generated synthetic data.
@@
 
-The algorithm was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9, http:\/\/julialang.org\/).
-One of the attractive features of this language is its large and dynamic library for numerical and scientific applications.
-From this, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions.
+The algorithm was implemented in C++ (version C++11) using the open source \emph{Adept} library (version 1.1)~\cite{Hogan_2014_adept}, which provides a powerful tool for fast reverse-mode automated differentiation (AD).
 Our algorithm benefits greatly from the use of AD.
@@
 The simulations were run on a serial implementation of the algorithm on a 64-bit laptop computer equipped with a Windows 7 operating system, a double-core 1.6~GHz CPU (Intel i5-4200U) and 8~GB of RAM.
-$n+1 = 11$ measurement points and $j = 10, 20,\dots, 50$ intermediate staging beads, with a total number of discretization points $N = 101, 201, \dots, 501$ were taken.
+We used $n+1 = 11$ measurement points and $j = 10, 20,\dots, 50$ intermediate staging beads, with a total number of discretization points $N = 101, 201, \dots, 501$.
 In the Hamiltonian propagator~(\ref{propagator}), we set $\Delta\tau = 0.25$ and $P = 3$, with a constant total observation time $T = 833$ (arbitrary units of time).

### 2015-10-26 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
chapters 5&6 rewritten

@@
 
-\title{Boosting Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models by Hamiltonian Scale Separation}
+\title{Boosting Bayesian Parameter Inference of Nonlinear Stochastic Differential Equation Models by Hamiltonian Scale Separation}
 
@@
 The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy.
-A minimal example is a perceptron [XXX], the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time.
-While this task can easily be achieved for clean data using gradient descent learning, for noisy input data this is largely impossible, as noise cannot be learned.
+A minimal example is a perceptron \cite{rosenblatt_1958_perceptron, steeb_2004_problems}, the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time.
+While this task can easily be achieved for clean data using gradient descent learning, for noisy input data, this is largely impossible, as noise cannot be learned.
 The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
@@
 
-In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions [REF].
+In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions (see, e.g.
+Ref.
+\cite{box_2011_bayesian}).
 If a constant noise term is added to the output of a deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
@@
 A major problem with these simulation-based methods is, however, their inefficiency in the presence of many data points or high dimensions.
-One solution is to map the output space to a smaller dimensional space of summary statistics, and accept/reject proposed model parameters depending on how well associated model runs conform with the data in terms of these summary statistics [Ref.].
+One solution is to map the output space to a smaller dimensional space of summary statistics, and accept/reject proposed model parameters depending on how well associated model runs conform with the data in terms of these summary statistics \cite{Fearnhead_2012_ABC}.
 However, how to choose the summary statistics to achieve a significant representation of the posterior parameter distribution is a largely unsolved problem.
@@
 Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series.
-In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretization points.
+In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points.
 %Our approach can be combined with the approach in \cite{girolami_2011_HMC}.
@@
 
-Consider, for simplicity and concreteness, a reservoir dynamics $S(t)$ that on the observation time-scale is linear, with other (inflow and outflow) processes happening at much shorter time scales, so that they can be described by white noise.
-Furthermore, assume that this noise scales linearly with the system state $S(t)$.
+Consider, for simplicity and concreteness, a reservoir dynamics S(t) that on the observation time-scale is linear, with other (inflow and outflow) processes happening at much shorter time scales, so that they can be described by white noise.
+Furthermore, assume that this noise scales linearly with the system state S(t).
 The model equation is thus given by the SDE
@@
 
-Our algorithms works with a constant diagonal mass matrix, with reasonable choices of masses - heavier for the measurement beads, lighter for the discretization (staging) beads.
-We expect this to work well in most application cases.
-However, if the curvature of the potential varies strongly, it might be beneficial to adapt the mass matrix to the local curvature as suggested in Ref. \cite{girolami_2011_HMC}.
-This, in turn, comes with some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved.
-Furthermore, we will no longer be able to solve part of the dynamics analytically.
-Nonetheless, for certain applications, a combination of the scale separation method proposed in this paper with a local adaptation of the mass matrix from Ref.
-\cite{girolami_2011_HMC} might  be the most efficient approach.
+In most application cases, our choice of a fixed diagonal mass matrix, for reasonable choices of masses - heavier for the measurement beads, lighter for the discretization (staging) beads - can be expected to work well.
+Nonetheless, if the curvature of the potential varies strongly, it might be beneficial to adapt the mass matrix to the local curvature as suggested in Ref.
+\cite{girolami_2011_HMC}.
+For such cases, a combination of the scale separation method proposed in this paper with the local mass matrix adaptation of Ref.
+\cite{girolami_2011_HMC} might be the most efficient solution.
+This extension, however, comes at the price of a computational overhead (second derivatives have to be calculated and implicit equations have to be solved), and we will no longer be able to solve part of the dynamics analytically.
+On a more general level, given the wide field of very distinct applications, optimal parameter inference for SDE models will not be provided by one single approach, but will require a set of tools, to make the optimal choice from.
+We expect statistical physics to continue making strong contributions toward this aim.
 

### 2015-10-27 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
version to be submitted

@@
 
-Modeling a dynamical process starts with a basic model that is usually obtained from more or less deep insight into the nature of the process.
+Modeling a dynamical process starts with a basic model that is usually obtained from a more or less deep insight into the nature of the process.
 The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy.
@@
 Problems of this kind are commonly solved by Monte Carlo (MC) methods that are based on simulating model realizations and comparing them to the data.
-Popular methods are particle filters \cite{chopin_2013_SMC2}, Metropolis-within-Gibbs algorithms \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters} or Approximate Bayes Computations (ABC) \cite{marin_2012_ABC, Albert_2014_ABC}.
+Popular methods are particle filters \cite{chopin_2013_SMC2}, Metropolis-within-Gibbs algorithms \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters} or Approximate Bayes Computations \cite{marin_2012_ABC, Albert_2014_ABC}.
 A major problem with these simulation-based methods is, however, their inefficiency in the presence of many data points or high dimensions.
@@
 These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanics system and by simulating the dynamics of the latter.
-After discretizing the time of the original problem, we are led to the statistical mechanics of a polymer %-like problem
-with harmonic bonds in an exterior potential.
+After discretizing the time of the original problem, we are led to a problem akin to the statistical mechanics of a polymer %-like problem
+with harmonic bonds in an exterior potential \cite{chandler_1981_polymer}.
 In this framework, the measurements are interpreted as an additional exterior potential that acts only on the polymer's `measurement beads', and model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer.
@@
 This suggests a multiple time scale integration technique for the simulation of the statistical mechanics system \cite{tuckerman_1993}.
-We will show that for 1D SDEs we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics.
-Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series.
+We will show that for 1D SDE we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics.
+Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDE amenable to a full-fledged Bayesian inference with time-series.
 In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points.
@@
 \end{equation}
-where $r(t)$ denotes the time varying rain input, $K$ denotes the retention time,$\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
+where $r(t)$ denotes the time varying rain input, $K$ denotes the retention time, $\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
 \begin{equation}\label{whitenoise}
@@
 By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested, e.g., as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
-In this setting, the input $r(t)$ is a smooth and nowhere vanishing function.
+In our setting, the input $r(t)$ is a smooth and nowhere vanishing function.
 We assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors,
@@
 For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters is expressed in terms of probability distributions.
+
 We start our treatise by assuming that we have prior knowledge about a parameter vector, $\bt$, in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
@@
 where $L(\vc y|\bt)$ is the probability distribution for model outputs given model parameters, evaluated at the measured data (the so-called {\em likelihood function}).
-We posit that parameters must be non-negative; otherwise we assume no additional prior knowledge (i.e., the posterior is ‘data-driven’ [XX]).
+%We posit that parameters must be non-negative; otherwise we assume no additional prior knowledge (i.e., the posterior is ‘data-driven’ \cite{box_2011_bayesian}).
 
@@
 In these new variables and parameters, Eq.~(\ref{sde}) becomes
-\begin{equation}
+\begin{equation}\label{standardform}
   \dot q(t)
@@
 %\end{equation}
-Scale-invariance of Eq.~(\ref{sde}) leads to power law tails in the probability distribution of $S(t)$.
+Invariance of Eq.~(\ref{sde}) under re-scaling of both $r(t)$ and $S(t)$ leads to power law tails in the probability distribution of $S(t)$.
 In real-world hydrology, for which Eq.~(\ref{sde}) is a model, indeed often fat-tailed error distributions are observed ~\cite{thyer_2009_fattails}.
 
-At this point, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred.
-This will permit us to perform an integration that is needed in a later step analytically.
-In a one dimensional model this can always be achieved through re-parametrization (see, e.g., chapter 5 in Ref.
-\cite{risken_1989_FockerPlanck}).
 The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ if it was in an initial state $q_0$ at time $t = 0$, is expressed as a {\em path-integral} as
@@
 $\langle S\rangle_{eq}=Kr_{0}$
-and its variance for $\gamma< 2$ is given by $ \langle (S - \langle S\rangle_{eq})^2\rangle_{eq}
+and its variance, for $\gamma< 2$, is given by $ \langle (S - \langle S\rangle_{eq})^2\rangle_{eq}
   =
   K^2r_{0}^2
-  \gamma/(2-\gamma)$, which for $\gamma\geq 2$ is seen to diverge.
+  \gamma/(2-\gamma)$, which, for $\gamma\geq 2$, is seen to diverge.
 The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.
@@
 
-Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field.
+Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field \cite{chandler_1981_polymer}.
 The latter consists of two parts, a field that results from the measurements and is felt by the measurement beads only (last term on the r.h.s.
-of eq.
-\ref{Vdisc}), and a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads.
-The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm.
-Since measurement beads are constrained more than intermediate beads, we will assign larger masses to the former.
-Fig.
-(\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer.
-The physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced as the vertical axis.
+of Eq.~(\ref{Vdisc})), and a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads.
+The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm. Since measurement beads are constrained more than intermediate beads, we will assign larger masses to the former. 
+Fig. (\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer. 
+%The physical time (horizontal axis) is interpreted as a spatial dimension.
 Measurement beads only move within the measurement uncertainty, whilst the intermediate beads explore much larger regions of phase space.
@@
 It is now essential to note that potential (\ref{Vdisc}) contains terms of distinct scaling in the potentially large numbers $N$ and $n$, that refer to dynamics on distinct time-scales.
-In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and to resolve its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.
+In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and to resolve its dynamics, brute force numerical integration of Hamilton's equations in step 2 of the HMC algorithm would require a very small discretization time-step.
 
@@
 with $  q^*_{(s-1)j+k}  =  ((k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} )/k$.
-For the boundary beads, corresponding to the original measurement points, we apply the coordinate transformations
-$  u_{sj+1} = q_{sj+1}$, $s=0,\dots,n$
-and for the intermediate staging beads
+The boundary beads, corresponding to the original measurement points, are not transformed,
+$  u_{sj+1} = q_{sj+1}$, $s=0,\dots,n$,
+while, for the intermediate staging beads, we apply the coordinate transformations
 $u_{sj+k} = q_{sj+k} - q^*_{sj+k}$, $s=0,\dots,n-1$, $k=2,\dots,j$.
@@
   +
-  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,
+  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,\label{analytical1}
   \\
@@
   -
-  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
+  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,\label{analytical2}
 \end{align}
 for $s=1,\dots,n$ and $k=2,\dots,j$.
-This elegant `analytical' solution is one main boosting part of our algorithm.
 For the inner propagator, we employ the time-reversible and volume preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
@@
 
+The analytical solution (\ref{analytical1}) and (\ref{analytical2}) is one main boosting part of our algorithm.
+To find such a solution, it is important to arrive at model equations of the form (\ref{standardform}), where the noise term neither depends on the state variables, nor on the parameters to be inferred.
+In a one dimensional model this can always be achieved through re-parametrization (see, e.g., chapter 5 in Ref. \cite{risken_1989_FockerPlanck}).
+In higher dimensions, this will not always be possible.
+But even in such cases, we will be able to boost our algorithm through assigning smaller time intervals $\Delta\tau$ to the fast dynamics and larger ones to the slow dynamics.
+
 \section{Results}\label{numerical_results}
@@
 \end{figure}
-A complete run with 20000 iterations with $n=10$ and $j=30$ (i.e., $N = 301$ discretization points) required only about 40 seconds (for implementation details see the next section).
+A complete run with 50000 iterations with $n=10$ and $j=30$ (i.e., $N = 301$ discretization points) required only about 100 seconds (for implementation details see the next section).
 A set of 200 simulated system realizations based on $j=30$ is shown in Fig.~(\ref{fig:spaghetti}), together with the generated synthetic data.
@@
 \end{figure}
-The Markov chains for parameters $K$ and $\gamma$, obtained after 50000 iterations of the HMC algorithm, are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively; they are fully compatible with the true, to be inferred, parameter values $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
+The Markov chains for parameters $K$ and $\gamma$, obtained after 50000 iterations of the HMC algorithm, are shown in Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG}), respectively.
 
@@
 The Markov chains (Figs.~(\ref{fig:chainK}) and (\ref{fig:chainG})) determine the probability density functions (PDF) for $K$ and $\gamma$, respectively.
-The result obtained by using the built-in kernel density estimator provided by Mathematica (version 10) is exhibited in  Fig.~(\ref{fig:KG_distr}).
+The results obtained by using the built-in kernel density estimator provided by Mathematica (version 10) are exhibited in  Fig.~(\ref{fig:KG_distr}); they are fully compatible with the true, to be inferred, parameter values $K_{\mbox{\small true}}$ and $\gamma_{\mbox{\small true}}$.
 
@@
 We presented a novel, extremely efficient and versatile approach for data-based SDE parameter estimation.
-Our algorithm obtains its strength from translating the problem of generating posterior parameter samples into the problem of simulating the dynamics of a statistical mechanics system.
-The main novelty in our algorithm is the exploitation of the fact that this dynamics generically happens on very different time scales.
+Our algorithm obtains its strength from translating the problem of generating posterior parameter samples into the problem of simulating the dynamics of a statistical mechanics system;
+the main novelty in our algorithm is the exploitation of the fact that this dynamics generically happens on very different time scales.
 Furthermore, at least for 1D systems, our approach also allows for an analytical, and therefore computationally efficient, integration of the fastest part of the dynamics.
